---
title: "Cameroon plaque taxonomy analysis - Kraken RefSeq database w/ & w/o Pasolli MAGs"
author: "Irina Velsko"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
    github_document:
        fig_width: 8
        fig_height: 6
        toc: true
---

```{r load_libraries, echo = F, message = F}
library(knitr)
library(data.table)
library(janitor)
library(taxa)
library(plyr)
library(tidyverse)
library(gplots)
library(ggrepel)
library(viridis)
library(cowplot)
opts_chunk$set(echo=F, warning=F, message=F, dpi=150)
```

```{r setup}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```


```{r load_data}
load("./05-results.backup/kraken_refseq_run1.RData")

# this file has the full taxonomy string for all samples (by hand added in all missing f__|g__, etc)
all_taxonomy <- fread("./00-documentation.backup/all_taxonomy_filtered_split.csv")
all_taxonomy <- all_taxonomy %>%
  as_tibble(.) %>%
  select(-Root) %>%
  mutate(Taxonomy = str_replace_all(Taxonomy, "root\\|", ""))

load("./05-results.backup/CMC_metadata.RData")

# remove the samples and columns we're not looking at
metadata_mcR <- metadata %>%
  filter(!str_detect(SampleID, "SRR164|10M|EXB|LIB")) %>%
  select(SampleID, Env, Description, Industry, Ethnic_Group, Village, Market_economy, Age_group, Sex, Tooth_site)

```

Create tables that present the number of read assignments at each taxonomic
level (Phylum, Genus, Species) with the RefSeqOnly and RefSeq+PasolliMAGs
databases, the difference between each, and the fold-difference between the
number of assignments to each taxon. Positive fold-difference means there were
more assignments to that taxon in the RefSeq+PasolliMAGs database, and negative
values means there were more assignments to that taxon in the RefSeqOnly
database.
```{r taxon_differences}

db_phylum_compare <- rso_reports_tax_run1 %>%
  filter(str_detect(SampleID, "CMC")) %>%
  filter(str_detect(Rank, "P")) %>%
  group_by(SampleID, SciName) %>%
  summarize(Rooted_avg = mean(Number_rooted)) %>%
  mutate(Database = "RefSeqOnly") %>%
  select(SciName, SampleID, Database, Rooted_avg) %>%
  bind_rows(., rsmp_reports_tax_run1 %>%
              filter(str_detect(SampleID, "CMC")) %>%
              filter(str_detect(Rank, "P")) %>%
              group_by(SampleID, SciName) %>%
              summarize(Rooted_avg = mean(Number_rooted)) %>%
              mutate(Database = "RefSeqPasolliMAGs") %>%
              select(SciName, SampleID, Database, Rooted_avg)) %>%
  group_by(Database, SciName) %>%
  summarize(Total_rooted = sum(Rooted_avg)) %>%
  spread(Database, Total_rooted) %>%
  select(SciName, RefSeqOnly, RefSeqPasolliMAGs) %>%
  mutate(Read_difference = RefSeqPasolliMAGs - RefSeqOnly, 
         Fold_difference = log2(RefSeqPasolliMAGs/RefSeqOnly)) %>%
  arrange(desc(Fold_difference))

db_genus_compare <- rso_reports_tax_run1 %>%
  filter(str_detect(SampleID, "CMC")) %>%
  filter(str_detect(Rank, "G")) %>%
  group_by(SampleID, SciName) %>%
  summarize(Rooted_avg = mean(Number_rooted)) %>%
  mutate(Database = "RefSeqOnly") %>%
  select(SciName, SampleID, Database, Rooted_avg) %>%
  bind_rows(., rsmp_reports_tax_run1 %>%
              filter(str_detect(SampleID, "CMC")) %>%
              filter(str_detect(Rank, "G")) %>%
              group_by(SampleID, SciName) %>%
              summarize(Rooted_avg = mean(Number_rooted)) %>%
              mutate(Database = "RefSeqPasolliMAGs") %>%
              select(SciName, SampleID, Database, Rooted_avg)) %>%
  group_by(Database, SciName) %>%
  summarize(Total_rooted = sum(Rooted_avg)) %>%
  spread(Database, Total_rooted) %>%
  select(SciName, RefSeqOnly, RefSeqPasolliMAGs) %>%
  mutate(Read_difference = RefSeqPasolliMAGs - RefSeqOnly, 
         Fold_difference = log2(RefSeqPasolliMAGs/RefSeqOnly)) %>%
  arrange(desc(Fold_difference))

db_species_compare <- rso_reports_tax_run1 %>%
  filter(str_detect(SampleID, "CMC")) %>%
  filter(str_detect(Rank, "S")) %>%
  group_by(SampleID, SciName) %>%
  summarize(Rooted_avg = mean(Number_rooted)) %>%
  mutate(Database = "RefSeqOnly") %>%
  select(SciName, SampleID, Database, Rooted_avg) %>%
  bind_rows(., rsmp_reports_tax_run1 %>%
              filter(str_detect(SampleID, "CMC")) %>%
              filter(str_detect(Rank, "S")) %>%
              group_by(SampleID, SciName) %>%
              summarize(Rooted_avg = mean(Number_rooted)) %>%
              mutate(Database = "RefSeqPasolliMAGs") %>%
              select(SciName, SampleID, Database, Rooted_avg)) %>%
  group_by(Database, SciName) %>%
  summarize(Total_rooted = sum(Rooted_avg)) %>%
  spread(Database, Total_rooted) %>%
  select(SciName, RefSeqOnly, RefSeqPasolliMAGs) %>%
  mutate(Read_difference = RefSeqPasolliMAGs - RefSeqOnly, 
         Fold_difference = log2(RefSeqPasolliMAGs/RefSeqOnly)) %>%
  arrange(desc(Fold_difference))


```

There's a lot to look through in the tables above, so to make it easier to
diges, look at the taxa that have at least 100 reads assigned by both databases,
and for which the read difference is at least 100.
```{r}
db_species_compare_short <- db_species_compare %>%
  filter(RefSeqOnly >=100, RefSeqPasolliMAGs >= 100, Read_difference >=100 | Read_difference <= -100)

db_genus_compare_short <- db_genus_compare %>%
  filter(RefSeqOnly >=100, RefSeqPasolliMAGs >= 100, Read_difference >=100 | Read_difference <= -100)

db_phylum_compare_short <- db_phylum_compare %>%
  filter(RefSeqOnly >=100, RefSeqPasolliMAGs >= 100, Read_difference >=100 | Read_difference <= -100)

db_species_compare_phage <- db_species_compare %>%
  filter(str_detect(SciName, "phage")) %>%
  filter(!str_detect(SciName, "Bacteria"))

```

Look at the taxonomy of the MAGs from the Pasolli paper taht were added to the
RefSeq database to create our RefSeq+PasolliMAGs database. Where were these taxa
in the taxonomic tree? What are the most abundant taxa at each level (can we
expect increases in assignments of certain taxa b/c there are a lot more of them
in the RSPM database?)
```{r mag_info}
# this is where the Pasolli MAG info is: 
# /projects1/microbiome_calculus/Cameroon_plaque/00-documentation.backup/Pasolli2019_SGBinfo.csv
# maybe parse this?


mag_info <- fread("00-documentation.backup/Pasolli2019_SGBinfo.csv")
mag_info <- as_tibble(mag_info)
mag_info %>% filter(str_detect(closestTaxLevel, "Corynebacterium")) %>% select(-estTaxonomy) %>% arrange(closestTaxLevel) %>% print(., n = 59)

# total counts at each level
mag_info %>%
  group_by(estTaxLevel) %>%
  count()

# counts of each entry at the specified levels
mag_info %>%
  filter(estTaxLevel == "Other") %>%
  select(closestTaxLevel) %>%
  mutate(phylum = sapply(closestTaxLevel, function(f) {
                                  unlist(str_split(f, "\\|"))[2]
                           })) %>%
  filter(!str_detect(phylum, "c__|o__|f__|g__|s__")) %>%
  count(phylum) %>%
  arrange(desc(n)) 
# %>%
#   summarize(total = sum(n))

mag_info %>%
  filter(estTaxLevel == "Family") %>%
  select(closestTaxLevel) %>%
  mutate(family = sapply(closestTaxLevel, function(f) {
                                  unlist(str_split(f, "\\|"))[5]
                           })) %>%
  filter(!str_detect(family, "p__|c__|o__|g__|s__")) %>%
  count(family) %>%
  arrange(desc(n)) 
# %>%
#   summarize(total = sum(n))

mag_info %>%
  filter(estTaxLevel == "Genus") %>%
  select(closestTaxLevel) %>%
  mutate(genus = sapply(closestTaxLevel, function(f) {
                                  unlist(str_split(f, "\\|"))[6]
                           })) %>%
  filter(!str_detect(genus, "p__|c__|o__|f__|s__")) %>%
  count(genus) %>%
  arrange(desc(n)) 
# %>%
#   summarize(total = sum(n))

mag_info %>%
  filter(estTaxLevel == "Species") %>%
  select(closestTaxLevel) %>%
  mutate(species = sapply(closestTaxLevel, function(f) {
                                  unlist(str_split(f, "\\|"))[7]
                           })) %>%
  filter(!str_detect(species, "p__|c__|o__|f__|g__")) %>%
  count(species) %>%
  arrange(desc(n)) 
# %>%
#   summarize(total = sum(n))


```

```{r, eval = F}
table_temp <- fread("./04-analysis/kraken/output/RefSeqPasolliMAGs/SRR514329_S0_L001_R1_000.fastq.combined.fq.prefixed.extractunmapped.bam.fastq.gz.output.1.kraken")

table_temp1 <- table_temp %>%
  filter(V3 == 0, V1 != "U")
```

Let's see if we can get the data into metacoder to visualize how the assignments change around the tree
```{r otu_tables}
# make an OTU table for the RefSeqOnly dataset
rso_otu_table <- rso_reports_tax_run1 %>%
  select(NCBItaxID, SampleID, Number_direct) %>%
  spread(SampleID, Number_direct, fill = 0) %>%
  filter(NCBItaxID != "0") %>%
  select(-matches("10M|SRR164|EXB|LIB"))

# make an OTU table for the RefSeq+PasolliMAGs dataset
rsmp_otu_table <- rsmp_reports_tax_run1 %>%
  select(NCBItaxID, SampleID, Number_direct) %>%
  spread(SampleID, Number_direct, fill = 0) %>%
  filter(NCBItaxID != "0") %>%
  select(-matches("10M|SRR164|EXB|LIB"))

```

```{r otu_table_tax}

rso_otu_table <- rso_otu_table %>%
  inner_join(., all_taxonomy, by = "NCBItaxID")

rsmp_otu_table <- rsmp_otu_table %>%
    inner_join(., all_taxonomy, by = "NCBItaxID")

```

```{r taxa_object}

obj <- parse_tax_data(rso_otu_table,
                      class_cols = "Taxonomy",
                      class_sep = "\\|",
                      # class_regex = "^([a-zA-Z]{0,1})_{0,2}(.*)$",
                      class_regex = "^(.+)__(.+)$",
                      class_key = c(tax_rank = "info",
                                    name = "taxon_name"))
print(obj)

```









































