---
title: "Cameroon hunter-gatherer calculus/plaque HUMAnN2"
author: "Irina Velsko"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
    github_document:
        fig_width: 8
        fig_height: 6
        toc: true
---


```{r load_libraries, echo = F, message = F}
library(knitr)
library(decontam)
library(data.table)
library(vegan)
library(ape)
library(mixOmics)
library(janitor)
library(rstatix)
library(pander)
library(tidyverse)
library(gplots)
library(ggrepel)
library(viridis)
library(cowplot)
opts_chunk$set(echo=F, warning=F, message=F, dpi=150)
```

Set the working directory of the notebook to the parent directory of the 
notebook itself.  

```{r setup}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```


# HUMAnN2 assignment stats
```{r load_data, eval = F}
## load the species and genus tables generated with HUMAnN2
humann2_path <- fread("./05-results.backup/humann2.pathabundance.all.tss.tsv")
humann2_path <- as_tibble(humann2_path)
humann2_KOs_full <- fread("./05-results.backup/humann2.genefamilies.all.i.tss.renorm.ko.tsv")
humann2_KOs_full <- as_tibble(humann2_KOs_full)

# clean the file names
humann2_KOs_full <- rename(humann2_KOs_full, Ortholog = `# Gene Family`)
colnames(humann2_KOs_full) <- gsub(".SG1.1_S0_L001_R1_001.fastq.combined.fq.prefixed.extractunmapped.bam_Abundance-RPKs","", colnames(humann2_KOs_full))
colnames(humann2_KOs_full) <- gsub("_S0_L001_R1_000.fastq.combined.fq.prefixed.extractunmapped.bam_Abundance-RPKs","", colnames(humann2_KOs_full))
colnames(humann2_KOs_full) <- gsub("_S0_L001_R1_001.fastq.combined.fq.prefixed.extractunmapped.bam.10M_Abundance-RPKs","_10M", colnames(humann2_KOs_full))
colnames(humann2_KOs_full) <- gsub("_S0_L001_R1_000.10M.fastq.combined.fq.prefixed.extractunmapped.bam_Abundance-RPKs","_10M", colnames(humann2_KOs_full))
colnames(humann2_KOs_full) <- gsub("_S0_L001_R1_001.fastq.combined.fq.prefixed.extractunmapped.bam_Abundance-RPKs","", colnames(humann2_KOs_full))
colnames(humann2_KOs_full) <- gsub(".fastq.truncated_Abundance-RPKs","", colnames(humann2_KOs_full))

humann2_path <- rename(humann2_path, Pathway = `# Pathway`)
colnames(humann2_path) <- gsub(".SG1.1_S0_L001_R1_001.fastq.combined.fq.prefixed.extractunmapped.bam_Abundance","", colnames(humann2_path))
colnames(humann2_path) <- gsub("_S0_L001_R1_000.fastq.combined.fq.prefixed.extractunmapped.bam_Abundance","", colnames(humann2_path))
colnames(humann2_path) <- gsub("_S0_L001_R1_001.fastq.combined.fq.prefixed.extractunmapped.bam.10M_Abundance","_10M", colnames(humann2_path))
colnames(humann2_path) <- gsub("_S0_L001_R1_000.10M.fastq.combined.fq.prefixed.extractunmapped.bam_Abundance","_10M", colnames(humann2_path))
colnames(humann2_path) <- gsub("_S0_L001_R1_001.fastq.combined.fq.prefixed.extractunmapped.bam_Abundance","", colnames(humann2_path))
colnames(humann2_path) <- gsub(".fastq.truncated_Abundance","", colnames(humann2_path))


```

```{r load_data}
load("./05-results.backup/CMC_humann2_tables.RData")
load("./05-results.backup/CMC_metadata.RData")

# remove unmapped and ungrouped reads
humann2_KOs <- humann2_KOs_full %>% filter(Ortholog != "UNMAPPED") %>% filter(!str_detect(Ortholog, "UNGROUPED"))

```


```{r functions}

# add functions for:
# making the input table for PCA?

# plotting PCA with colored dots
plot_pca <- function(df, pc1, pc2, color_group, shape_group, size_group) {
    metadata_group_colors <- get(paste(color_group, "_colors", sep = ""))
    metadata_group_shapes <- get(paste(shape_group, "_shapes", sep = ""))
    metadata_group_size <- get(paste(size_group, "_size", sep = ""))

    exp_var <- paste0(round(df$explained_variance * 100, 2), "%")
    df_X <- df$variates$X %>%
              as.data.frame() %>%
              rownames_to_column("SampleID") %>%
              inner_join(metadata %>%
                           select(SampleID, Description, Ethnic_Group, Age_group, Sex, Market_economy, Village, Tooth_site, Type), by = "SampleID")
    
    color_group = df_X[[color_group]]
    shape_group = df_X[[shape_group]]
    size_group = df_X[[size_group]]
                      
                       if (pc1 == 'PC1') {
                           pc1 <- df_X$PC1
                           exp_var_pc1 <- exp_var[1]
                           xaxis <- c("PC1")
                       }  else if (pc1 == 'PC2') {
                           pc1 <- df_X$PC2
                           exp_var_pc1 <- exp_var[2]
                           xaxis <- c("PC2")
                       } else if (pc1 == 'PC3') {
                           pc1 <- df_X$PC3
                           exp_var_pc1 <- exp_var[3]
                           xaxis <- c("PC3")
                       }

                       if (pc2 == 'PC1') {
                           pc2 <- df_X$PC1
                           exp_var_pc2 <- exp_var[1]
                           yaxis <- c("PC1")
                       }  else if (pc2 == 'PC2') {
                           pc2 <- df_X$PC2
                           exp_var_pc2 <- exp_var[2]
                           yaxis <- c("PC2")
                       } else if (pc2 == 'PC3') {
                           pc2 <- df_X$PC3
                           exp_var_pc2 <- exp_var[3]
                           yaxis <- c("PC3")
                       }

    pca_plot <- ggplot(df_X, aes(pc1, pc2, colour = color_group, shape = shape_group, size = size_group)) +
                        geom_point() +
                        scale_colour_manual(values = metadata_group_colors) +
                        scale_shape_manual(values = metadata_group_shapes) +
                        scale_size_manual(values = metadata_group_size) +
                        # stat_ellipse() +
                        xlab(paste(xaxis, " - ", exp_var_pc1)) +
                        ylab(paste(yaxis, " - ", exp_var_pc2)) +
                        theme_minimal(base_size = 10) +
                        theme(legend.position = "top") +
                        theme(legend.title = element_blank())

    return(pca_plot)
}

# this function plots the PCA with the sample names instead of points
plot_pca_names <- function(df, pc1, pc2, metadata_group) {
    metadata_group_colors <- get(paste(metadata_group, "_colors", sep = ""))

    exp_var <- paste0(round(df$explained_variance * 100, 2), "%")
    df_X <- df$variates$X %>%
              as.data.frame() %>%
              rownames_to_column("SampleID") %>%
              inner_join(metadata %>%
                           select(SampleID, Description, Ethnic_Group, Age_group, Sex, Market_economy, Village, Tooth_site, Type), by = "SampleID")
    
    metadata_group = df_X[[metadata_group]]
    
                      # select the correct PC column and explained variance for PC1
                       if (pc1 == 'PC1') {
                           pc1 <- df_X$PC1
                           exp_var_pc1 <- exp_var[1]
                           xaxis <- c("PC1")
                       }  else if (pc1 == 'PC2') {
                           pc1 <- df_X$PC2
                           exp_var_pc1 <- exp_var[2]
                           xaxis <- c("PC2")
                       } else if (pc1 == 'PC3') {
                           pc1 <- df_X$PC3
                           exp_var_pc1 <- exp_var[3]
                           xaxis <- c("PC3")
                       }

                       if (pc2 == 'PC1') {
                           pc2 <- df_X$PC1
                           exp_var_pc2 <- exp_var[1]
                           yaxis <- c("PC1")
                       }  else if (pc2 == 'PC2') {
                           pc2 <- df_X$PC2
                           exp_var_pc2 <- exp_var[2]
                           yaxis <- c("PC2")
                       } else if (pc2 == 'PC3') {
                           pc2 <- df_X$PC3
                           exp_var_pc2 <- exp_var[3]
                           yaxis <- c("PC3")
                       }

    pca_plot_names <- ggplot(df_X, aes(pc1, pc2, colour = metadata_group)) +
                        geom_text(aes(label = df_X$SampleID), size = 2) +
                        scale_colour_manual(values = metadata_group_colors) +
                        # stat_ellipse() +
                        xlab(paste(xaxis, " - ", exp_var_pc1)) +
                        ylab(paste(yaxis, " - ", exp_var_pc2)) +
                        theme_minimal(base_size = 10) +
                        theme(legend.position = "top") +
                        theme(legend.title = element_blank())

    return(pca_plot_names)
}

# this function plots the PCA with biplots to look at the categories with strongest loadings
plot_pca_bi <- function(df, pc1, pc2, metadata_group, columntitle, pathways = c("TRUE","FALSE")) {
    metadata_group_colors <- get(paste(metadata_group, "_colors", sep = ""))
    metadata_group_shapes <- get(paste(metadata_group, "_shapes", sep = ""))
   
    arrow_pc <- enquo(columntitle)
    
    exp_var <- paste0(round(df$explained_variance * 100, 2), "%") # explained variance for x- and y-labels
    
    # select only the PCs from the PCA and add metadata
    df_X <- df$variates$X %>%
              as.data.frame() %>%
              rownames_to_column("SampleID") %>%
              inner_join(metadata %>%
                           select(SampleID, Description, Ethnic_Group, Age_group, Sex, Market_economy, Village, Tooth_site, Type), by = "SampleID")
    
    metadata_group = df_X[[metadata_group]]
    
    corr_lam <- df$sdev[c("PC1", "PC2", "PC3")] * sqrt(nrow(df_X))

    df_X <- df_X %>%
      mutate(PC1 = PC1 / corr_lam[1],
             PC2 = PC2 / corr_lam[2], 
             PC3 = PC3 / corr_lam[3])

    # select the correct PC column and explained variance for PC1
    if (pc1 == 'PC1') {
        Pc1 <- df_X$PC1
        exp_var_pc1 <- exp_var[1]
        xaxis <- c("PC1")
    } else if (pc1 == 'PC2') {
        Pc1 <- df_X$PC2
        exp_var_pc1 <- exp_var[2]
        xaxis <- c("PC2")
    } else if (pc1 == 'PC3') {
        Pc1 <- df_X$PC3
        exp_var_pc1 <- exp_var[3]
        xaxis <- c("PC3")
   }
    
    # select the correct PC column and explained variance for PC2
    if (pc2 == 'PC1') {
        Pc2 <- df_X$PC1
        exp_var_pc2 <- exp_var[1]
        yaxis <- c("PC1")
 } else if (pc2 == 'PC2') {
       Pc2 <- df_X$PC2
       exp_var_pc2 <- exp_var[2]
       yaxis <- c("PC2")
    } else if (pc2 == 'PC3') {
       Pc2 <- df_X$PC3
       exp_var_pc2 <- exp_var[3]
       yaxis <- c("PC3")
   }

    # Identify the 10 pathways that have highest positive and negative loadings in the selected PC
    pws_10 <- df$loadings$X %>%
      as.data.frame(.) %>%
      rownames_to_column(var = "Function") %>%
      top_n(10, !!arrow_pc)
   if (pathways == T) {
      pws_10 <- pws_10 %>%
        rename(pathway = Function) %>%
        mutate(Function = sapply(pathway, function(f) {
                                  unlist(str_split(f, ":"))[1]
                           })) 
    }

    neg_10 <- df$loadings$X %>%
      as.data.frame(.) %>%
      rownames_to_column(var = "Function") %>%
      top_n(-10, !!arrow_pc)
    if (pathways == T) {
      neg_10 <- neg_10 %>%
      rename(pathway = Function) %>%
       mutate(Function = sapply(pathway, function(f) {
                                  unlist(str_split(f, ":"))[1]
                           })) 
    }


    pca_plot_bi <- ggplot(df_X, aes(x = Pc1, y = Pc2, colour = metadata_group)) +
      geom_point(size = 2, aes(shape = metadata_group) )+
      geom_segment(data = pws_10,
                   aes(xend = get(paste(pc1)), yend = get(paste(pc2))),
                   x = 0, y = 0, colour = "black",
                   size = 0.5,
                   arrow = arrow(length = unit(0.03, "npc"))) +
      geom_label_repel(data = pws_10,
                   aes(x = get(paste(pc1)), y = get(paste(pc2)), label = Function),
                   size = 1.5, colour = "grey20", label.padding = 0.2) +
      geom_segment(data = neg_10,
                   aes(xend = get(paste(pc1)), yend = get(paste(pc2))),
                   x = 0, y = 0, colour = "grey50",
                   size = 0.5,
                   arrow = arrow(length = unit(0.03, "npc"))) +
      geom_label_repel(data = neg_10,
                   aes(x = get(paste(pc1)), y = get(paste(pc2)), label = Function),
                   size = 1.5, colour = "grey20", label.padding = 0.2) +
      labs(x = paste(xaxis, " - ", exp_var_pc1),
           y = paste(yaxis, " - ", exp_var_pc2)) +
      scale_colour_manual(values = metadata_group_colors) +
      scale_shape_manual(values = metadata_group_shapes) +
      theme_minimal() + theme(text = element_text(size = 10)) +
      theme(legend.position = "top")

    return(pca_plot_bi)
}

```

Before we proceed with looking at differences in abundance of different enzymes,
let's compare the how many reads were identified by HUMAnN2 for each sample
type. With the first step we'll read in the default HUMAnN2 output table with
all UniRef90 assignments. We'll select out just the rows with no species
assignemtns so we can calculate total assignments. HUMAnN2 uses *reads per*
*kilobase* for the initial output, so we'll be calculating the percent of total
assigned/mapped reads per kilobase (RPK). Then write out the file with just the
unmapped and total reads assigned so we never have to load this huge file again.
This has to be done with a table for which the samples were renormed
individually and then joined, and not renormed after they were joined (with
HUMAnN2 scripts).

Plot the proportion of assigned reads in each group.
```{r read_assignment_stats}

humann2_alignment_stats <- fread("./05-results.backup/humann2_alignment_stats.tsv")
humann2_alignment_stats$SampleID <- gsub("_S0_L001_R1_001.fastq.combined.fq.prefixed.extractunmapped.bam.10M.fastq.gz","_10M", humann2_alignment_stats$SampleID)
humann2_alignment_stats$SampleID <- gsub("_S0_L001_R1_000.10M.fastq.combined.fq.prefixed.extractunmapped.bam.fastq.gz","_10M", humann2_alignment_stats$SampleID)
humann2_alignment_stats$SampleID <- gsub("_S0_L001_R1_000.fastq.combined.fq.prefixed.extractunmapped.bam.fastq.gz","", humann2_alignment_stats$SampleID)
humann2_alignment_stats$SampleID <- gsub(".fastq.truncated.prefixed.gz","", humann2_alignment_stats$SampleID)
humann2_alignment_stats$SampleID <- gsub(".SG1.1","", humann2_alignment_stats$SampleID)

humann2_alignment_stats <- humann2_alignment_stats %>%
  arrange(SampleID) %>%
  inner_join(., metadata %>% select(SampleID, Individual_Seq_Depth_NonHuman, Ethnic_Group, Village, Description), by = "SampleID") %>%
  mutate(UnalignedReads = round(Individual_Seq_Depth_NonHuman*(UnalignedPct/100)),
         AlignedReads = Individual_Seq_Depth_NonHuman - UnalignedReads, 
         AlignedPct = 100 - UnalignedPct) %>%
  select(SampleID, Individual_Seq_Depth_NonHuman, UnalignedReads, AlignedReads, UnalignedPct, AlignedPct, everything())

humann2_alignment_stats_box <- humann2_alignment_stats %>%
  filter(!str_detect(SampleID, "EXB|LIB|SRR164|10M")) %>%
  ggplot(., aes(Village, AlignedPct, fill = Village)) +
    geom_boxplot() +
    scale_fill_manual(values = Village_colors) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ylab("Percent") +
    ggtitle("Percent of reads aligned by HUMAnN2")
humann2_alignment_stats_box

# ggsave("./05-results.backup/presentation_pdfs/humann2_alignment_stats_box.pdf", plot = humann2_alignment_stats_box, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

humann2_alignment_stats_w <- humann2_alignment_stats %>%
  filter(!str_detect(SampleID, "EXB|LIB|SRR164|10M")) 
pairwise.wilcox.test(humann2_alignment_stats_w$AlignedPct, humann2_alignment_stats_w$Village, p.adjust.method = "fdr")

```

# Pathway analysis

How many pathways were identified in each of the groups? Are there significant
differences?
```{r pathways_boxes}

# all pathways
humann2_path_noblanks_long <- humann2_path %>%
  filter(!str_detect(Pathway, "\\|")) %>%
  select(-CMC032.B0101) %>% # this sample has very few reads
  select(-c("SRR1646032","SRR1646041")) %>% # these plot with EXB and LIB controls in the PCA above
  gather("SampleID","CPM", 2:ncol(.)) %>%
  mutate_if(is.numeric, ~1 * (. != 0)) %>%
  group_by(SampleID) %>%
  summarize(Paths = sum(CPM)) %>%
  inner_join(., metadata %>% select(SampleID, Ethnic_Group, Village), by = "SampleID")

humann2_path_noblanks_long_box <- humann2_path_noblanks_long %>%
  filter(!str_detect(SampleID, "EXB|LIB|SRR164|10M")) %>%
  ggplot(., aes(Village, Paths, fill = Village)) +
    geom_boxplot() +
    scale_fill_manual(values = Village_colors) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ylab("Number of pathways detected") +
    ggtitle("Total number of pathways ID'd by HUMAnN2")
humann2_path_noblanks_long_box

# ggsave("./05-results.backup/presentation_pdfs/humann2_path_noblanks_box.pdf", plot = humann2_path_noblanks_long_box, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

humann2_path_noblanks_long_w <- humann2_path_noblanks_long %>%
  filter(!str_detect(SampleID, "EXB|LIB|SRR164|10M")) 
pairwise.wilcox.test(humann2_path_noblanks_long_w$Paths, humann2_path_noblanks_long_w$Village, p.adjust.method = "fdr")


# potential contaminant pathways removed
humann2_path.decontam_noblanks_long <- humann2_path.decontam %>%
  filter(!str_detect(Pathway, "\\|")) %>%
  gather("SampleID","CPM", 2:ncol(.)) %>%
  mutate_if(is.numeric, ~1 * (. != 0)) %>%
  group_by(SampleID) %>%
  summarize(Paths = sum(CPM)) %>%
  inner_join(., metadata %>% select(SampleID, Ethnic_Group, Village), by = "SampleID")

humann2_path.decontam_noblanks_long_box <- humann2_path.decontam_noblanks_long %>%
  filter(!str_detect(SampleID, "EXB|LIB|10M")) %>%
  ggplot(., aes(Village, Paths, fill = Village)) +
    geom_boxplot() +
    scale_fill_manual(values = Village_colors) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ylab("Number of pathways detected") +
    ggtitle("Number of pathways with potential contaminants removed")
humann2_path.decontam_noblanks_long_box

# ggsave("./05-results.backup/presentation_pdfs/humann2_path.decontam_noblanks_long_box.pdf", plot = humann2_path.decontam_noblanks_long_box, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

humann2_path.decontam_noblanks_long_box_w <- humann2_path.decontam_noblanks_long %>%
  filter(!str_detect(SampleID, "EXB|LIB|10M")) 
pairwise.wilcox.test(humann2_path.decontam_noblanks_long_box_w$Paths, humann2_path.decontam_noblanks_long_box_w$Village, p.adjust.method = "fdr")


# potential contaminant pathways removed and only pathways present in at least 33% of one ethnic group
h2_path_test <- humann2_path.decontam %>%
  filter(!str_detect(Pathway, "\\|")) %>%
  mutate_if(is.numeric, ~1 * (. > 0)) %>%
  gather("SampleID","CPM",2:ncol(.)) %>%
  inner_join(., metadata %>%
               select(SampleID, Ethnic_Group), by = "SampleID") %>%
  group_by(Ethnic_Group, Pathway) %>%
  summarize(Percent_presence = sum(CPM)/length(SampleID)*100) %>%
  # select the SEED proteins that are present in at least 30% of at least one group
  ungroup() %>%
  filter(Percent_presence >= 33.33) %>%
  distinct(Pathway) %>%
  # pull out these proteins to keep
  select(Pathway)

humann2_path.decontam_noblanks_33_long <- humann2_path.decontam %>%
  inner_join(., h2_path_test, by = "Pathway") %>%
  gather("SampleID","CPM", 2:ncol(.)) %>%
  mutate_if(is.numeric, ~1 * (. != 0)) %>%
  group_by(SampleID) %>%
  summarize(Paths = sum(CPM)) %>%
  inner_join(., metadata %>% select(SampleID, Ethnic_Group, Village), by = "SampleID")

humann2_path.decontam_noblanks_33_long_box <- humann2_path.decontam_noblanks_33_long %>%
  filter(!str_detect(SampleID, "EXB|LIB|SRR164|10M")) %>%
  ggplot(., aes(Village, Paths, fill = Village)) +
    geom_boxplot() +
    scale_fill_manual(values = Village_colors) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ylab("Number of pathways detected") +
    ggtitle("Number of pathways w/o contaminants and 33% prevalence")
humann2_path.decontam_noblanks_33_long_box

# ggsave("./05-results.backup/presentation_pdfs/humann2_path.decontam_noblanks_33_long_box.pdf", plot = humann2_path.decontam_noblanks_33_long, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

humann2_path.decontam_noblanks_33_long_w <- humann2_path.decontam_noblanks_33_long %>%
  filter(!str_detect(SampleID, "EXB|LIB|10M")) 
pairwise.wilcox.test(humann2_path.decontam_noblanks_33_long_w$Paths, humann2_path.decontam_noblanks_33_long_w$Village, p.adjust.method = "fdr")


```
HMP and JAE both have significantly more pathways identified than each of the
villages, for all pathways, when potential contaminants are removed, as well as
when potential contaminants are removed and only pathways present in at least
1/3 of one ethnic group are included. There are no differences between the
villages in any of these.

Now plot some PCAs to see the relationship of the samples to eachother and between groups.
We don't need to look at PCAs with the Yanomamis samples b/c we're excluding them. 
```{r pca_paths_noblanks_metadata, eval = F}
# this uses the full pathway table (not decontaminated) and includes the Yanomami samples. We're not including those, so it's not necessary to run
# create the input matrix
humann2_path_noblanks_mat <- humann2_path %>%
  filter(!str_detect(Pathway, "\\|")) %>%
  select(-CMC032.B0101) %>% # this sample has very few reads
  select(-c("SRR1646032","SRR1646041")) %>% # these plot with EXB and LIB controls in the PCA above
  gather("SampleID","CPM", 2:ncol(.)) %>%
  mutate(CPM = CPM + 1) %>%
  spread(Pathway, CPM) %>%
  filter(!str_detect(SampleID, "EXB|LIB|10M")) %>%
  column_to_rownames("SampleID")

# check the number of components to retain by tuning the PCA
tune.humann2_path_pca <- tune.pca(humann2_path_noblanks_mat, logratio = 'CLR')

# perform a PCA to see how the data cluster
humann2_path_noblanks.pca <- pca(humann2_path_noblanks_mat, ncomp = 3, logratio = 'CLR')

# plot the PCA
pca_EG_plot <- plot_pca(humann2_path_noblanks.pca, "PC1", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")
pca_EG_plot
plot_pca(humann2_path_noblanks.pca, "PC3", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")

pca_AG_plot <- plot_pca(humann2_path_noblanks.pca, "PC1", "PC2", "Age_group", "Age_group", "Age_group")
pca_AG_plot

pca_ME_plot <- plot_pca(humann2_path_noblanks.pca, "PC1", "PC2", "Market_economy", "Market_economy", "Market_economy")
pca_ME_plot

pca_S_plot <- plot_pca(humann2_path_noblanks.pca, "PC1", "PC2", "Sex", "Sex", "Sex")
pca_S_plot

pca_TS_plot <- plot_pca(humann2_path_noblanks.pca, "PC1", "PC2", "Tooth_site", "Tooth_site", "Tooth_site")
pca_TS_plot

pca_V_plot <- plot_pca(humann2_path_noblanks.pca, "PC1", "PC2", "Village", "Village", "Village")
pca_V_plot

```
The plaque (HMP) and calculus (JAE) samples separate along PC2, but the CMC
samples have more overlap with JAE than HMP. There is much more spread of the
samples along PC1. Do any other metadata categories appear to drive separation
of the CMC samples along PC1?

Only the tooth site (anterior vs posterior) appears to be separating along PC1.
Let's also run these without the HMP and JAE samples to see how the CMC samples
cluster without industrialized samples included.

```{r adonis_paths_samples, eval = F}
# Use the CLR-transformed matrix from the mixOmics PCA in the section above
# Euclidean distance
humann2_path_noblanks_mat.euc <- vegdist(humann2_path_noblanks.pca$X, method = "euclidean", na.rm = T)
humann2_path_noblanks_mat.euc.pcoa <- ape::pcoa(humann2_path_noblanks_mat.euc)

humann2_path_noblanks_mat.euc.pcoavectors <- as.data.frame(humann2_path_noblanks_mat.euc.pcoa$vectors)
humann2_path_noblanks_mat.euc.pcoavectors <- humann2_path_noblanks_mat.euc.pcoavectors %>%
  rownames_to_column("SampleID")

humann2_path_noblanks_mat.euc.vecmet <- inner_join(humann2_path_noblanks_mat.euc.pcoavectors, metadata, by = "SampleID")

humann2_path_noblanks_mat.euc.vecmet %>%
  ggplot(., aes(Axis.1, Axis.2, colour = Tooth_site, shape = Tooth_site)) +
  geom_point(size = 4) +
  scale_colour_manual(values = Tooth_site_colors) +
  scale_shape_manual(values = Tooth_site_shapes) +
  stat_ellipse(aes(colour = Tooth_site, group = Tooth_site)) +
  theme_minimal(base_size = 7) +
  ggtitle("Tooth site")

# test for difference between different metadata categories 
adonis(humann2_path_noblanks_mat.euc ~ Tooth_site, humann2_path_noblanks_mat.euc.vecmet, perm=999)

adonis2(humann2_path_noblanks_mat.euc ~ ExtractionBatch + LibraryBatch, humann2_path_noblanks_mat.euc.vecmet, permutations = 999, by = "margin")
adonis2(humann2_path_noblanks_mat.euc ~ Village + Sex + Industry + Tooth_site + Ethnic_Group + Age_group + Market_economy, humann2_path_noblanks_mat.euc.vecmet, permutations = 999, by = "margin")

# controlled for tooth site
adonis2(humann2_path_noblanks_mat.euc ~ ExtractionBatch + Village + Sex + Industry + Ethnic_Group + Age_group +  LibraryBatch + Market_economy, humann2_path_noblanks_mat.euc.vecmet, permutations = 999, by = "margin", strata = humann2_path_noblanks_mat.euc.vecmet$Tooth_site)

```

Do the pathway profiles of the Cameroon plaque samples plot closer to industrial
plaque or industrial calculus?
```{r pca_paths_noyanomami_metadata}
# create the input matrix
humann2_path_noyanomami_mat <- humann2_path.decontam %>%
  gather("SampleID","CPM", 2:ncol(.)) %>%
  mutate(CPM = CPM + 1) %>%
  spread(Pathway, CPM) %>%
  filter(!str_detect(SampleID, "EXB|LIB|10M|SRR164")) %>%
  column_to_rownames("SampleID")

# check the number of components to retain by tuning the PCA
tune.humann2_path_pca <- tune.pca(humann2_path_noyanomami_mat, logratio = 'CLR')

# perform a PCA to see how the data cluster
humann2_path_noyanomami.pca <- pca(humann2_path_noyanomami_mat, ncomp = 3, logratio = 'CLR')

# plot the PCA
pca_EG_plot <- plot_pca(humann2_path_noyanomami.pca, "PC1", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")
pca_EG_plot
plot_pca(humann2_path_noyanomami.pca, "PC3", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")

pca_AG_plot <- plot_pca(humann2_path_noyanomami.pca, "PC1", "PC2", "Age_group", "Age_group", "Age_group")
pca_AG_plot

pca_ME_plot <- plot_pca(humann2_path_noyanomami.pca, "PC1", "PC2", "Market_economy", "Market_economy", "Market_economy")
pca_ME_plot

### there's a problem with this one about the size of the points. Using a different size_group lets it work
pca_S_plot <- plot_pca(humann2_path_noyanomami.pca, "PC1", "PC2", "Sex", "Sex", "Sex")
pca_S_plot

pca_TS_plot <- plot_pca(humann2_path_noyanomami.pca, "PC1", "PC2", "Tooth_site", "Tooth_site", "Tooth_site")
pca_TS_plot

pca_V_plot <- plot_pca(humann2_path_noyanomami.pca, "PC1", "PC2", "Village", "Village", "Village")
pca_V_plot

# ggsave("./05-results.backup/presentation_pdfs/pca_EG_noyanomami_paths.pdf", plot = pca_EG_plot, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

```
Some of the samles plot closer to the industrial calculus, but generally they
are distinct, and separate from both industrial sample types along PC1. This
appears to be related to anterior/posterior tooth sampling site, where anterior
samples plot closer to the industrial samples and posterior samples plot further
from industrial samples.

Are there significant differences in beta-diversity between any of the metadata groups?
```{r adonis_paths_noyanomami}
# Use the CLR-transformed matrix from the mixOmics PCA in the section above
# Euclidean distance
humann2_path_noyanomami_mat.euc <- vegdist(humann2_path_noyanomami.pca$X, method = "euclidean", na.rm = T)
humann2_path_noyanomami_mat.euc.pcoa <- ape::pcoa(humann2_path_noyanomami_mat.euc)

humann2_path_noyanomami_mat.euc.pcoavectors <- as.data.frame(humann2_path_noyanomami_mat.euc.pcoa$vectors)
humann2_path_noyanomami_mat.euc.pcoavectors <- humann2_path_noyanomami_mat.euc.pcoavectors %>%
  rownames_to_column("SampleID")

humann2_path_noyanomami_mat.euc.vecmet <- inner_join(humann2_path_noyanomami_mat.euc.pcoavectors, metadata, by = "SampleID")

humann2_path_noyanomami_mat.euc.vecmet %>%
  ggplot(., aes(Axis.1, Axis.2, colour = Tooth_site, shape = Tooth_site)) +
  geom_point(size = 4) +
  scale_colour_manual(values = Tooth_site_colors) +
  scale_shape_manual(values = Tooth_site_shapes) +
  stat_ellipse(aes(colour = Tooth_site, group = Tooth_site)) +
  theme_minimal(base_size = 7) +
  ggtitle("Tooth site")

# test for difference between different metadata categories 
adonis2(humann2_path_noyanomami_mat.euc ~ ExtractionBatch + Village + Sex + Industry + Tooth_site + Ethnic_Group + Age_group +  LibraryBatch + Market_economy, humann2_path_noyanomami_mat.euc.vecmet, permutations = 999, by = "margin")

# controlled for tooth site
adonis2(humann2_path_noyanomami_mat.euc ~ ExtractionBatch + Village + Sex + Industry + Ethnic_Group + Age_group +  LibraryBatch + Market_economy, humann2_path_noyanomami_mat.euc.vecmet, permutations = 999, by = "margin", strata = metadata_noblanks$Tooth_site)

```
Tooth site (p = 0.0001) and village (p < 0.05) are significantly different. When
controlling for tooth site, village is significant (p < 0.05).

What are the relationships of the Cameroon plaque samples to eachother without
the industrial samples?
```{r pca_paths_cmc_metadata}
humann2_path_cmc_mat <- humann2_path.decontam %>%
  gather("SampleID","CPM", 2:ncol(.)) %>%
  mutate(CPM = CPM + 1) %>%
  spread(Pathway, CPM) %>%
  filter(!str_detect(SampleID, "EXB|LIB|SRR|JAE")) %>%
  column_to_rownames("SampleID")

# check the number of components to retain by tuning the PCA
tune.humann2_path_pca <- tune.pca(humann2_path_cmc_mat, logratio = 'CLR')

# perform a PCA to see how the data cluster
humann2_path_cmc.pca <- pca(humann2_path_cmc_mat, ncomp = 3, logratio = 'CLR')

# plot the PCA
pca_EG_plot <- plot_pca(humann2_path_cmc.pca, "PC1", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")
pca_EG_plot

pca_AG_plot <- plot_pca(humann2_path_cmc.pca, "PC1", "PC2", "Age_group", "Age_group", "Age_group")
pca_AG_plot

pca_ME_plot <- plot_pca(humann2_path_cmc.pca, "PC1", "PC2", "Market_economy", "Market_economy", "Market_economy")
pca_ME_plot

pca_S_plot <- plot_pca(humann2_path_cmc.pca, "PC1", "PC2", "Sex", "Sex", "Sex")
pca_S_plot
plot_pca(humann2_path_cmc.pca, "PC3", "PC1", "Sex", "Sex", "Sex")


pca_TS_plot <- plot_pca(humann2_path_cmc.pca, "PC1", "PC2", "Tooth_site", "Tooth_site", "Tooth_site")
pca_TS_plot

pca_V_plot <- plot_pca(humann2_path_cmc.pca, "PC1", "PC2", "Village", "Village", "Village")
pca_V_plot

# ggsave("./05-results.backup/presentation_pdfs/pca_EG_cmc_paths.pdf", plot = pca_EG_plot, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

```
Again the only sort-of clustering that we see is by tooth site (anterior or
posterior).  

Are there significant differences in beta-diversity between any of the metadata groups?
```{r adonis_paths_cmc}
# Use the CLR-transformed matrix from the mixOmics PCA in the section above
# Euclidean distance
humann2_path_cmc_mat.euc <- vegdist(humann2_path_cmc.pca$X, method = "euclidean", na.rm = T)
humann2_path_cmc_mat.euc.pcoa <- ape::pcoa(humann2_path_cmc_mat.euc)

humann2_path_cmc_mat.euc.pcoavectors <- as.data.frame(humann2_path_cmc_mat.euc.pcoa$vectors)
humann2_path_cmc_mat.euc.pcoavectors <- humann2_path_cmc_mat.euc.pcoavectors %>%
  rownames_to_column("SampleID")

humann2_path_cmc_mat.euc.vecmet <- inner_join(humann2_path_cmc_mat.euc.pcoavectors, metadata, by = "SampleID")

humann2_path_cmc_mat.euc.vecmet %>%
  ggplot(., aes(Axis.1, Axis.2, colour = Tooth_site, shape = Tooth_site)) +
  geom_point(size = 4) +
  scale_colour_manual(values = Tooth_site_colors) +
  scale_shape_manual(values = Tooth_site_shapes) +
  # stat_ellipse(aes(colour = Tooth_site, group = Tooth_site)) +
  theme_minimal(base_size = 7) +
  ggtitle("Tooth site")

# test for difference between different metadata categories 
adonis2(humann2_path_cmc_mat.euc ~ ExtractionBatch + Village + Sex + Tooth_site + Ethnic_Group + Age_group +  LibraryBatch + Market_economy, humann2_path_cmc_mat.euc.vecmet, permutations = 999, by = "margin")

# controlled for tooth site
adonis2(humann2_path_cmc_mat.euc ~ ExtractionBatch + Village + Sex + Ethnic_Group + Age_group +  LibraryBatch + Market_economy, humann2_path_cmc_mat.euc.vecmet, permutations = 999, by = "margin", strata = humann2_path_cmc_mat.euc$Tooth_site)

```
Tooth site (p < 0.01) is significantly different, but when controlling for tooth
site none are significantly different.

Now let's look at which pathways are driving separation of the sample groups.
Which of the pathways have the strongest loadings in PC1 and PC2?
```{r pca_path_biplot}
# which paths distinguish sample types when HMP and JAE are included?
pca_EG_biplot_pc1 <- plot_pca_bi(humann2_path_noyanomami.pca, "PC1", "PC2", "Ethnic_Group", PC1, pathways = T)
pca_EG_biplot_pc1

# ggsave("./05-results.backup/presentation_pdfs/pca_EG_bi_noyanomami_paths.pdf", plot = pca_EG_biplot_pc1, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

# what are those pathways?
pandoc.table(pca_EG_biplot_pc1$plot_env$pws_10 %>% select(-Function) %>% arrange(desc(PC1)),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC1 positive loadings")

pandoc.table(pca_EG_biplot_pc1$plot_env$neg_10 %>% select(-Function) %>% arrange(PC1),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC1 negative loadings")

# make a table of the pathways to save, to use again later in another R notebook
# humann2_pathway_biplot_list <- pca_EG_biplot_pc1$plot_env$pws_10 %>% select(-Function) %>% arrange(desc(PC1)) %>% select(pathway) %>% mutate(Direction = "PC1+", Samples = "Industrial+CMC") 
# humann2_pathway_biplot_list <- humann2_pathway_biplot_list %>%
#   bind_rows(pca_EG_biplot_pc1$plot_env$neg_10 %>% select(-Function) %>% arrange(desc(PC1)) %>% select(pathway) %>% mutate(Direction = "PC1-", Samples = "Industrial+CMC"))

# which paths distinguish sample types when HMP and JAE are included?
pca_EG_biplot_pc2 <- plot_pca_bi(humann2_path_noyanomami.pca, "PC1", "PC2", "Ethnic_Group", PC2, pathways = T)
pca_EG_biplot_pc2

# what are those pathways?
pandoc.table(pca_EG_biplot_pc2$plot_env$pws_10 %>% select(-Function) %>% arrange(desc(PC2)),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC2 positive loadings")

pandoc.table(pca_EG_biplot_pc2$plot_env$neg_10 %>% select(-Function) %>% arrange(PC2),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC2 negative loadings")

# humann2_pathway_biplot_list <- humann2_pathway_biplot_list %>%
#   bind_rows(pca_EG_biplot_pc2$plot_env$pws_10 %>% select(-Function) %>% arrange(desc(PC2)) %>% select(pathway) %>% mutate(Direction = "PC2+", Samples = "Industrial+CMC"))
# humann2_pathway_biplot_list <- humann2_pathway_biplot_list %>%
#   bind_rows(pca_EG_biplot_pc2$plot_env$neg_10 %>% select(-Function) %>% arrange(desc(PC2)) %>% select(pathway) %>% mutate(Direction = "PC2-", Samples = "Industrial+CMC"))


# which paths distinguish sample groups when only CMC samples are included?
pca_EG_cmc_biplot_pc1 <- plot_pca_bi(humann2_path_cmc.pca, "PC1", "PC2", "Ethnic_Group", PC1, pathways = T)
pca_EG_cmc_biplot_pc1

# ggsave("./05-results.backup/presentation_pdfs/pca_EG_bi_cmc_paths.pdf", plot = pca_EG_cmc_biplot_pc1, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

# what are those pathways?
pandoc.table(pca_EG_cmc_biplot_pc1$plot_env$pws_10 %>% select(-Function) %>% arrange(desc(PC1)),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC1 positive loadings")

pandoc.table(pca_EG_cmc_biplot_pc1$plot_env$neg_10 %>% select(-Function) %>% arrange(PC1),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC1 negative loadings")

# humann2_pathway_biplot_list <- humann2_pathway_biplot_list %>% 
#   bind_rows(pca_EG_cmc_biplot_pc1$plot_env$pws_10 %>% select(-Function) %>% arrange(desc(PC1)) %>% select(pathway) %>% mutate(Direction = "PC1+", Samples = "CMC"))
# humann2_pathway_biplot_list <- humann2_pathway_biplot_list %>%
#   bind_rows(pca_EG_biplot_pc1$plot_env$neg_10 %>% select(-Function) %>% arrange(desc(PC1)) %>% select(pathway) %>% mutate(Direction = "PC1-", Samples = "CMC"))

pca_EG_cmc_biplot_pc2 <- plot_pca_bi(humann2_path_cmc.pca, "PC1", "PC2", "Ethnic_Group", PC2, pathways = T)
pca_EG_cmc_biplot_pc2

pandoc.table(pca_EG_cmc_biplot_pc1$plot_env$pws_10 %>% select(-Function) %>% arrange(desc(PC2)),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC1 positive loadings")

pandoc.table(pca_EG_cmc_biplot_pc1$plot_env$neg_10 %>% select(-Function) %>% arrange(PC2),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC1 negative loadings")

# humann2_pathway_biplot_list <- humann2_pathway_biplot_list %>% 
#   bind_rows(pca_EG_cmc_biplot_pc1$plot_env$pws_10 %>% select(-Function) %>% arrange(desc(PC2)) %>% select(pathway) %>% mutate(Direction = "PC2+", Samples = "CMC"))
# humann2_pathway_biplot_list <- humann2_pathway_biplot_list %>%
#   bind_rows(pca_EG_biplot_pc1$plot_env$neg_10 %>% select(-Function) %>% arrange(desc(PC2)) %>% select(pathway) %>% mutate(Direction = "PC2-", Samples = "CMC"))
# 
# fwrite(humann2_pathway_biplot_list, file = "./05-results.backup/humann2_pathway_biplot_list.tsv", quote = F, sep = "\t")

```
_*All samples*_
_PC1+:_
PWY-6285: superpathway of fatty acids biosynthesis (E. coli) - standard fatty acid biosynthesis 
PWY-5971: palmitate biosynthesis II (bacteria and plants) - common saturated fatty acid 
PWY-7391: isoprene biosynthesis II (engineered) - volatile compound produced by plants, incorporated into cell membranes in response to heat stress
PWY-6803: phosphatidylcholine acyl editing -  production of triacylglycerols
PWY-6113: superpathway of mycolate biosynthesis - found in mycobacterial cell walls
P122-PWY: heterolactic fermentation - found in Firmicutes, particularly lactic acid bacteria; this fermentation pathway produces substantial amounts of volatile compounds and C02 along with lactic acid; this produces lactate, EtOH, CO2 and trace acetate
PWY-5675: nitrate reduction V (assimilatory) - environmental nitrogen-containing compounds (nitrate, nitrite) must be reduced to be used by bacteria in producing their own N-containing compounds such as amino acids
PWY-5121: superpathway of geranylgeranyl diphosphate biosynthesis II (via MEP) - crucial compound involved in biosynthesis of a variety of terpenes and terpenoids, including central compounds such as ubiquinones and menaquinones
PWY-5910: superpathway of geranylgeranyldiphosphate biosynthesis I (via mevalonate) - crucial compound involved in biosynthesis of a variety of terpenes and terpenoids, including central compounds such as ubiquinones and menaquinones
PWY-6859: all-trans-farnesol biosynthesis - found in eukaryia such as yeast; part of terpenoid synthesis

_PC1-:_
GLYCOCAT-PWY: glycogen degradation I (bacterial) - breaks (internally stored?) long chain carbohydrates into short, processible ones that are further broken down into maltoses and glucose for glycolysis
PWY-7204: pyridoxal5'-phosphate salvage II (plants) - vitamin B6 salvage pathway; plants and bacteria have a similar system; roles in sphingolipid biosynthesis and carbohydrate metabolism
GALACTUROCAT-PWY: D-galacturonate degradation I - breaks down D-galacturonate (common in plant cell walls and the environment) to a modified glucose that enters the Entner-Doudoroff shunt
PWY-5104: L-isoleucine biosynthesis IV - common alternate pathway to synthesize L-isoleucine; known to exist in *Methanobrevibacter* and *Desulfovibrio*.
PWY-6630: superpathway of L-tyrosine biosynthesis - standard pathway
PWY-5005: biotin biosynthesis II - biotin is a cofactor that facilitates transfer of CO2 in fatty acid and carbohydrate metabolism
PWY-5941: glycogen degradation II (eukaryotic) - same idea as GLYCOCAT-PWY, but found in fungi and animals
GALACT-GLUCUROCAT-PWY: superpathway of hexuronide and hexuronate degradation - D-glucuronate, D-fructuronate and D-galacturonate (main monomer of pectin); shuffles these till they can enter the Entner-Doudoroff pathway
PWY-7371: 1,4-dihydroxy-6-naphthoate biosynthesis II - this is converted to menaquinone; menaquinones and demethylmenaquinones (vitamin K2) are reversible redox component of the electron transfer chain; Gram +s, and anaerobic G+ and G- have either; comes from chorismate (intermediate of aromatic amino acid biosynthesis)
PWY1G-0: mycothiol biosynthesis - major thiol found in Actinobacteria (opposed to glutathione); detoxifies stress-inducing factors, maintains reducing envt within the cell

_PC2+:_
PWY0-1241: ADP-L-glycero-&beta;-D-manno-heptose biosynthesis - glycero-manno-heptoses are found in the cell surface polymers of many bacteria, particularly LPS
PWY-6895: superpathway of thiamin diphosphate biosynthesis II - vitamin B1, an essential cofactor for enzymes involved in energy metabolism
THISYN-PWY: superpathway of thiamin diphosphate biosynthesis I - vitamin B1, an essential cofactor for enzymes involved in energy metabolism; different version from above, carried by different organisms
PWY-5083: NAD/NADH phosphorylation and dephosphorylation - very common cofactors
PWY-5838: superpathway of menaquinol-8 biosynthesis I - menaquinones and demethylmenaquinones (vitamin K2) are reversible redox component of the electron transfer chain; Gram +s, and anaerobic G+ and G- have either; menaquinol-8 is common in facultatively anaerobic Gram - organisms
PWY-5861: superpathway of demethylmenaquinol-8 biosynthesis - menaquinones and demethylmenaquinones (vitamin K2) are reversible redox component of the electron transfer chain; Gram +s, and anaerobic G+ and G- have either; dimethylmenaquinol-8 is common in facultatively anaerobic Gram - organisms
PWY-5675: nitrate reduction V (assimilatory) - environmental nitrogen-containing compounds (nitrate, nitrite) must be reduced to be used by bacteria in producing their own N-containing compounds such as amino acids
PWY-6892: thiazole biosynthesis I (E. coli) - only the synthesis of the thiazole moiety of thiamin; part of the thiamin phosphate synthesis pathway (vitamin B1) (E. coli version)
PWY-6891: thiazole biosynthesis II (Bacillus) - only the synthesis of the thiazole moiety of thiamin; part of the thiamin phosphate synthesis pathway (vitamin B1) (Bacillus version)
PWY-4041: gamma-glutamyl cycle - glutathione metabolism; maintains a reduced environment

_PC2-:_
P23-PWY: reductive TCA cycle I - carbon dioxide fixation pathway found in autotrophic eubacteria and archaea; considered to be a primordial pathway for production of starting organic molecules for biosynthesis of sugars, lipids, amino acids, pyrimidines and pyrroles
PWY-5104: L-isoleucine biosynthesis IV - synthesis from propanoate
PWY-7003: glycerol degradation to butanol - butanol biosynthesis pathway from glycerol
PWY4LZ-257: superpathway of fermentation (Chlamydomonas reinhardtii) - fermentation from pyruvate
PWY-6588: pyruvate fermentation to acetone - anaerobic
PWY66-398: TCA cycle III (animals) - energy metabolism
PWY-6269: adenosylcobalamin salvage from cobinamide II - production of cobamides, which are enzyme cofactors; found in bacteria and archaea
PWY-6549: L-glutamine biosynthesis III - amino acid biosynthesis in plants
PWY-7254: TCA cycle VII (acetate-producers) - used by acetic acid bacteria are Gram-negative acidophilic α-proteobacteria that oxidize ethanol to acetate
PWY-5690: TCA cycle II (plants and fungi) - aerobic respiration


_*CMC samples*_
_PC1+:_
GLYCOCAT-PWY: glycogen degradation I (bacterial) - breaks (internally stored?) long chain carbohydrates into short, processible ones that are further broken down into maltoses and glucose for glycolysis
PWY-7204: pyridoxal 5'-phosphate salvage II (plants) - vitamin B6 salvage pathway; plants and bacteria have a similar system; roles in sphingolipid biosynthesis and carbohydrate metabolism
PWY-5005: biotin biosynthesis II - biotin is a cofactor that facilitates transfer of CO2 in fatty acid and carbohydrate metabolism
GALACTUROCAT-PWY: D-galacturonate degradation I - breaks down D-galacturonate (common in plant cell walls and the environment) to a modified glucose that enters the Entner-Doudoroff shunt
PWY-6630: superpathway of L-tyrosine biosynthesis - standard pathway
GALACT-GLUCUROCAT-PWY: superpathway of hexuronide and hexuronate degradation - D-glucuronate, D-fructuronate and D-galacturonate (main monomer of pectin); shuffles these till they can enter the Entner-Doudoroff pathway
RHAMCAT-PWY: L-rhamnose degradation I - common bacterial cell wall component, also used for energy
PWY-5101: L-isoleucine biosynthesis II - synthesis from pyruvate
PWY-5941: glycogen degradation II (eukaryotic) - same idea as GLYCOCAT-PWY, but found in fungi and animals
PWY-7209: superpathway of pyrimidine ribonucleosides degradation - pyrimidine salvage pathway; uridine broken down to uracil and a ribose or to beta-alanine

_PC1-:_
PWY-6803: phosphatidylcholine acyl editing - production of triacylglycerols
PWY-7431: aromatic biogenic amine degradation (bacteria) - synthesis of aromatic and aliphatic biogenic amines by decarboxylation of amino acids; biogenic amines have physiologic roles in prokaryotes (not listed)
PWY-7391: isoprene biosynthesis II (engineered) - volatile compound produced by plants, incorporated into cell membranes in response to heat stress
PWY-5910: superpathway of geranylgeranyldiphosphate biosynthesis I (via mevalonate) - crucial compound involved in biosynthesis of a variety of terpenes and terpenoids, including central compounds such as ubiquinones and menaquinones
PWY-7392: taxadiene biosynthesis (engineered) - taxadiene is an intermediate in taxadiol biosynthesis (from Pacific yew tree, used in chemotherapy)
PWY-5121: superpathway of geranylgeranyl diphosphate biosynthesis II (via MEP) - crucial compound involved in biosynthesis of a variety of terpenes and terpenoids, including central compounds such as ubiquinones and menaquinones
GLYCOLYSIS-TCA-GLYOX-BYPASS: superpathway of glycolysis, pyruvate dehydrogenase, TCA, and glyoxylate bypass - energy metabolism
P122-PWY: heterolactic fermentation - found in Firmicutes, particularly lactic acid bacteria; this fermentation pathway produces substantial amounts of volatile compounds and C02 along with lactic acid; this produces lactate, EtOH, CO2 and trace acetate
PWY-6353: purine nucleotides degradation II (aerobic) - degradation of purine nucleotides to purine nucleosides, purine bases; nitrogen metabolism for nitrogen utilization
TCA-GLYOX-BYPASS: superpathway of glyoxylate bypass and TCA - bypasses steps that lead to loss of CO2, saving carbon

_PC2+:_
PWY-5101: L-isoleucine biosynthesis II - synthesis from pyruvate
GALACTUROCAT-PWY: D-galacturonate degradation I - breaks down D-galacturonate (common in plant cell walls and the environment) to a modified glucose that enters the Entner-Doudoroff shunt
GALACT-GLUCUROCAT-PWY: superpathway of hexuronide and hexuronate degradation - D-glucuronate, D-fructuronate and D-galacturonate (main monomer of pectin); shuffles these till they can enter the Entner-Doudoroff pathway
GLYCOCAT-PWY: glycogen degradation I (bacterial) - breaks (internally stored?) long chain carbohydrates into short, processible ones that are further broken down into maltoses and glucose for glycolysis
PWY-5005: biotin biosynthesis II - biotin is a cofactor that facilitates transfer of CO2 in fatty acid and carbohydrate metabolism
PWY-6630: superpathway of L-tyrosine biosynthesis - standard pathway
PWY-5941: glycogen degradation II (eukaryotic) - same idea as GLYCOCAT-PWY, but found in fungi and animals
RHAMCAT-PWY: L-rhamnose degradation I - common bacterial cell wall component, also used for energy
PWY-7204: pyridoxal 5'-phosphate salvage II (plants) - vitamin B6 salvage pathway; plants and bacteria have a similar system; roles in sphingolipid biosynthesis and carbohydrate metabolism
PWY-7209: superpathway of pyrimidine ribonucleosides degradation - pyrimidine salvage pathway; uridine broken down to uracil and a ribose or to beta-alanine

_PC2-:_
TCA-GLYOX-BYPASS: superpathway of glyoxylate bypass and TCA - bypasses steps that lead to loss of CO2, saving carbon
GLYCOLYSIS-TCA-GLYOX-BYPASS: superpathway of glycolysis, pyruvate dehydrogenase, TCA, and glyoxylate bypass - energy metabolism
PWY-7391: isoprene biosynthesis II (engineered) - volatile compound produced by plants, incorporated into cell membranes in response to heat stress
PWY-7431: aromatic biogenic amine degradation (bacteria) - synthesis of aromatic and aliphatic biogenic amines by decarboxylation of amino acids; biogenic amines have physiologic roles in prokaryotes (not listed)
P122-PWY: heterolactic fermentation - found in Firmicutes, particularly lactic acid bacteria; this fermentation pathway produces substantial amounts of volatile compounds and C02 along with lactic acid; this produces lactate, EtOH, CO2 and trace acetate
PWY-6803: phosphatidylcholine acyl editing - production of triacylglycerols
PWY-5910: superpathway of geranylgeranyldiphosphate biosynthesis I (via mevalonate) - crucial compound involved in biosynthesis of a variety of terpenes and terpenoids, including central compounds such as ubiquinones and menaquinones
PWY-5121: superpathway of geranylgeranyl diphosphate biosynthesis II (via MEP) - crucial compound involved in biosynthesis of a variety of terpenes and terpenoids, including central compounds such as ubiquinones and menaquinones
PWY-7392: taxadiene biosynthesis (engineered) - taxadiene is an intermediate in taxadiol biosynthesis (from Pacific yew tree, used in chemotherapy)
PWY-6353: purine nucleotides degradation II (aerobic) - degradation of purine nucleotides to purine nucleosides, purine bases; nitrogen metabolism for nitrogen utilization


# Community-level functional analysis with individual gene families assigned KEGG Orthology numbers
## KEGG ortholog distribution
We will start analysis of KEGG orthologs by looking at the total number of KOs
assigned to each population to see if this may bias further analyses. Is the
proportion of genefamilies that were re-grouped by HUMAnN2 script into KOs
consistent and high across all sample types? That would be good b/c we have no
major group biases.
```{r orthologs_boxes}
# use the full table
humann2_KO_long <- humann2_KOs %>%
  filter(!str_detect(Ortholog, "\\|")) %>%
  select(-CMC032.B0101) %>% # this sample has very few reads
  select(-c("SRR1646032","SRR1646041")) %>% # these plot with EXB and LIB controls in the PCA above
  gather("SampleID","CPM", 2:ncol(.)) %>%
  mutate_if(is.numeric, ~1 * (. != 0)) %>%
  group_by(SampleID) %>%
  summarize(Paths = sum(CPM)) %>%
  inner_join(., metadata %>% select(SampleID, Ethnic_Group, Village), by = "SampleID")

humann2_KO_box <- humann2_KO_long %>%
  filter(!str_detect(SampleID, "EXB|LIB|SRR164|10M")) %>%
  ggplot(., aes(Village, Paths, fill = Village)) +
    geom_boxplot() +
    scale_fill_manual(values = Village_colors) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ylab("Number of KEGG orthologs detected")
    ggtitle("Total number of KEGG orthologs ID'd by HUMAnN2 (no filtering)")
humann2_KO_box

# ggsave("./05-results.backup/presentation_pdfs/humann2_KO_box.pdf", plot = humann2_KO_box, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

humann2_KO_long_w <- humann2_KO_long %>%
  filter(!str_detect(SampleID, "EXB|LIB|SRR164|10M")) 
pairwise.wilcox.test(humann2_KO_long_w$Paths, humann2_KO_long_w$Village, p.adjust.method = "fdr")

# use the decontaminated table, with KOs identified as potential contaminants removed
humann2_KOs.decontam_long <- humann2_KOs.decontam %>%
  gather("SampleID","CPM", 2:ncol(.)) %>%
  mutate_if(is.numeric, ~1 * (. != 0)) %>%
  group_by(SampleID) %>%
  summarize(Paths = sum(CPM)) %>%
  inner_join(., metadata %>% select(SampleID, Ethnic_Group, Village), by = "SampleID")

humann2_KOs.decontam_box <- humann2_KOs.decontam_long %>%
  filter(!str_detect(SampleID, "EXB|LIB|SRR164|10M")) %>%
  ggplot(., aes(Village, Paths, fill = Village)) +
    geom_boxplot() +
    scale_fill_manual(values = Village_colors) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ylab("Number of KEGG orthologs detected")
    ggtitle("Number of KEGG orthologs ID'd by HUMAnN2 (decontaminated)")
humann2_KOs.decontam_box

# ggsave("./05-results.backup/presentation_pdfs/humann2_KOs.decontam_box.pdf", plot = humann2_KOs.decontam_box, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

humann2_KOs.decontam_long_w <- humann2_KOs.decontam_long %>%
  filter(!str_detect(SampleID, "EXB|LIB|10M")) 
pairwise.wilcox.test(humann2_KOs.decontam_long_w$Paths, humann2_KOs.decontam_long_w$Village, p.adjust.method = "fdr")


```
There are significantly more orthologs identified in HMP samples than each of
the villates, and in JAE samples than each of the villages. There are no
significant difrerences between any villages, or between HMP and JAE samples.
This is true for both the full table and the decontaminated table.

Now perform a PCA to see if there is more distinct clustering of the samples
based on any of the metadata categories than we see with the pathways.
```{r pca_KOs_with_controls}
# create the input matrix
humann2_KO_mat <- humann2_KOs.decontam %>%
  gather("SampleID","CPM", 2:ncol(.)) %>%
  mutate(CPM = CPM + 1) %>%
  spread(Ortholog, CPM) %>%
  column_to_rownames("SampleID")

# check the number of components to retain by tuning the PCA
tune.humann2_KO_pca <- tune.pca(humann2_KO_mat, logratio = 'CLR')

# perform a PCA to see how the data cluster
humann2_KO_allsamples.pca <- pca(humann2_KO_mat, ncomp = 3, logratio = 'CLR')

# plot the PCA
plot_pca(humann2_KO_allsamples.pca, "PC1", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")
plot_pca(humann2_KO_allsamples.pca, "PC3", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")


plot_pca_names(humann2_KO_allsamples.pca, "PC1", "PC2", "Ethnic_Group")
plot_pca_names(humann2_KO_allsamples.pca, "PC3", "PC2", "Ethnic_Group")

```
There are ExtractionGauze samples that plot with the HMP plaque samples, but
these probably had some spit on them as mentioned before. None of the CMC
samples plot with the extraction or library blanks, so none need to be removed
as outliers.

Plot a PCA without the blanks.
```{r pca_KOs_no_blanks}

# create the input matrix
humann2_KO_noblanks_mat <- humann2_KOs.decontam %>%
  gather("SampleID","CPM", 2:ncol(.)) %>%
  mutate(CPM = CPM + 1) %>%
  spread(Ortholog, CPM) %>%
  filter(!str_detect(SampleID, "EXB|LIB")) %>%
  column_to_rownames("SampleID")

# check the number of components to retain by tuning the PCA
tune.humann2_path_pca <- tune.pca(humann2_KO_noblanks_mat, logratio = 'CLR')

# perform a PCA to see how the data cluster
humann2_KO_noblanks.pca <- pca(humann2_KO_noblanks_mat, ncomp = 3, logratio = 'CLR')

# plot the PCA
pca_EG_plot <- plot_pca(humann2_KO_noblanks.pca, "PC1", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")
pca_EG_plot

```
The HMP and JAE samples that were subsampled to 10M reads each are plotting
exactly on top of their counterpart full samples, so we can keep using the full
samples. The difference in sequencing depth isn't responsible for the
differences we observe between the sample types.

How does the reationship of the samples change when we filter out orthologs that
are low abundance or low prevalence?
```{r pca_KOs_filtering}
# filter low-abundance orthologs
humann2_KO_mat <- humann2_KOs.decontam %>%
  gather("SampleID","CPM", 2:ncol(.)) %>%
  spread(Ortholog, CPM) %>%
  filter(!str_detect(SampleID, "EXB|LIB|10M"))%>%
  adorn_totals(where = "row", name = "Sum") %>%
  gather("Ortholog","CPM", 2:ncol(.)) %>%
  spread(SampleID, CPM) %>%
  mutate(Percent = Sum/sum(Sum)*100) %>%
  filter(Percent > 0.01) %>%
  select(-c("Sum","Percent")) %>%
  gather("SampleID","CPM", 2:ncol(.)) %>%
  mutate(CPM = CPM + 1) %>%
  spread(Ortholog, CPM) %>%
  column_to_rownames("SampleID")

# check the number of components to retain by tuning the PCA
tune.humann2_KO_pca <- tune.pca(humann2_KO_mat, logratio = 'CLR')

# perform a PCA to see how the data cluster
humann2_KO_noblanks.pca <- pca(humann2_KO_mat, ncomp = 3, logratio = 'CLR')

# plot the PCA
plot_pca(humann2_KO_noblanks.pca, "PC1", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")


# filter low-prevalence orthologs
h2_KO_lowprev <- humann2_KOs.decontam %>%
  mutate_if(is.numeric, ~1 * (. > 0)) %>%
  gather("SampleID","CPM",2:ncol(.)) %>%
  inner_join(., metadata %>%
               select(SampleID, Ethnic_Group), by = "SampleID") %>%
  group_by(Ethnic_Group, Ortholog) %>%
  summarize(Percent_presence = sum(CPM)/length(SampleID)*100) %>%
  # select the KEGG orthologs that are present in at least 30% of at least one group
  ungroup() %>%
  filter(Percent_presence >= 33.33) %>%
  distinct(Ortholog) %>%
  # pull out these proteins to keep
  select(Ortholog)

# this still has ~3000 orthologs
humann2_KOs.decontam_noblanks_33 <- humann2_KOs.decontam %>%
  inner_join(., h2_KO_lowprev, by = "Ortholog") %>%
  gather("SampleID","CPM",2:ncol(.)) %>%
  mutate(CPM = CPM + 1) %>%
  spread(Ortholog,CPM) %>%
  filter(!str_detect(SampleID, "EXB|LIB|10M")) %>%
  column_to_rownames("SampleID")

# check the number of components to retain by tuning the PCA
tune.pca(humann2_KOs.decontam_noblanks_33, logratio = 'CLR')

# perform a PCA to see how the data cluster
humann2_KOs.decontam_noblanks_33.pca <- pca(humann2_KOs.decontam_noblanks_33, ncomp = 3, logratio = 'CLR')

# plot the PCA
plot_pca(humann2_KOs.decontam_noblanks_33.pca, "PC1", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")


```

```{r pca_filtering_plots}
# decontam and abundance >0.01% filtered plots
plot_pca(humann2_KO_noblanks.pca, "PC1", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")
plot_pca(humann2_KO_noblanks.pca, "PC1", "PC2", "Age_group", "Age_group", "Age_group")
plot_pca(humann2_KO_noblanks.pca, "PC1", "PC2", "Market_economy", "Market_economy", "Market_economy")
plot_pca(humann2_KO_noblanks.pca, "PC1", "PC2", "Sex", "Sex", "Sex")
plot_pca(humann2_KO_noblanks.pca, "PC1", "PC2", "Tooth_site", "Tooth_site", "Tooth_site")
plot_pca(humann2_KO_noblanks.pca, "PC1", "PC2", "Village", "Village", "Village")

# # decontam and prevalence >33% filtered plots
# plot_pca(humann2_KOs.decontam_noblanks_33.pca, "PC1", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")
# plot_pca(humann2_KOs.decontam_noblanks_33.pca, "PC1", "PC2", "Age_group", "Age_group", "Age_group")
# plot_pca(humann2_KOs.decontam_noblanks_33.pca, "PC1", "PC2", "Market_economy", "Market_economy", "Market_economy")
# plot_pca(humann2_KOs.decontam_noblanks_33.pca, "PC1", "PC2", "Sex", "Sex", "Sex")
# plot_pca(humann2_KOs.decontam_noblanks_33.pca, "PC1", "PC2", "Tooth_site", "Tooth_site", "Tooth_site")
# plot_pca(humann2_KOs.decontam_noblanks_33.pca, "PC1", "PC2", "Village", "Village", "Village")

```

```{r mr1_orthologs}
# Now keep only the orthologs that are in MR1 from the factor analysis
mr1_proteins <- fread("./05-results.backup/minres_oblimin_noblanks.fa_KO.tsv")

humann2_KOs.decontam_mr1 <- humann2_KOs.decontam %>%
  select(-matches("EXB|LIB|10M")) %>%
  inner_join(., mr1_proteins %>%
               filter(Residual == "MR1") %>%
               select(Ortholog), by = "Ortholog") %>%
  gather("SampleID","Counts",2:ncol(.)) %>%
  mutate(Counts = Counts + 1) %>%
  spread(Ortholog,Counts) %>%
  column_to_rownames("SampleID")

# check the number of components to retain by tuning the PCA
tune.pca(humann2_KOs.decontam_mr1, logratio = 'CLR')

# perform a PCA to see how the data cluster
humann2_KOs.decontam_mr1.pca <- pca(humann2_KOs.decontam_mr1, ncomp = 3, logratio = 'CLR')

# plot the PCA
plot_pca(humann2_KOs.decontam_mr1.pca, "PC1", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")
plot_pca(humann2_KOs.decontam_mr1.pca, "PC1", "PC2", "Age_group", "Age_group", "Age_group")
plot_pca(humann2_KOs.decontam_mr1.pca, "PC1", "PC2", "Market_economy", "Market_economy", "Market_economy")
plot_pca(humann2_KOs.decontam_mr1.pca, "PC1", "PC2", "Sex", "Sex", "Sex")
plot_pca(humann2_KOs.decontam_mr1.pca, "PC1", "PC2", "Tooth_site", "Tooth_site", "Tooth_site")
plot_pca(humann2_KOs.decontam_mr1.pca, "PC1", "PC2", "Village", "Village", "Village")


```


We'll use the mr1 orthologs, not the abundance cut-off or the prevalence cut-off. Are there
significant differences between metadata groups?
```{r adonis_KOs_noblanks}
# Use the CLR-transformed matrix from the mixOmics PCA in the section above
# Euclidean distance
humann2_KO_mat.euc <- vegdist(humann2_KOs.decontam_mr1.pca$X, method = "euclidean", na.rm = T)
humann2_KO_mat.euc.pcoa <- ape::pcoa(humann2_KO_mat.euc)

humann2_KO_mat.euc.pcoavectors <- as.data.frame(humann2_KO_mat.euc.pcoa$vectors)
humann2_KO_mat.euc.pcoavectors <- humann2_KO_mat.euc.pcoavectors %>%
  rownames_to_column("SampleID")

humann2_KO_mat.euc.vecmet <- inner_join(humann2_KO_mat.euc.pcoavectors, metadata, by = "SampleID")

humann2_KO_mat.euc.vecmet %>%
  ggplot(., aes(Axis.1, Axis.2, colour = Tooth_site, shape = Tooth_site)) +
  geom_point(size = 3) +
  scale_colour_manual(values = Tooth_site_colors) +
  scale_shape_manual(values = Tooth_site_shapes) +
  # stat_ellipse(aes(colour = Tooth_site, group = Tooth_site)) +
  theme_minimal(base_size = 7) +
  ggtitle("Tooth site")

# test for difference between different metadata categories 
adonis2(humann2_KO_mat.euc ~ ExtractionBatch + Village + Sex + Industry + Tooth_site + Ethnic_Group + Age_group +  LibraryBatch + Market_economy, humann2_KO_mat.euc.vecmet, permutations = 999, by = "margin")

# controlled for tooth site
adonis2(humann2_KO_mat.euc ~ ExtractionBatch + Village + Sex + Industry + Ethnic_Group + Age_group +  LibraryBatch + Market_economy, humann2_KO_mat.euc.vecmet, permutations = 999, by = "margin", strata = humann2_KO_mat.euc.vecmet$Tooth_site)

```
When using the MR1 orthologs, there are no significant differences between
groups, both when not controlling for tooth site and when controlling for tooth
site.  If using the abundance cut-off, only tooth site is significantly
different (p < 0.01), and when controlling for tooth site, none are
significantly different.

What is the relationship of the Cameroon plaque samples without the industrial samples?
```{r pca_KOs_cmc_metadata}
# filter low-abundance orthologs
humann2_KO_cmc_mat <- humann2_KOs.decontam %>%
  gather("SampleID","CPM", 2:ncol(.)) %>%
  spread(Ortholog, CPM) %>%
  filter(!str_detect(SampleID, "EXB|LIB|SRR|JAE")) %>%
  adorn_totals(where = "row", name = "Sum") %>%
  gather("Ortholog","CPM", 2:ncol(.)) %>%
  spread(SampleID, CPM) %>%
  mutate(Percent = Sum/sum(Sum)*100) %>%
  filter(Percent > 0.01) %>%
  select(-c("Sum","Percent")) %>%
  gather("SampleID","CPM", 2:ncol(.)) %>%
  mutate(CPM = CPM + 1) %>%
  spread(Ortholog, CPM) %>%
  column_to_rownames("SampleID")

# check the number of components to retain by tuning the PCA
tune.pca(humann2_KO_cmc_mat, logratio = 'CLR')

# perform a PCA to see how the data cluster
humann2_KO_cmc.pca <- pca(humann2_KO_cmc_mat, ncomp = 3, logratio = 'CLR')

# plot the PCA
plot_pca(humann2_KO_cmc.pca, "PC1", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")

# ggsave("./05-results.backup/presentation_pdfs/pca_EG_KOs_cmc.pdf", plot = pca_EG_plot, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)


# filter low-prevalence samples
h2_KO_lowprev <- humann2_KOs.decontam %>%
  select(-matches("EXB|LIB|JAE|SRR")) %>%
  mutate_if(is.numeric, ~1 * (. > 0)) %>%
  gather("SampleID","CPM",2:ncol(.)) %>%
  inner_join(., metadata %>%
               select(SampleID, Ethnic_Group), by = "SampleID") %>%
  group_by(Ethnic_Group, Ortholog) %>%
  summarize(Percent_presence = sum(CPM)/length(SampleID)*100) %>%
  # select the KEGG orthologs that are present in at least 30% of at least one group
  ungroup() %>%
  filter(Percent_presence >= 33.33) %>%
  distinct(Ortholog) %>%
  # pull out these proteins to keep
  select(Ortholog)

humann2_KOs.decontam_cmc_33 <- humann2_KOs.decontam %>%
  inner_join(., h2_KO_lowprev, by = "Ortholog") %>%
  gather("SampleID","CPM",2:ncol(.)) %>%
  mutate(CPM = CPM + 1) %>%
  spread(Ortholog,CPM) %>%
  filter(!str_detect(SampleID, "EXB|LIB|JAE|SRR")) %>%
  column_to_rownames("SampleID")

# check the number of components to retain by tuning the PCA
tune.pca(humann2_KOs.decontam_cmc_33, logratio = 'CLR')

# perform a PCA to see how the data cluster
humann2_KOs.decontam_cmc_33.pca <- pca(humann2_KOs.decontam_cmc_33, ncomp = 3, logratio = 'CLR')

# plot the PCA
plot_pca(humann2_KOs.decontam_cmc_33.pca, "PC1", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")


```

```{r pca_filtering_plots}
# decontam and abundance >0.01% filtered plots
plot_pca(humann2_KO_cmc.pca, "PC1", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")
plot_pca(humann2_KO_cmc.pca, "PC1", "PC2", "Age_group", "Age_group", "Age_group")
plot_pca(humann2_KO_cmc.pca, "PC1", "PC2", "Market_economy", "Market_economy", "Market_economy")
plot_pca(humann2_KO_cmc.pca, "PC1", "PC2", "Sex", "Sex", "Sex")
plot_pca(humann2_KO_cmc.pca, "PC1", "PC2", "Tooth_site", "Tooth_site", "Tooth_site")
plot_pca(humann2_KO_cmc.pca, "PC1", "PC2", "Village", "Village", "Village")

# # decontam and prevalence >33% filtered plots
# plot_pca(humann2_KOs.decontam_cmc_33.pca, "PC1", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")
# plot_pca(humann2_KOs.decontam_cmc_33.pca, "PC1", "PC2", "Age_group", "Age_group", "Age_group")
# plot_pca(humann2_KOs.decontam_cmc_33.pca, "PC1", "PC2", "Market_economy", "Market_economy", "Market_economy")
# plot_pca(humann2_KOs.decontam_cmc_33.pca, "PC1", "PC2", "Sex", "Sex", "Sex")
# plot_pca(humann2_KOs.decontam_cmc_33.pca, "PC1", "PC2", "Tooth_site", "Tooth_site", "Tooth_site")
# plot_pca(humann2_KOs.decontam_cmc_33.pca, "PC1", "PC2", "Village", "Village", "Village")

```

```{r mr1_orthologs_cmc}
# Now keep only the orthologs that are in MR1 from the factor analysis
mr1_orthologs_cmc <- fread("./05-results.backup/minres_oblimin_cmc.fa_KO.tsv")

humann2_KOs.decontam_cmc_mr1 <- humann2_KOs.decontam %>%
  select(-matches("EXB|LIB|JAE|SRR")) %>%
  inner_join(., mr1_orthologs_cmc %>%
               filter(Residual == "MR1") %>%
               select(Ortholog), by = "Ortholog") %>%
  gather("SampleID","Counts",2:ncol(.)) %>%
  mutate(Counts = Counts + 1) %>%
  spread(Ortholog,Counts) %>%
  column_to_rownames("SampleID")

# check the number of components to retain by tuning the PCA
tune.pca(humann2_KOs.decontam_cmc_mr1, logratio = 'CLR')

# perform a PCA to see how the data cluster
humann2_KOs.decontam_cmc_mr1.pca <- pca(humann2_KOs.decontam_cmc_mr1, ncomp = 3, logratio = 'CLR')

# plot the PCA
plot_pca(humann2_KOs.decontam_cmc_mr1.pca, "PC1", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")
plot_pca(humann2_KOs.decontam_cmc_mr1.pca, "PC1", "PC2", "Age_group", "Age_group", "Age_group")
plot_pca(humann2_KOs.decontam_cmc_mr1.pca, "PC1", "PC2", "Market_economy", "Market_economy", "Market_economy")
plot_pca(humann2_KOs.decontam_cmc_mr1.pca, "PC1", "PC2", "Sex", "Sex", "Sex")
plot_pca(humann2_KOs.decontam_cmc_mr1.pca, "PC1", "PC2", "Tooth_site", "Tooth_site", "Tooth_site")
plot_pca(humann2_KOs.decontam_cmc_mr1.pca, "PC1", "PC2", "Village", "Village", "Village")


```


We'll use the orthologs in MR1 from the factor analysis, not the abundance
cut-off or the prevalence cut-off. Are there significant differences between
metadata groups?
```{r adonis_KOs_cmc}
# Use the CLR-transformed matrix from the mixOmics PCA in the section above
# Euclidean distance
humann2_KO_cmc_mat.euc <- vegdist(humann2_KOs.decontam_cmc_mr1.pca$X, method = "euclidean", na.rm = T)
humann2_KO_cmc_mat.euc.pcoa <- ape::pcoa(humann2_KO_cmc_mat.euc)

humann2_KO_cmc_mat.euc.pcoavectors <- as.data.frame(humann2_KO_cmc_mat.euc.pcoa$vectors)
humann2_KO_cmc_mat.euc.pcoavectors <- humann2_KO_cmc_mat.euc.pcoavectors %>%
  rownames_to_column("SampleID")

humann2_KO_cmc_mat.euc.vecmet <- inner_join(humann2_KO_cmc_mat.euc.pcoavectors, metadata, by = "SampleID")

humann2_KO_cmc_mat.euc.vecmet %>%
  ggplot(., aes(Axis.1, Axis.2, colour = Tooth_site, shape = Tooth_site)) +
  geom_point(size = 3) +
  scale_colour_manual(values = Tooth_site_colors) +
  scale_shape_manual(values = Tooth_site_shapes) +
  # stat_ellipse(aes(colour = Tooth_site, group = Tooth_site)) +
  theme_minimal(base_size = 7) +
  ggtitle("Tooth site")

# test for difference between different metadata categories 
adonis2(humann2_KO_cmc_mat.euc ~ ExtractionBatch + Village + Sex + Tooth_site + Ethnic_Group + Age_group +  LibraryBatch + Market_economy, humann2_KO_cmc_mat.euc.vecmet, permutations = 999, by = "margin")

# controlled for tooth site
adonis2(humann2_KO_cmc_mat.euc ~ ExtractionBatch + Village + Sex + Ethnic_Group + Age_group +  LibraryBatch + Market_economy, humann2_KO_cmc_mat.euc.vecmet, permutations = 999, by = "margin", strata = humann2_KO_cmc_mat.euc.vecmet$Tooth_site)

```
When using only the orthologs in MR1 from the factor analysis, none are
significantly different, either when not controlling for tooth site, or when
controlling for tooth site. When using the abundance cut-off, only tooth site is
significantly different (p < 0.01), and when controlling for tooth site, only
sex is significantly different (p < 0.05).


Now to see which of the orthologs have the strongest loadings in PC1 and PC2
```{r pca_KO_biplot}
# which paths distinguish sample types when HMP and JAE are included?
pca_EG_KO_biplot_pc1 <- plot_pca_bi(humann2_KOs.decontam_mr1.pca, "PC1", "PC2", "Ethnic_Group", PC1, pathways = F)
pca_EG_KO_biplot_pc1

# ggsave("./05-results.backup/presentation_pdfs/humann2_KO_bi_noblanks.pdf", plot = pca_EG_KO_biplot_pc1, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

# what are those pathways?
pandoc.table(pca_EG_KO_biplot_pc1$plot_env$pws_10 %>% arrange(desc(PC1)),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC1 positive loadings")

pandoc.table(pca_EG_KO_biplot_pc1$plot_env$neg_10 %>% arrange(PC1),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC1 negative loadings")

# humann2_KO_biplot_list <- pca_EG_KO_biplot_pc1$plot_env$pws_10 %>% arrange(desc(PC1)) %>% select(Function) %>% mutate(Direction = "PC1+", Samples = "Industrial+CMC") 
# humann2_KO_biplot_list <- humann2_KO_biplot_list %>%
#   bind_rows(pca_EG_KO_biplot_pc1$plot_env$neg_10 %>% arrange(desc(PC1)) %>% select(Function) %>% mutate(Direction = "PC1-", Samples = "Industrial+CMC"))

# which paths distinguish sample types when HMP and JAE are included?
pca_EG_biplot_pc2 <- plot_pca_bi(humann2_KOs.decontam_mr1.pca, "PC1", "PC2", "Ethnic_Group", PC2, pathways = F)
pca_EG_biplot_pc2

# what are those pathways?
pandoc.table(pca_EG_biplot_pc2$plot_env$pws_10 %>% arrange(desc(PC2)),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC2 positive loadings")

pandoc.table(pca_EG_biplot_pc2$plot_env$neg_10 %>% arrange(PC2),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC2 negative loadings")

# humann2_KO_biplot_list <- humann2_KO_biplot_list %>%
#   bind_rows(pca_EG_biplot_pc2$plot_env$pws_10 %>% arrange(desc(PC2)) %>% select(Function) %>% mutate(Direction = "PC2+", Samples = "Industrial+CMC"))
# humann2_KO_biplot_list <- humann2_KO_biplot_list %>%
#   bind_rows(pca_EG_biplot_pc2$plot_env$neg_10 %>% arrange(desc(PC2)) %>% select(Function) %>% mutate(Direction = "PC2-", Samples = "Industrial+CMC"))

# which paths distinguish sample groups when only CMC samples are included?
pca_EG_cmc_biplot_pc1 <- plot_pca_bi(humann2_KOs.decontam_cmc_mr1.pca, "PC1", "PC2", "Ethnic_Group", PC1, pathways = F)
pca_EG_cmc_biplot_pc1

# ggsave("./05-results.backup/presentation_pdfs/humann2_KO_bi_cmc.pdf", plot = pca_EG_cmc_biplot_pc1, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

# what are those pathways?
pandoc.table(pca_EG_cmc_biplot_pc1$plot_env$pws_10 %>% arrange(desc(PC1)),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC1 positive loadings")

pandoc.table(pca_EG_cmc_biplot_pc1$plot_env$neg_10 %>% arrange(PC1),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC1 negative loadings")

# humann2_KO_biplot_list <- humann2_KO_biplot_list %>%
#   bind_rows(pca_EG_cmc_biplot_pc1$plot_env$pws_10 %>% arrange(desc(PC1)) %>% select(Function) %>% mutate(Direction = "PC1+", Samples = "CMC"))
# humann2_KO_biplot_list <- humann2_KO_biplot_list %>%
#   bind_rows(pca_EG_cmc_biplot_pc1$plot_env$neg_10 %>% arrange(desc(PC1)) %>% select(Function) %>% mutate(Direction = "PC1-", Samples = "CMC"))

pca_EG_cmc_biplot_pc2 <- plot_pca_bi(humann2_KOs.decontam_cmc_mr1.pca, "PC1", "PC2", "Ethnic_Group", PC2, pathways = F)
pca_EG_cmc_biplot_pc2

pandoc.table(pca_EG_cmc_biplot_pc2$plot_env$pws_10 %>% arrange(desc(PC2)),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC2 positive loadings")

pandoc.table(pca_EG_cmc_biplot_pc2$plot_env$neg_10  %>% arrange(PC2),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC2 negative loadings")

# humann2_KO_biplot_list <- humann2_KO_biplot_list %>%
#   bind_rows(pca_EG_cmc_biplot_pc2$plot_env$pws_10 %>% arrange(desc(PC2)) %>% select(Function) %>% mutate(Direction = "PC2+", Samples = "CMC"))
# humann2_KO_biplot_list <- humann2_KO_biplot_list %>%
#   bind_rows(pca_EG_cmc_biplot_pc2$plot_env$neg_10 %>% arrange(desc(PC2)) %>% select(Function) %>% mutate(Direction = "PC2-", Samples = "CMC"))
# 
# fwrite(humann2_KO_biplot_list, file = "./05-results.backup/humann2_KO_biplot_list.tsv", quote = F, sep = "\t")

```

_*All samples MR1 orthologs*_
_PC1+_
K17320 - putative aldouronate transport system permease protein lplC; membrane transport
K04653 - hydrogenase expression/formation protein HypC
K09777 - uncharacterized protein
K08722 - 5'-deoxynucleotidase yfbR; nucleotide metabolism
K17319 - putative aldouronate transport system permease protein lplB; membrane transport
K11705 - iron/zinc/manganese/copper transport system permease protein mtsC; membrane transport
K02471 - vitamin B12/bleomycin/antimicrobial peptide transport system ATP-binding/permease protein bacA; membrane transport
K05352 - ribitol-5-phosphate 2-dehydrogenase (NADP+) tarJ; carbohydrate metabolism
K04651 - hydrogenase nickel incorporation protein HypA/HybF; chaperones and folding catalysts
K07015 - uncharacterized protein

_PC1-_
K09138 - uncharacterized protein
K00096 - glycerol-1-phosphate dehydrogenase [NAD(P)+] [EC:1.1.1.261] araM, egsA; lipid metabolism
K07131 - uncharacterized protein
K02438 - glycogen debranching enzyme [EC:3.2.1.196] glgX; carbohydrate metabolism
K13541 - cobalt-precorrin 5A hydrolase / precorrin-3B C17-methyltransferase [EC:3.7.1.12 2.1.1.131] cbiGH-cobJ; metabolism of cofactors and vitamins
K03795 - sirohydrochlorin cobaltochelatase [EC:4.99.1.3] cbiX; metabolism of cofactors and vitamins
K07101 - uncharacterized protein
K07653 - two-component system, OmpR family, sensor histidine kinase MprB [EC:2.7.13.3] mprB; membrane transport
K07979 - GntR family transcriptional regulator ytrA; transcription factors
K05567 - multicomponent Na+:H+ antiporter subunit C mnhC, mrpC; membrane transport

_PC2+_
K09791 - uncharacterized protein
K09858 - SEC-C motif domain protein; poorly characterized (?)
K07062 - toxin FitB [EC:3.1.-.-]; hydrolase
K02783 - glucitol/sorbitol PTS system EIIC component srlA; carbohydrate metabolism
K00658 - 2-oxoglutarate dehydrogenase E2 component (dihydrolipoamide succinyltransferase) [EC:2.3.1.61] DLST, sucB; carbohydrate metabolism
K06221 - 2,5-diketo-D-gluconate reductase A [EC:1.1.1.346] dkgA; Oxidoreductases
K06980 - tRNA-modifying protein YgfZ; transfer RNA biogenesis
K03684 - ribonuclease D [EC:3.1.13.5] rnd; transfer RNA biogenesis
K00982 - [glutamine synthetase] adenylyltransferase / [glutamine synthetase]-adenylyl-L-tyrosine phosphorylase [EC:2.7.7.42 2.7.7.89] glnE; Nucleotidyltransferases
K03578 - ATP-dependent helicase HrpA [EC:3.6.4.13]; DNA replication and repair

_PC2-_
K11784 - cyclic dehypoxanthinyl futalosine synthase [EC:1.21.98.1] mqnC; metabolism of cofactors and vitamins
K18285 - aminodeoxyfutalosine synthase [EC:2.5.1.120] mqnE; metabolism of cofactors and vitamins
K06001 - tryptophan synthase beta chain [EC:4.2.1.20] trpB; amino acid metabolism
K02032 -  peptide/nickel transport system ATP-binding protein ddpF; membrane transport
K00175 - 2-oxoglutarate/2-oxoacid ferredoxin oxidoreductase subunit beta [EC:1.2.7.3 1.2.7.11] korB, oorB, oforB; citrate cycle
K00852 - ribokinase [EC:2.7.1.15] rbsK, RBKS; carbohydrate metabolism
K02401 - flagellar biosynthetic protein FlhB; cell motility
K16789 - thiamine transporter thiT; transport
K15876 - cytochrome c nitrite reductase small subunit nrfH; energy metabolism
K01950 - NAD+ synthase (glutamine-hydrolysing) [EC:6.3.5.1] NADSYN1, QNS1, nadE; metabolism of cofactors and vitamins

_*CMC samples only*_
_PC1+_
K03406 - methyl-accepting chemotaxis protein mcp; signal transduction
K00266 - glutamate synthase (NADPH) small chain [EC:1.4.1.13] gltD; energy metabolism (N fixation)
K10206 - LL-diaminopimelate aminotransferase [EC:2.6.1.83]; amino acid metabolims
K09777 - uncharacterized protein; poorly characterized (?)
K01678 - fumarate hydratase subunit beta [EC:4.2.1.2] fumB; carbon fixation
K02499 - tetrapyrrole methylase family protein / MazG family protein yabN; DNA replication
K11784 - cyclic dehypoxanthinyl futalosine synthase [EC:1.21.98.1] mqnC; metabolism of cofactors and vitamins
K06409 - stage V sporulation protein B spoVB; transporter
K18285 - aminodeoxyfutalosine synthase [EC:2.5.1.120] mqnE; metabolism of cofactors and vitamins
K09691 - lipopolysaccharide transport system ATP-binding protein ABC-2.LPSE.A; membrane transport

_PC1-_
K09791 - uncharacterized protein
K02783 - glucitol/sorbitol PTS system EIIC component srlA; carbohydrate metabolism
K00096 - glycerol-1-phosphate dehydrogenase [NAD(P)+] [EC:1.1.1.261] araM, egsA; lipid metabolism
K16163 - maleylpyruvate isomerase [EC:5.2.1.4]; amino acid metabolism
K07101 - uncharacterized protein
K07131 - uncharacterized protein
K02438 - glycogen debranching enzyme [EC:3.2.1.196] glgX; carbohydrate metabolism
K03795 - sirohydrochlorin cobaltochelatase [EC:4.99.1.3] cbiX; metabolism of cofactors and vitamins
K09138 - uncharacterized protein
K07979 - GntR family transcriptional regulator ytrA; transcription factor

_PC2+_
K03406 - methyl-accepting chemotaxis protein mcp; signal transduction
K02406 - flagellin fliC; cell motility
K02392 - flagellar basal-body rod protein FlgG; coll motility
K07138 - uncharacterized protein
K02385 - flagellar protein FlbD; cell motility
K02241 - competence protein ComFB; competence
K10907 - aminotransferase [EC:2.6.1.-]; amino acid transferrase
K02588 - nitrogenase iron protein NifH; energy metabolism (N metabolism)
K11358 - aspartate aminotransferase [EC:2.6.1.1] yhdR; amino acid metabolism
K06284 - AbrB family transcriptional regulator, transcriptional pleiotropic regulator of transition state genes abrB; transcription factor

_PC2-_
K03571 - rod shape-determining protein MreD; DNA replication and repair
K02821 - ascorbate PTS system EIIA or EIIAB component [EC:2.7.1.194] ulaC, sgaA; carbohydrate metabolism
K09780 - uncharacterized protein
K05337 - ferredoxin fer; energy metabolism
K02794 - mannose PTS system EIIAB component [EC:2.7.1.191] manX; carbohydrate metabolism
K11072 - spermidine/putrescine transport system ATP-binding protein [EC:7.6.2.11] potA; membrane transport
K07461 - putative endonuclease; DNA replication and repair
K03801 - lipoyl(octanoyl) transferase [EC:2.3.1.181] lipB; metabolism of cofactors and vitamins
K06920 - 7-cyano-7-deazaguanine synthase [EC:6.3.4.20] queC; metabolism of cofactors and vitamins
K00991 - 2-C-methyl-D-erythritol 4-phosphate cytidylyltransferase [EC:2.7.7.60] ispd; metabolism of terpenoids and polyketides



*_Abundance cut-off_* - *Needs to be checked, abundance cut-off calculation was wrong earlier*
_*All samples*_
_PC1+:_
K01839 - phosphopentomutase [EC:5.4.2.7]; carbohydrate metabolism
K01989 - putative ABC transport system substrate-binding protein; membrane transport
K01819 - galactose-6-phosphate isomerase [EC:5.3.1.26]; carbohydrate metabolism
K07088 - uncharacterized protein
K01223 - 6-phospho-beta-glucosidase [EC:3.2.1.86]; carbohydrate metabolism
K03839 - flavodoxin I; energy metabolism
K15986 - manganese-dependent inorganic pyrophosphatase [EC:3.6.1.1]; energy metabolism
K03650 - tRNA modification GTPase [EC:3.6.-.-]; tRNA biosynthesis
K03569 - rod shape-determining protein MreB and related proteins
K15580 - oligopeptide transport system substrate-binding protein; membrane transport

_PC1-:_
K09791 - uncharacterized protein (ycaR)
K07481 - transposase, IS5 family
K01494 - dCTP deaminase [EC:3.5.4.13]; nucleotide metabolism
K03801 - lipoyl(octanoyl) transferase [EC:2.3.1.181]; metabolism of cofactors and vitamins
K00033 - 6-phosphogluconate dehydrogenase [EC:1.1.1.44 1.1.1.343]; carbohydrate metabolism
K07054 - uncharacterized protein (ypfJ)
K02897 - large subunit ribosomal protein L25; translation
K02050 - NitT/TauT family transport system permease protein; membrane transport
K00340 - NADH-quinone oxidoreductase subunit K [EC:7.1.1.2]; energy metabolism
K02437 - glycine cleavage system H protein; carbohydrate metabolism

_PC2+:_
K02032 - peptide/nickel transport system ATP-binding protein; membrane transport
K07114 - Ca-activated chloride channel homolog; membrane transport
K07133 - uncharacterized protein
K11710 - manganese/zinc/iron transport system ATP- binding protein [EC:7.2.2.5]; membrane transport
K02031 - peptide/nickel transport system ATP-binding protein; membrane transport
K10441 - ribose transport system ATP-binding protein [EC:7.5.2.7]; membrane transport
K03924 - MoxR-like ATPase [EC:3.6.3.-]
K02034 - peptide/nickel transport system permease protein; membrane transport
K02033 - peptide/nickel transport system permease protein; membrane transport
K02026 - multiple sugar transport system permease protein; membrane transport

_PC2-:_
K07729 - putative transcriptional regulator; transcription factor
K06199 - fluoride exporter; membrane transport
K02825 - pyrimidine operon attenuation protein / uracil phosphoribosyltransferase [EC:2.4.2.9]; nucleotide metabolism
K03621 - phosphate acyltransferase [EC:2.3.1.274]; lipid metabolism
K00991 - 2-C-methyl-D-erythritol 4-phosphate cytidylyltransferase [EC:2.7.7.60]; biosynthesis of secondary metabolites
K02793 - mannose PTS system EIIA component [EC:2.7.1.191]; carbohydrate metabolism
K02794 - mannose PTS system EIIAB component [EC:2.7.1.191]; carbohydrate metabolism
K07166 - ACT domain-containing protein; signaling protein
K01091 - phosphoglycolate phosphatase [EC:3.1.3.18]; carbohydrate metabolism
K07058 - membrane protein (yhjD, yihH, rbn)

_*CMC samples only*_
_PC1+:_
K00260 - glutamate dehydrogenase [EC:1.4.1.2]; amino acid metabolism, arginine biosynthesis
K01819 - galactose-6-phosphate isomerase [EC:5.3.1.26]; carbohydrate metabolism
K01839 - phosphopentomutase [EC:5.4.2.7]; carbohydrate metabolism
K03856 - 3-deoxy-7-phosphoheptulonate synthase [EC:2.5.1.54]; amino acid metabolism
K01989 - putative ABC transport system substrate-binding protein
K01223 - 6-phospho-beta-glucosidase [EC:3.2.1.86]; carbohydrate metabolism
K07088 - uncharacterized protein
K16957 - L-cystine transport system substrate-binding protein; membrane transport
K15986 - manganese-dependent inorganic pyrophosphatase [EC:3.6.1.1]; energy metabolism
K04486 - histidinol-phosphatase (PHP family) [EC:3.1.3.15]; amino acid metabolism

_PC1-:_
K09791 - uncharacterized protein (ycaR)
K07481 - transposase, IS5 family
K07054 - uncharacterized protein (ypfJ)
K01494 - dCTP deaminase [EC:3.5.4.13]; nucleotide metabolism
K04759 - ferrous iron transport protein B; membrane transport
K02050 - NitT/TauT family transport system permease protein
K10441 - ribose transport system ATP-binding protein [EC:7.5.2.7]; membrane transport
K02027 - multiple sugar transport system substrate-binding protein (ycjN); membrane transport
K02897 - large subunit ribosomal protein L25
K00340 - NADH-quinone oxidoreductase subunit K [EC:7.1.1.2]; energy metabolism

_PC2+:_
K00260 - glutamate dehydrogenase [EC:1.4.1.2]; energy metabolism, amino acid metabolism
K04031 - ethanolamine utilization protein EutS
K01623 - fructose-bisphosphate aldolase, class I [EC:4.1.2.13]; carbohydrate metabolism
K11708 - manganese/zinc/iron transport system permease protein; membrane transport
K02032 - peptide/nickel transport system ATP-binding protein; membrane transport
K02009 - cobalt/nickel transport protein; membrane transport
K01995 - branched-chain amino acid transport system ATP-binding protein; membrane transport
K03839 - flavodoxin I; eneregy metabolism
K11707 - manganese/zinc/iron transport system substrate-binding protein; membrane transport
K03856 - 3-deoxy-7-phosphoheptulonate synthase [EC:2.5.1.54]; amino acid metabolism

_PC2-:_
K03060 - DNA-directed RNA polymerase subunit omega [EC:2.7.7.6]; transcription
K02036 - phosphate transport system ATP-binding protein [EC:7.3.2.1]; membrane transport
K03402 - transcriptional regulator of arginine metabolism; transcription factor
K01626 - 3-deoxy-7-phosphoheptulonate synthase [EC:2.5.1.54]; amino acid metabolism
K00969 - nicotinate-nucleotide adenylyltransferase [EC:2.7.7.18]; metabolism of cofactors and vitamins
K07166 - ACT domain-containing protein; signaling protein (gvcR)
K02529 - LacI family transcriptional regulator; transcription factor
K02970 - small subunit ribosomal protein S21
K05896 - segregation and condensation protein A; DNA replication
K00016 - L-lactate dehydrogenase [EC:1.1.1.27]; carbohydrate metabolism

## KEGG Carbohydrate orthologs
Just because, let's look at only the orthologs in KEGG Carbohydrate pathways. 
```{r carbohydrate_KOs}
kegg_carbohydrate_KOs <- fread("./05-results.backup/kegg_carbohydrates.tsv")
kegg_carbohydrate_KOs <- kegg_carbohydrate_KOs %>% rename(Ortholog = KOrtholog)
```

Industrial and non-industrial samples
```{r carb_KOs_pca_noblanks}

humann2_KO_carbs <- humann2_KOs.decontam %>%
  select(-matches("EXB|LIB|10M")) %>%
  inner_join(., kegg_carbohydrate_KOs %>%
               select(Ortholog) %>% distinct(), by = "Ortholog") %>%
  gather("SampleID","CPM",2:ncol(.)) %>%
  mutate(CPM = CPM + 1) %>%
  spread(Ortholog,CPM) %>%
  column_to_rownames("SampleID")

# check the number of components to retain by tuning the PCA
tune.pca(humann2_KO_carbs, logratio = 'CLR')

# perform a PCA to see how the data cluster
humann2_KO_carbs.pca <- pca(humann2_KO_carbs, ncomp = 3, logratio = 'CLR')

# plot the PCA
plot_pca(humann2_KO_carbs.pca, "PC1", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")
plot_pca(humann2_KO_carbs.pca, "PC1", "PC2", "Age_group", "Age_group", "Age_group")
plot_pca(humann2_KO_carbs.pca, "PC1", "PC2", "Market_economy", "Market_economy", "Market_economy")
plot_pca(humann2_KO_carbs.pca, "PC1", "PC2", "Sex", "Sex", "Sex")
plot_pca(humann2_KO_carbs.pca, "PC1", "PC2", "Tooth_site", "Tooth_site", "Tooth_site")
plot_pca(humann2_KO_carbs.pca, "PC1", "PC2", "Village", "Village", "Village")

```

Are there significant differences between metadata groups?
```{r adonis_carb_KOs_cmc}
# Use the CLR-transformed matrix from the mixOmics PCA in the section above
# Euclidean distance
humann2_KO_carbs_mat.euc <- vegdist(humann2_KO_carbs.pca$X, method = "euclidean", na.rm = T)
humann2_KO_carbs_mat.euc.pcoa <- ape::pcoa(humann2_KO_carbs_mat.euc)

humann2_KO_carbs_mat.euc.pcoavectors <- as.data.frame(humann2_KO_carbs_mat.euc.pcoa$vectors)
humann2_KO_carbs_mat.euc.pcoavectors <- humann2_KO_carbs_mat.euc.pcoavectors %>%
  rownames_to_column("SampleID")

humann2_KO_carbs_mat.euc.vecmet <- inner_join(humann2_KO_carbs_mat.euc.pcoavectors, metadata, by = "SampleID")

humann2_KO_carbs_mat.euc.vecmet %>%
  ggplot(., aes(Axis.1, Axis.2, colour = Tooth_site, shape = Tooth_site)) +
  geom_point(size = 3) +
  scale_colour_manual(values = Tooth_site_colors) +
  scale_shape_manual(values = Tooth_site_shapes) +
  # stat_ellipse(aes(colour = Tooth_site, group = Tooth_site)) +
  theme_minimal(base_size = 7) +
  ggtitle("Tooth site")

# test for difference between different metadata categories 
adonis2(humann2_KO_carbs_mat.euc ~ ExtractionBatch + Village + Sex + Industry + Tooth_site + Ethnic_Group + Age_group +  LibraryBatch + Market_economy, humann2_KO_carbs_mat.euc.vecmet, permutations = 999, by = "margin")

# controlled for tooth site
adonis2(humann2_KO_carbs_mat.euc ~ ExtractionBatch + Village + Sex + Industry + Ethnic_Group + Age_group +  LibraryBatch + Market_economy, humann2_KO_carbs_mat.euc.vecmet, permutations = 999, by = "margin", strata = humann2_KO_carbs_mat.euc.vecmet$Tooth_site)

```
Only tooth site is significantly different (p < 0.01). When controlling for
tooth site, no categories are different.


Cameroon plaque samples only
```{r carb_KOs_pca_cmc}

humann2_KO_cmc_carbs <- humann2_KOs.decontam %>%
  select(-matches("EXB|LIB|JAE|SRR")) %>%
  inner_join(., kegg_carbohydrate_KOs %>%
               select(Ortholog) %>% distinct(), by = "Ortholog") %>%
  gather("SampleID","CPM",2:ncol(.)) %>%
  mutate(CPM = CPM + 1) %>%
  spread(Ortholog,CPM) %>%
  column_to_rownames("SampleID")

# check the number of components to retain by tuning the PCA
tune.pca(humann2_KO_cmc_carbs, logratio = 'CLR')

# perform a PCA to see how the data cluster
humann2_KO_cmc_carbs.pca <- pca(humann2_KO_cmc_carbs, ncomp = 3, logratio = 'CLR')

# plot the PCA
plot_pca(humann2_KO_cmc_carbs.pca, "PC1", "PC2", "Ethnic_Group", "Ethnic_Group", "Ethnic_Group")
plot_pca(humann2_KO_cmc_carbs.pca, "PC1", "PC2", "Age_group", "Age_group", "Age_group")
plot_pca(humann2_KO_cmc_carbs.pca, "PC1", "PC2", "Market_economy", "Market_economy", "Market_economy")
plot_pca(humann2_KO_cmc_carbs.pca, "PC1", "PC2", "Sex", "Sex", "Sex")
plot_pca(humann2_KO_cmc_carbs.pca, "PC1", "PC2", "Tooth_site", "Tooth_site", "Tooth_site")
plot_pca(humann2_KO_cmc_carbs.pca, "PC1", "PC2", "Village", "Village", "Village")

```

Are there significant differences between metadata groups?
```{r adonis_carb_KOs_cmc}
# Use the CLR-transformed matrix from the mixOmics PCA in the section above
# Euclidean distance
humann2_KO_cmc_carbs_mat.euc <- vegdist(humann2_KO_cmc_carbs.pca$X, method = "euclidean", na.rm = T)
humann2_KO_cmc_carbs_mat.euc.pcoa <- ape::pcoa(humann2_KO_cmc_carbs_mat.euc)

humann2_KO_cmc_carbs_mat.euc.pcoavectors <- as.data.frame(humann2_KO_cmc_carbs_mat.euc.pcoa$vectors)
humann2_KO_cmc_carbs_mat.euc.pcoavectors <- humann2_KO_cmc_carbs_mat.euc.pcoavectors %>%
  rownames_to_column("SampleID")

humann2_KO_cmc_carbs_mat.euc.vecmet <- inner_join(humann2_KO_cmc_carbs_mat.euc.pcoavectors, metadata, by = "SampleID")

humann2_KO_cmc_carbs_mat.euc.vecmet %>%
  ggplot(., aes(Axis.1, Axis.2, colour = Tooth_site, shape = Tooth_site)) +
  geom_point(size = 3) +
  scale_colour_manual(values = Tooth_site_colors) +
  scale_shape_manual(values = Tooth_site_shapes) +
  # stat_ellipse(aes(colour = Tooth_site, group = Tooth_site)) +
  theme_minimal(base_size = 7) +
  ggtitle("Tooth site")

# test for difference between different metadata categories 
adonis2(humann2_KO_cmc_carbs_mat.euc ~ ExtractionBatch + Village + Sex + Tooth_site + Ethnic_Group + Age_group +  LibraryBatch + Market_economy, humann2_KO_cmc_carbs_mat.euc.vecmet, permutations = 999, by = "margin")

# controlled for tooth site
adonis2(humann2_KO_cmc_carbs_mat.euc ~ ExtractionBatch + Village + Sex + Ethnic_Group + Age_group +  LibraryBatch + Market_economy, humann2_KO_cmc_carbs_mat.euc.vecmet, permutations = 999, by = "margin", strata = humann2_KO_cmc_carbs_mat.euc.vecmet$Tooth_site)

```
Only tooth site is significantly different (p < 0.01). When controlling for
tooth site, no categories are different.

Now to see which of the orthologs have the strongest loadings in PC1 and PC2
```{r pca_KO_biplot}
# which paths distinguish sample types when HMP and JAE are included?
pca_EG_KO_biplot_pc1 <- plot_pca_bi(humann2_KO_carbs.pca, "PC1", "PC2", "Ethnic_Group", PC1, pathways = F)
pca_EG_KO_biplot_pc1

# ggsave("./05-results.backup/presentation_pdfs/humann2_KO_bi_noblanks.pdf", plot = pca_EG_KO_biplot_pc1, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

# what are those pathways?
pandoc.table(pca_EG_KO_biplot_pc1$plot_env$pws_10 %>% arrange(desc(PC1)),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC1 positive loadings")

pandoc.table(pca_EG_KO_biplot_pc1$plot_env$neg_10 %>% arrange(PC1),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC1 negative loadings")

humann2_KO_carbs_biplot_list <- pca_EG_KO_biplot_pc1$plot_env$pws_10 %>% arrange(desc(PC1)) %>% select(Function) %>% mutate(Direction = "PC1+", Samples = "Industrial+CMC")
humann2_KO_carbs_biplot_list <- humann2_KO_carbs_biplot_list %>%
  bind_rows(pca_EG_KO_biplot_pc1$plot_env$neg_10 %>% arrange(desc(PC1)) %>% select(Function) %>% mutate(Direction = "PC1-", Samples = "Industrial+CMC"))

# which paths distinguish sample types when HMP and JAE are included?
pca_EG_biplot_pc2 <- plot_pca_bi(humann2_KO_carbs.pca, "PC1", "PC2", "Ethnic_Group", PC2, pathways = F)
pca_EG_biplot_pc2

# what are those pathways?
pandoc.table(pca_EG_biplot_pc2$plot_env$pws_10 %>% arrange(desc(PC2)),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC2 positive loadings")

pandoc.table(pca_EG_biplot_pc2$plot_env$neg_10 %>% arrange(PC2),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC2 negative loadings")

humann2_KO_carbs_biplot_list <- humann2_KO_carbs_biplot_list %>%
  bind_rows(pca_EG_biplot_pc2$plot_env$pws_10 %>% arrange(desc(PC2)) %>% select(Function) %>% mutate(Direction = "PC2+", Samples = "Industrial+CMC"))
humann2_KO_carbs_biplot_list <- humann2_KO_carbs_biplot_list %>%
  bind_rows(pca_EG_biplot_pc2$plot_env$neg_10 %>% arrange(desc(PC2)) %>% select(Function) %>% mutate(Direction = "PC2-", Samples = "Industrial+CMC"))

# which paths distinguish sample groups when only CMC samples are included?
pca_EG_cmc_biplot_pc1 <- plot_pca_bi(humann2_KO_cmc_carbs.pca, "PC1", "PC2", "Ethnic_Group", PC1, pathways = F)
pca_EG_cmc_biplot_pc1

# ggsave("./05-results.backup/presentation_pdfs/humann2_KO_bi_cmc.pdf", plot = pca_EG_cmc_biplot_pc1, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

# what are those pathways?
pandoc.table(pca_EG_cmc_biplot_pc1$plot_env$pws_10 %>% arrange(desc(PC1)),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC1 positive loadings")

pandoc.table(pca_EG_cmc_biplot_pc1$plot_env$neg_10 %>% arrange(PC1),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC1 negative loadings")

humann2_KO_carbs_biplot_list <- humann2_KO_carbs_biplot_list %>%
  bind_rows(pca_EG_cmc_biplot_pc1$plot_env$pws_10 %>% arrange(desc(PC1)) %>% select(Function) %>% mutate(Direction = "PC1+", Samples = "CMC"))
humann2_KO_carbs_biplot_list <- humann2_KO_carbs_biplot_list %>%
  bind_rows(pca_EG_cmc_biplot_pc1$plot_env$neg_10 %>% arrange(desc(PC1)) %>% select(Function) %>% mutate(Direction = "PC1-", Samples = "CMC"))

pca_EG_cmc_biplot_pc2 <- plot_pca_bi(humann2_KO_cmc_carbs.pca, "PC1", "PC2", "Ethnic_Group", PC2, pathways = F)
pca_EG_cmc_biplot_pc2

pandoc.table(pca_EG_cmc_biplot_pc2$plot_env$pws_10 %>% arrange(desc(PC2)),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC2 positive loadings")

pandoc.table(pca_EG_cmc_biplot_pc2$plot_env$neg_10  %>% arrange(PC2),
             digits = 5,
             keep.trailing.zeros = T,
             justify = "lccr",
             split.tables = Inf,
             caption = "PC2 negative loadings")

humann2_KO_carbs_biplot_list <- humann2_KO_carbs_biplot_list %>%
  bind_rows(pca_EG_cmc_biplot_pc2$plot_env$pws_10 %>% arrange(desc(PC2)) %>% select(Function) %>% mutate(Direction = "PC2+", Samples = "CMC"))
humann2_KO_carbs_biplot_list <- humann2_KO_carbs_biplot_list %>%
  bind_rows(pca_EG_cmc_biplot_pc2$plot_env$neg_10 %>% arrange(desc(PC2)) %>% select(Function) %>% mutate(Direction = "PC2-", Samples = "CMC"))

# fwrite(humann2_KO_biplot_list, file = "./05-results.backup/humann2_KO_biplot_list.tsv", quote = F, sep = "\t")

humann2_KO_carbs_biplot_list <- humann2_KO_carbs_biplot_list %>%
  rename(Ortholog = Function) %>%
  inner_join(., kegg_carbohydrate_KOs, by = "Ortholog")

# here look at which pathways are most dominant in each group. None are specific to CMC samples
humann2_KO_carbs_biplot_list %>%
  group_by(Samples, Direction, Pathway_name) %>%
  summarize(Counts = length(Pathway_name)) %>%
  arrange(Pathway_name)

```

_*All samples*_
_PC1+:_
K01848
K15633
K00209
K01678
K01849
K03737
K00030
K01711
K01571
K01858

_PC1-:_
K02821
K02783
K01433
K03335
K02782
K00068
K13788
K07248
K00010
K09788

_PC2+:_
K00008
K02438
K17489
K13922
K16148
K06044
K03338
K05341
K03336
K01236

_PC2-:_
K18369
K00123
K01792
K01654
K11258
K03781
K00246
K01092
K00127
K01512

_*CMC samples only*_
_PC1+:_
K01848
K01849
K15633
K01678
K03737
K01220
K00030
K02787
K02788
K00209

_PC1-:_
K02783
K02821
K01433
K00854
K03079
K00694
K03078
K02438
K03335
K02782

_PC2+:_
K00009
K01006
K17489
K00008
K01218
K01580
K03737
K03336
K00692
K03416

_PC2-:_
K07248
K01092
K18369
K01595
K03777
K00029
K01690
K00121
K01069
K09788
















