---
title: "Cameroon hunter-gatherer calculus/plaque MALT"
author: "Irina Velsko"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
    github_document:
        fig_width: 8
        fig_height: 6
        toc: true
---


```{r load_libraries, echo = F, message = F}
library(knitr)
library(decontam)
library(data.table)
library(psych)
library(mixOmics)
library(compositions)
library(vegan)
library(janitor)
library(tidyverse)
library(gplots)
library(ggrepel)
library(viridis)
library(cowplot)
opts_chunk$set(echo=F, warning=F, message=F, dpi=150)
```

Set the working directory of the notebook to the parent directory of the 
notebook itself.  

```{r setup}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

# MALT read stats
The number of reads assigned in each sample appears very low (~50%). We want to check this so we know how many were assinged in each sample. We used the cleaned and merged reads that didn't align to the human genome (hg19), with the following database: /projects1/malt/databases/indexed/index040/full-nt_2017-10. 
```{r load_data}
## load the species and genus tables generated in MEGAN
# malt_species <- fread("05-results.backup/CMC_MALT_nt_species_bacteria_archaea.txt")
# malt_genus <- fread("05-results.backup/CMC_MALT_nt_genus_bacteria_archaea.txt")

# malt_species <- fread("05-results.backup/CMC_MALT_RSC_species_bacteria_archaea.txt")
# malt_genus <- fread("05-results.backup/CMC_MALT_RSC_genus_bacteria_archaea.txt")

# malt_species <- fread("05-results.backup/CMC_someHMP_MALT_RSC_species_bacteria_archaea.txt")
# malt_genus <- fread("05-results.backup/CMC_someHMP_MALT_RSC_genus_bacteria_archaea.txt")

# malt_species <- fread("./05-results.backup/CMC_HMP_MALT_RSC_species_bacteria_archaea.txt")
# malt_genus <- fread("./05-results.backup/CMC_HMP_MALT_RSC_genus_bacteria_archaea.txt")

malt_species <- fread("./05-results.backup/CMC_HMP_JAE_MALT_RSC_species_bacteria_archaea.txt")
malt_genus <- fread("./05-results.backup/CMC_HMP_JAE_MALT_RSC_genus_bacteria_archaea.txt")

# clean the file names
malt_species <- rename(malt_species, Species = `#Datasets`)
colnames(malt_species) <- gsub(".SG1.1_S0_L001_R1_001.fastq.combined.fq.prefixed.extractunmapped.bam","", colnames(malt_species))
colnames(malt_species) <- gsub("_S0_L001_R1_000.fastq.combined.fq.prefixed.extractunmapped.bam","", colnames(malt_species))
colnames(malt_species) <- gsub("_S0_L001_R1_001.fastq.combined.fq.prefixed.extractunmapped.bam.10M","_10M", colnames(malt_species))
colnames(malt_species) <- gsub("_S0_L001_R1_000.10M.fastq.combined.fq.prefixed.extractunmapped.bam","_10M", colnames(malt_species))
colnames(malt_species) <- gsub("_S0_L001_R1_001.fastq.combined.fq.prefixed.extractunmapped.bam","", colnames(malt_species))


# remove human and unassigned reads
malt_species <- malt_species %>% filter(!str_detect(Species, "Homo sapiens"))
malt_species <- malt_species %>% filter(!str_detect(Species, "Not assigned"))
# get rid of the extra column at the begining
malt_species <- malt_species %>% select(-CMC_HMP_comparison)

malt_genus <- rename(malt_genus, Species = `#Datasets`)
colnames(malt_genus) <- gsub(".SG1.1_S0_L001_R1_001.fastq.combined.fq.prefixed.extractunmapped.bam","", colnames(malt_genus))
colnames(malt_genus) <- gsub("_S0_L001_R1_000.fastq.combined.fq.prefixed.extractunmapped.bam","", colnames(malt_genus))
colnames(malt_genus) <- gsub("_S0_L001_R1_000.10M.fastq.combined.fq.prefixed.extractunmapped.bam","_10M", colnames(malt_genus))
colnames(malt_genus) <- gsub("_S0_L001_R1_001.fastq.combined.fq.prefixed.extractunmapped.bam.10M","_10M", colnames(malt_genus))
colnames(malt_genus) <- gsub("_S0_L001_R1_001.fastq.combined.fq.prefixed.extractunmapped.bam","", colnames(malt_genus))

# remove human and unassigned reads
malt_genus <- malt_genus %>% filter(!str_detect(Species, "Homo sapiens"))
malt_genus <- malt_genus %>% filter(!str_detect(Species, "Not assigned"))
# get rid of the extra column at the begining
malt_genus <- malt_genus %>% select(-CMC_HMP_comparison)

# load the metadata file
metadata <- fread("00-documentation.backup/01-cameroon_hunter_gatherer_metadata.tsv")
metadata <- metadata %>%
  rename(SampleID = `#SampleID`)  %>%
  as_tibble() %>%
  mutate(Description = fct_relevel(Description, "HunterGatherer", "HMP_supPlaque", "HMP_subPlaque",
                                   "HMP10M_supPlaque", "HMP10M_subPlaque","JAE","JAE10M",
                                   "ExtractionGauze","ExtractionBlank","LibraryBlank"),
         Ethnic_Group = fct_relevel(Ethnic_Group, "Baka","Nzime","HMP","HMP10M","JAE","JAE10M",
                                     "ExtractionGauze","ExtractionBlank","LibraryBlank"),
         Env = fct_relevel(Env, "HunterGatherer","HMP","HMP10M","JAE","JAE10M",
                                     "ExtractionGauze","ExtractionBlank","LibraryBlank"),
         Market_economy = fct_relevel(Market_economy, "logging_road","forrest_camp","farming","Industrial",
                                      "ExtractionGauze","ExtractionBlank","LibraryBlank"),
         Age_group = fct_relevel(Age_group, "18-29","30-39","40-49","50-59","60+","HMP","HMP10M"),
         Sex = fct_relevel(Sex, "Male","Female","HMP","HMP10M"),
         Tooth_site = fct_relevel(Tooth_site, "Anterior","Posterior","HMP","HMP10M"),
         Type = fct_relevel(Type, "Calculus","Plaque","Blank"),
         Village = fct_relevel(Village, "202","204","205","203","206","HMP","HMP10M","JAE","JAE10M",
                               "ExtractionGauze","ExtractionBlank","LibraryBlank"))

fullSRR <- metadata %>%
  select(SampleID) %>%
  filter(str_detect(SampleID, "SRR")) %>%
  filter(!str_detect(SampleID, "10M")) %>%
  pull
```

```{r colors_shapes_passionfruit, eval = F}
Village_colors <- c(`202` = "#C70E7B", `203` = "#E68561", `204` = "#60CB57", `205` = "#577EA1", `206` = "#172869",
                     ExtractionGauze = "#252525", ExtractionBlank = "#737373", LibraryBlank = "#bdbdbd")
village_shapes <- c(`202` = 15, `203` = 16, `204` = 17,`205` = 18, `206` = 25,
                    ExtractionGauze = 8, ExtractionBlank = 4, LibraryBlank = 3)

Ethnic_Group_colors <- c(Baka = "#C70E7B", Nzime = "#172869",
                         ExtractionGauze = "#252525", ExtractionBlank = "#737373", LibraryBlank = "#bdbdbd")
Ethnic_Group_shapes <- c(Baka = 19, Nzime = 17, 
                         ExtractionGauze = 8, ExtractionBlank = 4, LibraryBlank = 3)

Env_colors <- c(HunterGatherer = "#C70E7B", ExtractionGauze = "#881C00",
                ExtractionBlank = "#A34D1A", LibraryBlank = "#CC986A")
Env_shapes <- c(HunterGatherer = 19, ExtractionGauze = 8, ExtractionBlank = 4, LibraryBlank = 3)

market_economy_colors <- c(logging_road = "#C70E7B", forrest_camp = "#172869", farming = "#60CB57",
                           ExtractionGauze = "#252525", ExtractionBlank = "#737373", LibraryBlank = "#bdbdbd")
market_economy_shapes <- c(logging_road = 19, forrest_camp = 17, other = 18,
                           ExtractionGauze = 8, ExtractionBlank = 4, LibraryBlank = 3)

age_group_colors <- c(`18-29` = "#C70E7B", `30-39` = "#E68561", `40-49` = "#60CB57", `50-59` = "#577EA1", `60+` = "#172869",
                      ExtractionGauze = "#252525", ExtractionBlank = "#737373", LibraryBlank = "#bdbdbd")
age_group_shapes <- c(`18-29` = 15, `30-39` = 16, `40-49` = 17, `50-59` = 18, `60+` = 25,
                      ExtractionGauze = 8, ExtractionBlank = 4, LibraryBlank = 3)

sex_colors <- c(Female = "#C70E7B", Male = "#172869", ExtractionGauze = "#252525", ExtractionBlank = "#737373", LibraryBlank = "#bdbdbd")
sex_shapes <- c(Female = 19, Male = 17, ExtractionGauze = 8, ExtractionBlank = 4, LibraryBlank = 3)

tooth_site_colors <- c(Anterior = "#C70E7B", Posterior = "#172869", 
                       ExtractionGauze = "#252525", ExtractionBlank = "#737373", LibraryBlank = "#bdbdbd")
tooth_site_shapes <- c(Anterior = 19, Posterior = 17, 
                       ExtractionGauze = 8, ExtractionBlank = 4, LibraryBlank = 3)

```

```{r colors_shapes_peachpear}
# lacroix_palette("PeachPear", n = 8, type = "continuous")
# "#FF3200" "#EF8158" "#E9BD8E" "#CBDDA7" "#38BCAD" "#0B91B5" "#065FA3" "#172869"

Village_colors <- c(`202` = "#FF3200", `204` = "#EF8158", `205` = "#E9BD8E", `203` = "#CBDDA7", `206` = "#38BCAD",
                    HMP = "#172869", HMP10M = "#065FA3", JAE10M = "#ffeda0", JAE = "#feb24c",
                     ExtractionGauze = "#252525", ExtractionBlank = "#737373", LibraryBlank = "#bdbdbd")
Village_shapes <- c(`202` = 15, `203` = 16, `204` = 17,`205` = 18, `206` = 25, HMP = 7, HMP10M = 9, JAE = 7, JAE10M = 9,
                    ExtractionGauze = 8, ExtractionBlank = 4, LibraryBlank = 3)

Ethnic_Group_colors <- c(Baka = "#FF3200", Nzime = "#172869", HMP = "#969696", HMP10M = "#d9d9d9", JAE10M = "#ffeda0", JAE = "#feb24c",
                         ExtractionGauze = "#252525", ExtractionBlank = "#737373", LibraryBlank = "#bdbdbd")
Ethnic_Group_shapes <- c(Baka = 19, Nzime = 17,  HMP = 7, HMP10M = 9, JAE = 7, JAE10M = 9,
                         ExtractionGauze = 8, ExtractionBlank = 4, LibraryBlank = 3)

Env_colors <- c(HunterGatherer = "#FF3200", HMP = "#172869", HMP10M = "#065FA3", JAE10M = "#ffeda0", JAE = "#feb24c",
                ExtractionGauze = "#252525", ExtractionBlank = "#737373", LibraryBlank = "#bdbdbd")
Env_shapes <- c(HunterGatherer = 19, HMP = 7, HMP10M = 9, JAE = 7, JAE10M = 9, ExtractionGauze = 8, ExtractionBlank = 4, LibraryBlank = 3)

Market_economy_colors <- c(logging_road = "#FF3200", forrest_camp = "#172869", farming = "#82CDAA", Industrial = "#969696",
                           ExtractionGauze = "#252525", ExtractionBlank = "#737373", LibraryBlank = "#bdbdbd")
Market_economy_shapes <- c(logging_road = 19, forrest_camp = 17, farming = 18, HMP = 7, HMP10M = 9, Industrial = 15,
                           ExtractionGauze = 8, ExtractionBlank = 4, LibraryBlank = 3)

Age_group_colors <- c(`18-29` = "#FF3200", `30-39` = "#E9B186", `40-49` = "#82CDAA", `50-59` = "#0686B8", `60+` = "#172869",
                      HMP = "#969696", HMP10M = "#d9d9d9",
                      ExtractionGauze = "#252525", ExtractionBlank = "#737373", LibraryBlank = "#bdbdbd")
Age_group_shapes <- c(`18-29` = 15, `30-39` = 16, `40-49` = 17, `50-59` = 18, `60+` = 25, HMP = 7, HMP10M = 9,
                      ExtractionGauze = 8, ExtractionBlank = 4, LibraryBlank = 3)

Sex_colors <- c(Female = "#FF3200", Male = "#172869", HMP = "#969696", HMP10M = "#d9d9d9",
                ExtractionGauze = "#252525", ExtractionBlank = "#737373", LibraryBlank = "#bdbdbd")
Sex_shapes <- c(Female = 19, Male = 17,  HMP = 7, HMP10M = 9,ExtractionGauze = 8, ExtractionBlank = 4, LibraryBlank = 3)

Tooth_site_colors <- c(Anterior = "#FF3200", Posterior = "#172869", HMP = "#969696", HMP10M = "#d9d9d9",
                       ExtractionGauze = "#252525", ExtractionBlank = "#737373", LibraryBlank = "#bdbdbd")
Tooth_site_shapes <- c(Anterior = 19, Posterior = 17,  HMP = 7, HMP10M = 9,
                       ExtractionGauze = 8, ExtractionBlank = 4, LibraryBlank = 3)

```

```{r pca_plotting_functions}
# plotting PCA with colored dots
plot_pca <- function(df, pc1, pc2, metadata_group) {
    metadata_group_colors <- get(paste(metadata_group, "_colors", sep = ""))
    metadata_group_shapes <- get(paste(metadata_group, "_shapes", sep = ""))

    exp_var <- paste0(round(df$explained_variance * 100, 2), "%")
    df_X <- df$variates$X %>%
              as.data.frame() %>%
              rownames_to_column("SampleID") %>%
              inner_join(metadata %>%
                           select(SampleID, Env, Description, Ethnic_Group, Age_group, Sex, Market_economy, Village, Tooth_site, Type), by = "SampleID")
    
    metadata_group = df_X[[metadata_group]]
    
                       if (pc1 == 'PC1') {
                           pc1 <- df_X$PC1
                           exp_var_pc1 <- exp_var[1]
                       }  else if (pc1 == 'PC2') {
                           pc1 <- df_X$PC2
                           exp_var_pc1 <- exp_var[2]
                       } else if (pc1 == 'PC3') {
                           pc1 <- df_X$PC3
                           exp_var_pc1 <- exp_var[3]
                       }

                       if (pc2 == 'PC1') {
                           pc2 <- df_X$PC1
                           exp_var_pc2 <- exp_var[1]
                       }  else if (pc2 == 'PC2') {
                           pc2 <- df_X$PC2
                           exp_var_pc2 <- exp_var[2]
                       } else if (pc2 == 'PC3') {
                           pc2 <- df_X$PC3
                           exp_var_pc2 <- exp_var[3]
                       }

    pca_plot <- ggplot(df_X, aes(pc1, pc2, colour = metadata_group, shape = metadata_group)) +
                        geom_point(size = 4) +
                        scale_colour_manual(values = metadata_group_colors) +
                        scale_shape_manual(values = metadata_group_shapes) +
                        # stat_ellipse() +
                        xlab(paste(colnames(pc1), " - ", exp_var_pc1)) +
                        ylab(paste(colnames(pc2), " - ", exp_var_pc2)) +
                        theme_minimal(base_size = 10) +
                        theme(legend.position = "top") +
                        theme(legend.title = element_blank())

    return(pca_plot)
}

# this function plots the PCA with the sample names instead of points
plot_pca_names <- function(df, pc1, pc2, metadata_group) {
    metadata_group_colors <- get(paste(metadata_group, "_colors", sep = ""))

    exp_var <- paste0(round(df$explained_variance * 100, 2), "%")
    df_X <- df$variates$X %>%
              as.data.frame() %>%
              rownames_to_column("SampleID") %>%
              inner_join(metadata %>%
                           select(SampleID, Env, Description, Ethnic_Group, Age_group, Sex, Market_economy, Village, Tooth_site, Type), by = "SampleID")
    
    metadata_group = df_X[[metadata_group]]
    
                       if (pc1 == 'PC1') {
                           pc1 <- df_X$PC1
                           exp_var_pc1 <- exp_var[1]
                       }  else if (pc1 == 'PC2') {
                           pc1 <- df_X$PC2
                           exp_var_pc1 <- exp_var[2]
                       } else if (pc1 == 'PC3') {
                           pc1 <- df_X$PC3
                           exp_var_pc1 <- exp_var[3]
                       }

                       if (pc2 == 'PC1') {
                           pc2 <- df_X$PC1
                           exp_var_pc2 <- exp_var[1]
                       }  else if (pc2 == 'PC2') {
                           pc2 <- df_X$PC2
                           exp_var_pc2 <- exp_var[2]
                       } else if (pc2 == 'PC3') {
                           pc2 <- df_X$PC3
                           exp_var_pc2 <- exp_var[3]
                       }

    pca_plot_names <- ggplot(df_X, aes(pc1, pc2, colour = metadata_group)) +
                        geom_text(aes(label = df_X$SampleID), size = 2) +
                        scale_colour_manual(values = metadata_group_colors) +
                        # stat_ellipse() +
                        xlab(paste(colnames(pc1), " - ", exp_var_pc1)) +
                        ylab(paste(colnames(pc2), " - ", exp_var_pc2)) +
                        theme_minimal(base_size = 10) +
                        theme(legend.position = "top") +
                        theme(legend.title = element_blank())

    return(pca_plot_names)
}


# plot PCA with colored dots and the title including the # of species or genera
plot_pca_titles <- function(df, pc1, pc2, color_group, shape_group, ncomps) {
    metadata_group_colors <- get(paste(color_group, "_colors", sep = ""))
    metadata_group_shapes <- get(paste(shape_group, "_shapes", sep = ""))

    pca.list <- mixOmics::pca(df, ncomp = ncomps, logratio = 'CLR')

    exp_var <- paste0(round(pca.list$explained_variance * 100, 2), "%")
    df_X <- pca.list$variates$X %>%
              as.data.frame() %>%
              rownames_to_column("SampleID") %>%
              inner_join(metadata %>%
                           select(SampleID, Env, Description, Ethnic_Group, Age_group, Sex, Market_economy, Village, Tooth_site, Type), by = "SampleID")
    
    color_group = df_X[[color_group]]
    shape_group = df_X[[shape_group]]
   
                       if (pc1 == 'PC1') {
                           pc1 <- df_X$PC1
                           exp_var_pc1 <- exp_var[1]
                       }  else if (pc1 == 'PC2') {
                           pc1 <- df_X$PC2
                           exp_var_pc1 <- exp_var[2]
                       } else if (pc1 == 'PC3') {
                           pc1 <- df_X$PC3
                           exp_var_pc1 <- exp_var[3]
                       }

                       if (pc2 == 'PC1') {
                           pc2 <- df_X$PC1
                           exp_var_pc2 <- exp_var[1]
                       }  else if (pc2 == 'PC2') {
                           pc2 <- df_X$PC2
                           exp_var_pc2 <- exp_var[2]
                       } else if (pc2 == 'PC3') {
                           pc2 <- df_X$PC3
                           exp_var_pc2 <- exp_var[3]
                       }

    title_text <- df %>%
                     rownames_to_column("SampleID") %>%
                     gather("Species","Counts",2:ncol(.)) %>%
                     spread(SampleID, Counts) %>%
                     count
    
    pca_plot <- ggplot(df_X, aes(pc1, pc2, colour = color_group, shape = shape_group)) +
                        geom_point(size = 4) +
                        scale_colour_manual(values = metadata_group_colors) +
                        scale_shape_manual(values = metadata_group_shapes) +
                        # stat_ellipse() +
                        xlab(paste(colnames(pc1), " - ", exp_var_pc1)) +
                        ylab(paste(colnames(pc2), " - ", exp_var_pc2)) +
                        theme_minimal(base_size = 10) +
                        theme(legend.position = "top") +
                        ggtitle(paste(title_text))

    return(pca_plot)
}
# get(paste(metadata_group, "_colors", sep = ""))
# ggtitle(paste('MALT - unfiltered', malt_species_unfiltered %>%
#                                      rownames_to_column("SampleID") %>%
#                                      gather("Species","Counts",2:ncol(.)) %>%
#                                      spread(SampleID, Counts) %>%
#                                      count, 'species'))
  

```

```{r box_plotting_functions}

plot_box_whisker <- function(df, stat_info, metadata_group, fill_group, title_text) {
    fill_group_colors <- get(paste(fill_group, "_colors", sep = ""))

      df_X <- df %>%
              inner_join(metadata %>%
                           select(SampleID, Env, Description, Ethnic_Group, Age_group,
                                  Sex, Market_economy, Village, Tooth_site, Type), by = "SampleID")

    stat_info_table = df_X[[stat_info]]
    metadata_group = df_X[[metadata_group]]
    fill_group = df_X[[fill_group]]

    bx_wh_plot <- ggplot(df_X, aes(x = metadata_group, y = stat_info_table, fill = fill_group)) +
                         geom_boxplot() +
                         scale_fill_manual(values = fill_group_colors) +
                         theme_classic() +
                         #scale_y_continuous(trans='log10') +
                         #scale_y_continuous(trans='pseudo_log') +
                         theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
                         ylab(stat_info) +
                         ggtitle(title_text)

    return(bx_wh_plot)
}

```


Before running decontam, let's plot the samples in a PCA to see if there are any that are clearly not good-quality
```{r raw_plot}
# malt_species_raw <- malt_species %>%
#                                          adorn_totals(where = "col", name = "Sum_species") %>%
#                                          mutate(Percent = Sum_species/sum(Sum_species)*100) %>%
#                                          # filter(Percent > 0.01) %>%
#                                          select(-one_of(c("Sum_species","Percent"))) %>%
#                                          gather("SampleID","Counts",2:ncol(.)) %>%
#                                          spread(Species,Counts)


malt_species_raw <- malt_species %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts)

rownames(malt_species_raw) <- malt_species_raw$SampleID

malt_species_raw <- malt_species_raw %>% select(-SampleID)

malt_species_raw = malt_species_raw+1

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.malt_species_pca <- tune.pca(malt_species_raw, logratio = 'CLR')
tune.malt_species_pca

# perform a PCA to see how the data cluster
malt_species.pca <- mixOmics::pca(malt_species_raw, ncomp = 3, logratio = 'CLR')
plot(malt_species.pca)


plot_pca(malt_species.pca, "PC1", "PC2", "Env")
plot_pca_names(malt_species.pca,  "PC1", "PC2", "Env")

```

List the outliers in the plot above and remove them before running decontam to ensure they're not influencing the contaminant assessment. 
```{r outliers}
outliers <- c("CMC002.B0201","CMC032.B0101","CMC037.B0201","CMC047.A0201","EXB059.A1901","EXB059.A1301","EXB059.A1101","EXB059.A0901")

```

# Decontamination with Decontam
We want to know if there are any species that can be removed b/c they appear to be contaminants. Use the package decontam to determine which species may be contaminants. Prepare the matrix for decontam. 
```{r decontam_data_prep}
# list out the samples that have all 0 or NA values to remove from the input matrix
NA_filter_list <- metadata %>% filter((is.na(SumLibraryConcentration_copies_ul) | SumLibraryConcentration_copies_ul <= 0)) %>% select(SampleID)

# remove NA value samples from the input data frame
malt_species.dataframe <- malt_species %>%
  filter(!str_detect(Species, "Not assigned")) %>%
  select(-one_of(outliers)) %>%
  gather("SampleID","Counts", 2:ncol(.)) %>%
  spread(Species,Counts) %>%
  anti_join(., NA_filter_list)

# make a matrix for decontam
malt_species.matrix <- data.matrix(malt_species.dataframe[2:ncol(malt_species.dataframe)])
rownames(malt_species.matrix) <- malt_species.dataframe$SampleID

# remove NA value samples from the input data frame
malt_genus.dataframe <- malt_genus %>%
  filter(!str_detect(Species, "Not assigned")) %>%
  select(-one_of(outliers)) %>%
  gather("SampleID","Counts", 2:ncol(.)) %>%
  spread(Species,Counts) %>%
  anti_join(., NA_filter_list)

# make a matrix for decontam
malt_genus.matrix <- data.matrix(malt_genus.dataframe[2:ncol(malt_genus.dataframe)])
rownames(malt_genus.matrix) <- malt_genus.dataframe$SampleID


# make a metadata dataframe with the same NA-containing samples removed
metadata.decontam <- metadata %>%
  anti_join(., NA_filter_list, by = "SampleID") %>%
  filter(!str_detect(SampleID, outliers %>% str_c(collapse = "|")))

```

Make a species abundance curve for the samples based on Env to get an idea of the number of species identified in each prior to filtering, and how long the tail is. 
```{r rank_abundance_curves}

malt_species.dataframeL <- malt_species.dataframe %>%
  adorn_percentages(denominator = "row", na.rm=F) %>%
  gather("Species","Relab",2:ncol(.)) %>%
  inner_join(., metadata)

Envmt <- unique(malt_species.dataframeL$Env)

species_orders <- lapply(Envmt, function(envmt) {
malt_species.dataframeL_check <- malt_species.dataframeL %>%
  filter(Env == envmt) %>%
  group_by(Species) %>%
  summarise(mean_relab = mean(Relab)) %>%
  arrange(desc(mean_relab)) %>%
  filter(mean_relab > 0) %>%
  rownames_to_column("order") %>%
  mutate(order = as.numeric(order)) %>%
  mutate(Env = envmt)
}) %>%
  bind_rows(.)
  
species_abundance_curves <- function(df, envt, title_text) {
      options(scipen = 10)
      
    species_abundance_plot <- species_orders %>%
        filter(Env == envt) %>%
           ggplot(., aes(x=order, y=mean_relab)) +
           geom_point() +
           theme_minimal() +
           scale_y_log10() +
           geom_hline(yintercept=0.01, linetype="dashed", color = "red") +
           xlim(NA, 600) +
           ylab("mean relative abundance") +
           xlab("species order") +
           ggtitle(title_text)
  
    return(species_abundance_plot)
}

sp_curve_HG <- species_abundance_curves(species_orders, "HunterGatherer", "Hunter Gatherer - species")
sp_curve_EG <- species_abundance_curves(species_orders, "ExtractionGauze", "Extraction Gauze - species")
sp_curve_EB <- species_abundance_curves(species_orders, "ExtractionBlank", "Extraction Blank - species")
sp_curve_LB <- species_abundance_curves(species_orders, "LibraryBlank", "Library Blank - species")

plot_grid(sp_curve_HG, sp_curve_EG,
          sp_curve_EB, sp_curve_LB,
          nrow=2, labels = c("A", "B", "C", "D"), rel_widths = c(1, 1))


# repeat for genus level
malt_genus.dataframeL <- malt_genus.dataframe %>%
  adorn_percentages(denominator = "row", na.rm=F) %>%
  gather("genus","Relab",2:ncol(.)) %>%
  inner_join(., metadata)

Envmt <- unique(malt_genus.dataframeL$Env)

genus_orders <- lapply(Envmt, function(envmt) {
malt_genus.dataframeL_check <- malt_genus.dataframeL %>%
  filter(Env == envmt) %>%
  group_by(genus) %>%
  summarise(mean_relab = mean(Relab)) %>%
  arrange(desc(mean_relab)) %>%
  filter(mean_relab > 0) %>%
  rownames_to_column("order") %>%
  mutate(order = as.numeric(order)) %>%
  mutate(Env = envmt)
}) %>%
  bind_rows(.)
  
genus_abundance_curves <- function(df, envt, title_text) {
      options(scipen = 10)
      
    genus_abundance_plot <- genus_orders %>%
        filter(Env == envt) %>%
           ggplot(., aes(x=order, y=mean_relab)) +
           geom_point() +
           theme_minimal() +
           scale_y_log10() +
           geom_hline(yintercept=0.01, linetype="dashed", color = "red") +
           xlim(NA, 250) +
           ylab("mean relative abundance") +
           xlab("genus order") +
           ggtitle(title_text)
  
    return(genus_abundance_plot)
}

gn_curve_HG <- genus_abundance_curves(genus_orders, "HunterGatherer", "Hunter Gatherer - genus")
gn_curve_EG <- genus_abundance_curves(genus_orders, "ExtractionGauze", "Extraction Gauze - genus")
gn_curve_EB <- genus_abundance_curves(genus_orders, "ExtractionBlank", "Extraction Blank - genus")
gn_curve_LB <- genus_abundance_curves(genus_orders, "LibraryBlank", "Library Blank - genus")

plot_grid(gn_curve_HG, gn_curve_EG,
          gn_curve_EB, gn_curve_LB,
          nrow=2, labels = c("A", "B", "C", "D"), rel_widths = c(1, 1))

```



Now we're ready to run Decontam. Start with the frequency method. 
```{r decontam_frequency, eval = F}
# Check for contamination by fequency
contam.freq.species <- isContaminant(malt_species.matrix, method="frequency", 
                                 conc=metadata.decontam$SumLibraryConcentration_copies_ul, 
                                 threshold = 0.9)

# Print the # of species identified as contaminants by frequency
table(contam.freq.species$contaminant)

cont_freq_species_list <- contam.freq.species[which(contam.freq.species$contaminant=="TRUE"), ]

# Make a histogram to look at the distribution of the p-value (probability for classifying contaminants)

ggplot(contam.freq.species, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Frequency - Species')

# Check for contamination by fequency
contam.freq.genus <- isContaminant(malt_genus.matrix, method="frequency", 
                                 conc=metadata.decontam$SumLibraryConcentration_copies_ul, 
                                 threshold = 0.9)

# Print the # of genus identified as contaminants by frequency
table(contam.freq.genus$contaminant)

cont_freq_genus_list <- contam.freq.genus[which(contam.freq.genus$contaminant=="TRUE"), ]

# Make a histogram to look at the distribution of the p-value (probability for classifying contaminants)

ggplot(contam.freq.genus, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Frequency - Genus')

```

Try the prevalence method to compare. 
```{r decontam_prevalence}
# Assign sample-type status to identify control samples
metadata.decontam$is.control <- metadata.decontam$Sample_or_Control=="Control"

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam.prev.species <- isContaminant(malt_species.matrix, method="prevalence", 
                                 neg=metadata.decontam$is.control, 
                                 threshold = 0.6)

# Check number of contaminants
table(contam.prev.species$contaminant)

# List of contaminants
cont_prev_species_list <- contam.prev.species[which(contam.prev.species$contaminant=="TRUE"), ]

# Look at the histogram - does the probability threshold need to be changed?
ggplot(contam.prev.species, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Prevalence - Species')

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam.prev.genus <- isContaminant(malt_genus.matrix, method="prevalence", 
                                 neg=metadata.decontam$is.control, 
                                 threshold = 0.6)

# Check number of contaminants
table(contam.prev.genus$contaminant)

# List of contaminants
cont_prev_genus_list <- contam.prev.genus[which(contam.prev.genus$contaminant=="TRUE"), ]

# Look at the histogram - does the probability threshold need to be changed?
ggplot(contam.prev.genus, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Prevalence -Genus')

```

Check the rank abundance curves to see how changing the cut-off threshold changes the tail.
```{r decontam_rank_abund_curve}

contaminants_species <- cont_prev_species_list %>% rownames_to_column("Species") %>% select(Species)
malt_species.decontam <- malt_species %>%
  anti_join(., contaminants_species)

# write.table(malt_species.decontam, file = "05-results.backup/malt_species.decontam.tsv", sep="\t", quote=F, row.names = F)
# write.table(malt_species, "05-results.backup/malt_species.tsv", sep="\t", quote=F, row.names = F)

malt_species.decontamL <- malt_species.decontam %>%
  adorn_percentages(denominator = "col", na.rm=F) %>%
  gather("SampleID","Relab",2:ncol(.)) %>%
  inner_join(., metadata %>%
              select(SampleID, Env), by = "SampleID")

Envmt_decontam <- unique(malt_species.decontamL$Env)

species_decontam_orders <- lapply(Envmt_decontam, function(envmt) {
malt_species.dataframeL_check <- malt_species.decontamL %>%
  filter(Env == envmt) %>%
  group_by(Species) %>%
  summarise(mean_relab = mean(Relab)) %>%
  arrange(desc(mean_relab)) %>%
  filter(mean_relab > 0) %>%
  rownames_to_column("order") %>%
  mutate(order = as.numeric(order)) %>%
  mutate(Env = envmt)
}) %>%
  bind_rows(.)
  
species_decontam_abundance_curves <- function(df, envt, title_text) {
      options(scipen = 10)
      
    species_abundance_plot <- species_decontam_orders %>%
        filter(Env == envt) %>%
           ggplot(., aes(x=order, y=mean_relab)) +
           geom_point() +
           theme_minimal() +
           scale_y_log10() +
           geom_hline(yintercept=0.01, linetype="dashed", color = "red") +
           xlim(NA, 500) +
           ylab("mean relative abundance") +
           xlab("species order") +
           ggtitle(title_text) + 
           theme(plot.title = element_text(size = 8))
  
    return(species_abundance_plot)
}

sp_decontam_curve_HG <- species_decontam_abundance_curves(species_decontam_orders, "HunterGatherer", "HG, decontam")
sp_decontam_curve_EG <- species_decontam_abundance_curves(species_decontam_orders, "ExtractionGauze", "EG, decontam")
sp_decontam_curve_EB <- species_decontam_abundance_curves(species_decontam_orders, "ExtractionBlank", "EB, decontam")
sp_decontam_curve_LB <- species_decontam_abundance_curves(species_decontam_orders, "LibraryBlank", "LB, decontam")
sp_decontam_curve_HMP <- species_decontam_abundance_curves(species_decontam_orders, "HMP", "HMP, decontam")
sp_decontam_curve_HMP10 <- species_decontam_abundance_curves(species_decontam_orders, "HMP10M", "HMP10Ms, decontam")
sp_decontam_curve_JAE <- species_decontam_abundance_curves(species_decontam_orders, "JAE", "JAE, decontam")
sp_decontam_curve_JAE10 <- species_decontam_abundance_curves(species_decontam_orders, "JAE10M", "JAE10M, decontam")

plot_grid(sp_decontam_curve_HG, sp_decontam_curve_EG, sp_decontam_curve_EB, sp_decontam_curve_LB,
          sp_decontam_curve_HMP, sp_decontam_curve_HMP10, sp_decontam_curve_JAE, sp_decontam_curve_JAE10,
          nrow=2, labels = c("A", "B", "C", "D", "E", "F", "G", "H"), rel_widths = c(1, 1))

# Now repeatw with genus
contaminants_genus <- cont_prev_genus_list %>% rownames_to_column("Species") %>% select(Species)
malt_genus.decontam <- malt_genus %>%
  anti_join(., contaminants_genus)

malt_genus.decontamL <- malt_genus.decontam %>%
  adorn_percentages(denominator = "col", na.rm=F) %>%
  gather("SampleID","Relab",2:ncol(.)) %>%
  inner_join(., metadata)


Envmt_decontam <- unique(malt_genus.decontamL$Env)

genus_decontam_orders <- lapply(Envmt_decontam, function(envmt) {
malt_genus.dataframeL_check <- malt_genus.decontamL %>%
  filter(Env == envmt) %>%
  group_by(Species) %>%
  summarise(mean_relab = mean(Relab)) %>%
  arrange(desc(mean_relab)) %>%
  filter(mean_relab > 0) %>%
  rownames_to_column("order") %>%
  mutate(order = as.numeric(order)) %>%
  mutate(Env = envmt)
}) %>%
  bind_rows(.)
  
genus_decontam_abundance_curves <- function(df, envt, title_text) {
      options(scipen = 10)
      
    genus_abundance_plot <- genus_decontam_orders %>%
        filter(Env == envt) %>%
           ggplot(., aes(x=order, y=mean_relab)) +
           geom_point() +
           theme_minimal() +
           scale_y_log10() +
           geom_hline(yintercept=0.01, linetype="dashed", color = "red") +
           xlim(NA, 150) +
           ylab("mean relative abundance") +
           xlab("genus order") +
           ggtitle(title_text) + 
           theme(plot.title = element_text(size = 8))
  
    return(genus_abundance_plot)
}

gn_decontam_curve_HG <- genus_decontam_abundance_curves(genus_decontam_orders, "HunterGatherer", "HG, decontam")
gn_decontam_curve_EG <- genus_decontam_abundance_curves(genus_decontam_orders, "ExtractionGauze", "EG, decontam")
gn_decontam_curve_EB <- genus_decontam_abundance_curves(genus_decontam_orders, "ExtractionBlank", "EB, decontam")
gn_decontam_curve_LB <- genus_decontam_abundance_curves(genus_decontam_orders, "LibraryBlank", "LB, decontam")
gn_decontam_curve_HMP <- genus_decontam_abundance_curves(genus_decontam_orders, "HMP", "HMP, decontam")
gn_decontam_curve_HMP10 <- genus_decontam_abundance_curves(genus_decontam_orders, "HMP10M", "HMP10M, decontam")
gn_decontam_curve_JAE <- genus_decontam_abundance_curves(genus_decontam_orders, "JAE", "JAE, decontam")
gn_decontam_curve_JAE10 <- genus_decontam_abundance_curves(genus_decontam_orders, "JAE10M", "JAE10M, decontam")

plot_grid(gn_decontam_curve_HG, gn_decontam_curve_EG, gn_decontam_curve_EB, gn_decontam_curve_LB,
          gn_decontam_curve_HMP, gn_decontam_curve_HMP10,gn_decontam_curve_JAE, gn_decontam_curve_JAE10,
          nrow=2, labels = c("A", "B", "C", "D", "E", "F","G","H"), rel_widths = c(1, 1))

```


Check the combined prevalence and frequency method to see how it differs from the individual methods. 
```{r decontam_combined, eval=F}
# Assign sample-type status to identify control samples (same as was done for prevalence)
metadata.decontam$is.control <- metadata.decontam$Sample_or_Control=="Control"

# Assign contaminant status
contam.comb.species <- isContaminant(malt_species.matrix, method="combined", 
                                 conc=metadata.decontam$SumLibraryConcentration_copies_ul, 
                                 neg=metadata.decontam$is.control,
                                 threshold = 0.9)

## Check how many contaminants
table(contam.comb.species$contaminant)

## List of contaminants
cont_comb_species_list <- contam.comb.species[which(contam.comb.species$contaminant=="TRUE"), ]

# Graph it to look at the histogram and pick a suitable probability-values cut-off
ggplot(contam.comb.species, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  geom_hline(yintercept=10, linetype="dashed", color = "red") +
  geom_hline(yintercept=2, linetype="dashed", color = "red") +
  ggtitle('Combined - Species')

# Assign contaminant status
contam.comb.genus <- isContaminant(malt_genus.matrix, method="combined", 
                                 conc=metadata.decontam$SumLibraryConcentration_copies_ul, 
                                 neg=metadata.decontam$is.control,
                                 threshold = 0.9)

## Check how many contaminants
table(contam.comb.genus$contaminant)

## List of contaminants
cont_comb_genus_list <- contam.comb.genus[which(contam.comb.genus$contaminant=="TRUE"), ]

# Graph it to look at the histogram and pick a suitable probability-values cut-off
ggplot(contam.comb.genus, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  geom_hline(yintercept=10, linetype="dashed", color = "red") +
  geom_hline(yintercept=2, linetype="dashed", color = "red") +
  ggtitle('Combined - Genus')

```

We'll use the prevalence method, because this doesn't include known oral taxa in the contaminant list (frequency method does), and it picks up a lot more than the combined method does (but is that important if we cut off the tail of low-abundance taxa (say <0.01% abundance) anyway?). But which cut-off to use? How does changing the cut-off affect the tail of low-abundance taxa?
List out the pathways that are considered contaminants, to be removed from the table prior to all analyses. Then remove them from the table to make the decontaminated input table for analysis.
```{r contaminants}
contaminants_species <- cont_prev_species_list %>% rownames_to_column("Species") %>% select(Species)
malt_species.decontam <- malt_species %>%
  # gather("SampleID","Counts", 2:ncol(.)) %>%
  # spread(Species, Counts) %>%
  anti_join(., contaminants_species)

# write.table(malt_species.decontam, file = "05-results.backup/malt_species.decontam.tsv", sep="\t", quote=F, row.names = F)
# write.table(malt_species, "05-results.backup/malt_species.tsv", sep="\t", quote=F, row.names = F)

contaminants_genus <- cont_prev_genus_list %>% rownames_to_column("Species") %>% select(Species)
malt_genus.decontam <- malt_genus %>%
  # gather("SampleID","Counts", 2:ncol(.)) %>%
  # spread(Species, Counts) %>%
  anti_join(., contaminants_genus)

```

Let's do a quick and not at all enough data filtering so we can run a quick PCA to see if there are trends
```{r pca_all_species}
# unfiltered and not decontaminated
malt_species_unfiltered <- malt_species %>%
                                         adorn_totals(where = "col", name = "Sum_species") %>%
                                         select(-one_of(outliers)) %>%
                                         select(-contains("10M")) %>%
                                         mutate(Percent = Sum_species/sum(Sum_species)*100) %>%
                                         select(-one_of(c("Sum_species","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts)

rownames(malt_species_unfiltered) <- malt_species_unfiltered$SampleID

malt_species_unfiltered <- malt_species_unfiltered %>% select(-SampleID)

malt_species_unfiltered = malt_species_unfiltered+1

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.species_pca <- tune.pca(malt_species_unfiltered, logratio = 'CLR')
tune.species_pca

# perform and plot a PCA to see how the data cluster
plot_pca_titles(malt_species_unfiltered, "PC1", "PC2", "Env", "Env", 3)
plot_pca_titles(malt_species_unfiltered, "PC3", "PC2", "Env", "Env", 3)


# ggsave("05-results.backup/pca_species_unfiltered_pc1pc2.pdf", plot = pca_species_unfiltered_pc1pc2, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"),
#        dpi = 300)


# filtered only
malt_species_filtered <- malt_species %>%
                                         adorn_totals(where = "col", name = "Sum_species") %>%
                                         select(-one_of(outliers)) %>%
                                         select(-contains("10M")) %>%
                                         mutate(Percent = Sum_species/sum(Sum_species)*100) %>%
                                         filter(Percent > 0.005) %>%
                                         select(-one_of(c("Sum_species","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts)

rownames(malt_species_filtered) <- malt_species_filtered$SampleID

malt_species_filtered <- malt_species_filtered %>% select(-SampleID)

malt_species_filtered = malt_species_filtered+1

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.species_pca <- tune.pca(malt_species_filtered, logratio = 'CLR')
tune.species_pca

# perform and plot a PCA to see how the data cluster
plot_pca_titles(malt_species_filtered, "PC1", "PC2", "Env", "Env", 3)
plot_pca_titles(malt_species_filtered, "PC3", "PC2", "Env", "Env", 3)


# decontaminated only
malt_species.decontam_unfilt <- malt_species.decontam %>%
                                         adorn_totals(where = "col", name = "Sum_species.decontam") %>%
                                         select(-one_of(outliers)) %>%
                                         select(-contains("10M")) %>%
                                         mutate(Percent = Sum_species.decontam/sum(Sum_species.decontam)*100) %>%
                                         select(-one_of(c("Sum_species.decontam","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts)

rownames(malt_species.decontam_unfilt) <- malt_species.decontam_unfilt$SampleID

malt_species.decontam_unfilt <- malt_species.decontam_unfilt %>% select(-SampleID)

malt_species.decontam_unfilt = malt_species.decontam_unfilt+1

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.species.decontam_pca <- tune.pca(malt_species.decontam_unfilt, logratio = 'CLR')
tune.species.decontam_pca

# perform and plot a PCA to see how the data cluster
plot_pca_titles(malt_species.decontam_unfilt, "PC1", "PC2", "Env", "Env", 3)
plot_pca_titles(malt_species.decontam_unfilt, "PC3", "PC2", "Env", "Env", 3)

# decontaminated and filtered
malt_species.decontam_filtered <- malt_species.decontam %>%
                                         adorn_totals(where = "col", name = "Sum_species.decontam_filtered") %>%
                                         select(-one_of(outliers)) %>%
                                         select(-contains("10M")) %>%
                                         mutate(Percent = Sum_species.decontam_filtered/sum(Sum_species.decontam_filtered)*100) %>%
                                         filter(Percent > 0.005) %>%
                                         select(-one_of(c("Sum_species.decontam_filtered","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts)

rownames(malt_species.decontam_filtered) <- malt_species.decontam_filtered$SampleID

malt_species.decontam_filtered <- malt_species.decontam_filtered %>% select(-SampleID)

malt_species.decontam_filtered = malt_species.decontam_filtered+1

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.species.decontam_filtered_pca <- tune.pca(malt_species.decontam_filtered, logratio = 'CLR')
tune.species.decontam_filtered_pca

# perform and plot a PCA to see how the data cluster
plot_pca_titles(malt_species.decontam_filtered, "PC1", "PC2", "Env", "Env", 3)
plot_pca_titles(malt_species.decontam_filtered, "PC3", "PC2", "Env", "Env", 3)

```

```{r pca_all_genus}
# unfiltered and not decontaminated
malt_genus_unfiltered <- malt_genus %>%
                                         adorn_totals(where = "col", name = "Sum_genus") %>%
                                         select(-one_of(outliers)) %>%
                                         select(-contains("10M")) %>%
                                         mutate(Percent = Sum_genus/sum(Sum_genus)*100) %>%
                                         select(-one_of(c("Sum_genus","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts)

rownames(malt_genus_unfiltered) <- malt_genus_unfiltered$SampleID

malt_genus_unfiltered <- malt_genus_unfiltered %>% select(-SampleID)

malt_genus_unfiltered = malt_genus_unfiltered+1

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.nt_genus_pca <- tune.pca(malt_genus_unfiltered, logratio = 'CLR')
tune.nt_genus_pca

# perform a PCA to see how the data cluster
nt_genus.pca <- mixOmics::pca(malt_genus_unfiltered, ncomp = 3, logratio = 'CLR')

# perform and plot a PCA to see how the data cluster
plot_pca_titles(malt_genus_unfiltered, "PC1", "PC2", "Env", "Env", 3)
plot_pca_titles(malt_genus_unfiltered, "PC3", "PC2", "Env", "Env", 3)


# filtered only
malt_genus_filtered <- malt_genus %>%
                                         adorn_totals(where = "col", name = "Sum_genus") %>%
                                         select(-one_of(outliers)) %>%
                                         select(-contains("10M")) %>%
                                         mutate(Percent = Sum_genus/sum(Sum_genus)*100) %>%
                                         filter(Percent > 0.01) %>%
                                         select(-one_of(c("Sum_genus","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts)

rownames(malt_genus_filtered) <- malt_genus_filtered$SampleID

malt_genus_filtered <- malt_genus_filtered %>% select(-SampleID)

malt_genus_filtered = malt_genus_filtered+1

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.nt_genus_pca <- tune.pca(malt_genus_filtered, logratio = 'CLR')
tune.nt_genus_pca

# perform and plot a PCA to see how the data cluster
plot_pca_titles(malt_genus_filtered, "PC1", "PC2", "Env", "Env", 3)
plot_pca_titles(malt_genus_filtered, "PC3", "PC2", "Env", "Env", 3)


# ggsave("05-results.backup/pca_nt_genus_pc1pc2.pdf", plot = pca_nt_genus_pc1pc2, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"),
#        dpi = 300)


# decontaminated only
malt_genus.decontam_unfiltered <- malt_genus.decontam %>%
                                         adorn_totals(where = "col", name = "Sum_genus.decontam") %>%
                                         select(-one_of(outliers)) %>%
                                         select(-contains("10M")) %>%
                                         mutate(Percent = Sum_genus.decontam/sum(Sum_genus.decontam)*100) %>%
                                         select(-one_of(c("Sum_genus.decontam","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts)

rownames(malt_genus.decontam_unfiltered) <- malt_genus.decontam_unfiltered$SampleID

malt_genus.decontam_unfiltered <- malt_genus.decontam_unfiltered %>% select(-SampleID)

malt_genus.decontam_unfiltered = malt_genus.decontam_unfiltered+1

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.nt_genus.decontam_pca <- tune.pca(malt_genus.decontam_unfiltered, logratio = 'CLR')
tune.nt_genus.decontam_pca

# perform and plot a PCA to see how the data cluster
plot_pca_titles(malt_genus.decontam_unfiltered, "PC1", "PC2", "Env", "Env", 3)
plot_pca_titles(malt_genus.decontam_unfiltered, "PC3", "PC2", "Env", "Env", 3)


# decontaminated and filtered
malt_genus.decontam_filtered <- malt_genus.decontam %>%
                                         adorn_totals(where = "col", name = "Sum_genus.decontam") %>%
                                         select(-one_of(outliers)) %>%
                                         select(-contains("10M")) %>%
                                         mutate(Percent = Sum_genus.decontam/sum(Sum_genus.decontam)*100) %>%
                                         filter(Percent > 0.01) %>%
                                         select(-one_of(c("Sum_genus.decontam","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts)

rownames(malt_genus.decontam_filtered) <- malt_genus.decontam_filtered$SampleID

malt_genus.decontam_filtered <- malt_genus.decontam_filtered %>% select(-SampleID)

malt_genus.decontam_filtered = malt_genus.decontam_filtered+1

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.nt_genus.decontam_pca <- tune.pca(malt_genus.decontam_filtered, logratio = 'CLR')
tune.nt_genus.decontam_pca

# perform and plot a PCA to see how the data cluster
plot_pca_titles(malt_genus.decontam_filtered, "PC1", "PC2", "Env", "Env", 3)
plot_pca_titles(malt_genus.decontam_filtered, "PC3", "PC2", "Env", "Env", 3)

```

```{r pca_samples_species}
malt_species.decontam_filtered_noblanks <- malt_species.decontam %>%
                                         select(-starts_with("EXB")) %>%
                                         select(-starts_with("LIB")) %>%
                                         select(-contains("10M")) %>%
                                         # select(-one_of(fullSRR)) %>%
                                         adorn_totals(where = "col", name = "Sum_species") %>%
                                         mutate(Percent = Sum_species/sum(Sum_species)*100) %>%
                                         filter(Percent > 0.001) %>%
                                         select(-one_of(c("Sum_species","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts)

rownames(malt_species.decontam_filtered_noblanks) <- malt_species.decontam_filtered_noblanks$SampleID

malt_species.decontam_filtered_noblanks <- malt_species.decontam_filtered_noblanks %>% select(-SampleID)

malt_species.decontam_filtered_noblanks = malt_species.decontam_filtered_noblanks+1

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.nt_species_pca <- tune.pca(malt_species.decontam_filtered_noblanks, logratio = 'CLR')
tune.nt_species_pca

# perform and plot a PCA to see how the data cluster
# by Village
plot_pca_titles(malt_species.decontam_filtered_noblanks, "PC1", "PC2", "Ethnic_Group", "Village", 3)
plot_pca_titles(malt_species.decontam_filtered_noblanks, "PC3", "PC2", "Ethnic_Group", "Village", 3)

# by Sex
plot_pca_titles(malt_species.decontam_filtered_noblanks, "PC1", "PC2", "Ethnic_Group", "Sex", 3)
plot_pca_titles(malt_species.decontam_filtered_noblanks, "PC3", "PC2", "Ethnic_Group", "Sex", 3)

# by Age group
plot_pca_titles(malt_species.decontam_filtered_noblanks, "PC1", "PC2", "Ethnic_Group", "Age_group", 3)
plot_pca_titles(malt_species.decontam_filtered_noblanks, "PC3", "PC2", "Ethnic_Group", "Age_group", 3)

# by Market economy
plot_pca_titles(malt_species.decontam_filtered_noblanks, "PC1", "PC2", "Ethnic_Group", "Market_economy", 3)
plot_pca_titles(malt_species.decontam_filtered_noblanks, "PC3", "PC2", "Ethnic_Group", "Market_economy", 3)

# by Tooth site
plot_pca_titles(malt_species.decontam_filtered_noblanks, "PC1", "PC2", "Ethnic_Group", "Tooth_site", 3)
plot_pca_titles(malt_species.decontam_filtered_noblanks, "PC3", "PC2", "Ethnic_Group", "Tooth_site", 3)


# by Individual sequencing depth
# perform a PCA to see how the data cluster
malt_species.pca <- mixOmics::pca(malt_species.decontam_filtered_noblanks, ncomp = 3, logratio = 'CLR')
plot(malt_species.pca)

# Select out the PC variates into a new table and add the metadata for plotting
malt_species.pca.variates <- as.data.frame(malt_species.pca$variates$X)
malt_species.pca.variates <- malt_species.pca.variates %>%
  rownames_to_column("SampleID")
  
malt_species.pca.varmet <- inner_join(malt_species.pca.variates, metadata, by = "SampleID")

pca_malt_species_pc1pc2 <- ggplot(malt_species.pca.varmet, aes(PC1, PC2, colour = Individual_Seq_Depth_NonHuman, shape = Ethnic_Group)) +
  geom_point(size = 4) +
  # scale_colour_manual(values = common_colors) +
  # scale_shape_manual(values = village_shapes) +
  xlab(paste("PC1 - ", round(malt_species.pca$explained_variance[1], 2), sep = "")) +
  ylab(paste("PC2 - ", round(malt_species.pca$explained_variance[2], 2), sep = "")) +
  theme_minimal(base_size = 14) 

pca_malt_species_pc1pc2

# ggsave("05-results.backup/pca_nt_species_pc1pc2.pdf", plot = pca_nt_species_pc1pc2, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"),
#        dpi = 300)

```

```{r pca_samples_genus}
malt_genus.decontam_filtered_noblanks <- malt_genus.decontam %>%
                                         select(-starts_with("EXB")) %>%
                                         select(-starts_with("LIB")) %>%
                                         select(-contains("10M")) %>%
                                         # select(-one_of(fullSRR)) %>%
                                         adorn_totals(where = "col", name = "Sum_genus") %>%
                                         mutate(Percent = Sum_genus/sum(Sum_genus)*100) %>%
                                         filter(Percent > 0.01) %>%
                                         select(-one_of(c("Sum_genus","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts)

rownames(malt_genus.decontam_filtered_noblanks) <- malt_genus.decontam_filtered_noblanks$SampleID

malt_genus.decontam_filtered_noblanks <- malt_genus.decontam_filtered_noblanks %>% select(-SampleID)

malt_genus.decontam_filtered_noblanks = malt_genus.decontam_filtered_noblanks+1

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.nt_genus_pca <- tune.pca(malt_genus.decontam_filtered_noblanks, logratio = 'CLR')
tune.nt_genus_pca

# perform and plot a PCA to see how the data cluster
# by Village
plot_pca_titles(malt_genus.decontam_filtered_noblanks, "PC1", "PC2", "Ethnic_Group", "Village", 3)
plot_pca_titles(malt_genus.decontam_filtered_noblanks, "PC3", "PC2", "Ethnic_Group", "Village", 3)

# by Sex
plot_pca_titles(malt_genus.decontam_filtered_noblanks, "PC1", "PC2", "Ethnic_Group", "Sex", 3)
plot_pca_titles(malt_genus.decontam_filtered_noblanks, "PC3", "PC2", "Ethnic_Group", "Sex", 3)

# by Age group
plot_pca_titles(malt_genus.decontam_filtered_noblanks, "PC1", "PC2", "Ethnic_Group", "Age_group", 3)
plot_pca_titles(malt_genus.decontam_filtered_noblanks, "PC3", "PC2", "Ethnic_Group", "Age_group", 3)

# by Market economy
plot_pca_titles(malt_genus.decontam_filtered_noblanks, "PC1", "PC2", "Ethnic_Group", "Market_economy", 3)
plot_pca_titles(malt_genus.decontam_filtered_noblanks, "PC3", "PC2", "Ethnic_Group", "Market_economy", 3)

# by Tooth site
plot_pca_titles(malt_genus.decontam_filtered_noblanks, "PC1", "PC2", "Ethnic_Group", "Tooth_site", 3)
plot_pca_titles(malt_genus.decontam_filtered_noblanks, "PC3", "PC2", "Ethnic_Group", "Tooth_site", 3)


# by Individual sequencing depth
# perform a PCA to see how the data cluster
malt_genus.pca <- mixOmics::pca(malt_genus.decontam_filtered_noblanks, ncomp = 3, logratio = 'CLR')
plot(malt_genus.pca)

# Select out the PC variates into a new table and add the metadata for plotting
malt_genus.pca.variates <- as.data.frame(malt_genus.pca$variates$X)
malt_genus.pca.variates <- malt_genus.pca.variates %>%
  rownames_to_column("SampleID")
  
malt_genus.pca.varmet <- inner_join(malt_genus.pca.variates, metadata, by = "SampleID")

pca_malt_genus_pc1pc2 <- ggplot(malt_genus.pca.varmet, aes(PC1, PC2, colour = Individual_Seq_Depth_NonHuman, shape = Ethnic_Group)) +
  geom_point(size = 4) +
  # scale_colour_manual(values = common_colors) +
  # scale_shape_manual(values = village_shapes) +
  xlab(paste("PC1 - ", round(malt_genus.pca$explained_variance[1], 2), sep = "")) +
  ylab(paste("PC2 - ", round(malt_genus.pca$explained_variance[2], 2), sep = "")) +
  theme_minimal(base_size = 14) 

pca_malt_genus_pc1pc2


# ggsave("05-results.backup/pca_malt_genus_pc1pc2.pdf", plot = pca_malt_genus_pc1pc2, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"),
#        dpi = 300)

```

# Alpha-diversity
First we want to know if there are differences in the total diveristy of identified species between *XXXMETADATAXXX*. We'll look at alpha-diversity for this, using both the observed species and Shannon diverity metrics. Observed species to tell us which *XXXMETADATAXXX* have more species/genera, and Shannon to tell us which has more even distribution of species/genera. 
```{r observed_species}
# make Relab binary to make a presence/absence table
presabs_table <- malt_species.decontam %>%
                                         select(-one_of(outliers)) %>%
                                         adorn_totals(where = "col", name = "Sum_species.decontam") %>%
                                         mutate(Percent = Sum_species.decontam/sum(Sum_species.decontam)*100) %>%
                                         select(-one_of(c("Sum_species.decontam","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts) %>%
                                         gather("Species","Counts",2:ncol(.)) %>%
                                         mutate(Counts = ifelse(Counts > 0, 1, 0)) %>%
                                         spread(Species, Counts) %>%
                                         column_to_rownames("SampleID")

# calculate observed species for each sample with janitor 'adorn_totals'
malt_species.decontam.os <- presabs_table %>%
  adorn_totals(., where = "col", name = "Observed_species") %>%
  select(Observed_species) %>%
  rownames_to_column("SampleID") 

plot_box_whisker(malt_species.decontam.os, "Observed_species", "Env", "Env", "Number of species")
plot_box_whisker(malt_species.decontam.os, "Observed_species", "Village", "Village", "Number of species")


# now on filtered tables
presabs_table_f <- malt_species.decontam %>%
                                         select(-one_of(outliers)) %>%
                                         adorn_totals(where = "col", name = "Sum_species.decontam") %>%
                                         mutate(Percent = Sum_species.decontam/sum(Sum_species.decontam)*100) %>%
                                         filter(Percent > 0.001) %>%
                                         select(-one_of(c("Sum_species.decontam","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts) %>%
                                         gather("Species","Counts",2:ncol(.)) %>%
                                         mutate(Counts = ifelse(Counts > 0, 1, 0)) %>%
                                         spread(Species, Counts) %>%
                                         column_to_rownames("SampleID")
  
# calculate observed species for each sample with janitor 'adorn_totals'
malt_species.decontam_filtered.os <- presabs_table_f %>%
  adorn_totals(., where = "col", name = "Observed_species") %>%
  select(Observed_species) %>%
  rownames_to_column("SampleID") 

plot_box_whisker(malt_species.decontam_filtered.os, "Observed_species", "Env", "Env", "Number of species")
plot_box_whisker(malt_species.decontam_filtered.os, "Observed_species", "Village", "Village", "Number of species")

```

```{r shannon_diversity}
# Shannon diversity with the package vegan
malt_species.decontam.matrix <- as.matrix(malt_species.decontam %>%
                                         select(-one_of(outliers)) %>%
                                         adorn_totals(where = "col", name = "Sum_species.decontam") %>%
                                         mutate(Percent = Sum_species.decontam/sum(Sum_species.decontam)*100) %>%
                                         select(-one_of(c("Sum_species.decontam","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts) %>% 
                                         column_to_rownames("SampleID"))
 
malt_species.decontam.shannon <- as.data.frame(diversity(malt_species.decontam.matrix, index = "shannon"))
names(malt_species.decontam.shannon)[1] <- "Shannon_index"

# merge with the metadata file and plot
malt_species.decontam.shannon <- malt_species.decontam.shannon %>%
  rownames_to_column("SampleID") 

plot_box_whisker(malt_species.decontam.shannon, "Shannon_index", "Env", "Env", "Shannon Index")
plot_box_whisker(malt_species.decontam.shannon, "Shannon_index", "Village", "Village", "Shannon Index")


# now with the filtered data
malt_species.decontam_filtered.matrix <- as.matrix(malt_species.decontam %>%
                                         select(-one_of(outliers)) %>%
                                         adorn_totals(where = "col", name = "Sum_species.decontam") %>%
                                         mutate(Percent = Sum_species.decontam/sum(Sum_species.decontam)*100) %>%
                                         filter(Percent > 0.001) %>%
                                         select(-one_of(c("Sum_species.decontam","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts) %>% 
                                         column_to_rownames("SampleID"))

malt_species.decontam_filtered.shannon <- as.data.frame(diversity(malt_species.decontam_filtered.matrix, index = "shannon"))
names(malt_species.decontam_filtered.shannon)[1] <- "Shannon_index"

# merge with the metadata file and plot
malt_species.decontam_filtered.shannon <- malt_species.decontam_filtered.shannon %>%
  rownames_to_column("SampleID") 

plot_box_whisker(malt_species.decontam_filtered.shannon, "Shannon_index", "Env", "Env", "Shannon Index")
plot_box_whisker(malt_species.decontam_filtered.shannon, "Shannon_index", "Village", "Village", "Shannon Index")


```


Let's also look specifically at evenness of the samples using Pielou's evenness. This calculation (derived from Shannon's index) gives us a number from 0 to 1, where 0 is no evenness and 1 is perfect evenness. We want to see if there are any groups that have substantially greater CAZyme evenness than the others, we're not interested in what the actual number is, or how even the samples appear to be, except relative to eachother. After calculating plotting Pielou's evenness, make histograms to look at the evenness of CAZymes in each population.
```{r evenness_pielou}
evenness.species <- diversity(malt_species.decontam.matrix)
evenness.species <- as.data.frame(evenness.species/log(specnumber(malt_species.decontam_filtered.matrix)))
names(evenness.species)[1] <- "Pielou_evenness"

# merge with the metadata file and plot
evenness.species <- evenness.species %>%
  rownames_to_column("SampleID") 

plot_box_whisker(evenness.species, "Pielou_evenness", "Env", "Env", "Pielou's evenness")
plot_box_whisker(evenness.species, "Pielou_evenness", "Village", "Village", "Pielou's evenness")


# now with filtered data
evenness.species_filtered <- diversity(malt_species.decontam_filtered.matrix)
evenness.species_filtered <- as.data.frame(evenness.species_filtered/log(specnumber(malt_species.decontam_filtered.matrix)))
names(evenness.species_filtered)[1] <- "Pielou_evenness"

# merge with the metadata file and plot
evenness.species_filtered <- evenness.species_filtered %>%
  rownames_to_column("SampleID") 

plot_box_whisker(evenness.species_filtered, "Pielou_evenness", "Env", "Env", "Pielou's evenness")
plot_box_whisker(evenness.species_filtered, "Pielou_evenness", "Village", "Village", "Pielou's evenness")

```
In this dataset EXB059.A1601 has no detected genera, and this is the samples that was removed in the warning notice (it ends up with NA for the evenness calculation).

Now alpha-diversity at the genus level
```{r observed_genera}
presabs_table <- malt_genus.decontam %>%
                                         select(-one_of(outliers)) %>%
                                         adorn_totals(where = "col", name = "Sum_genus.decontam") %>%
                                         mutate(Percent = Sum_genus.decontam/sum(Sum_genus.decontam)*100) %>%
                                         select(-one_of(c("Sum_genus.decontam","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts) %>%
                                         gather("Species","Counts",2:ncol(.)) %>%
                                         mutate(Counts = ifelse(Counts > 0, 1, 0)) %>%
                                         spread(Species, Counts) %>%
                                         column_to_rownames("SampleID")
  
# calculate observed genus for each sample with janitor 'adorn_totals'
malt_genus.decontam.os <- presabs_table %>%
  adorn_totals(., where = "col", name = "Observed_genera") %>%
  select(Observed_genera) %>%
  rownames_to_column("SampleID") 

plot_box_whisker(malt_genus.decontam.os, "Observed_genera", "Env", "Env", "Number of genera")
plot_box_whisker(malt_genus.decontam.os, "Observed_genera", "Village", "Village", "Number of genera")


# now on the filtered table
presabs_table_filtered <- malt_genus.decontam %>%
                                         select(-one_of(outliers)) %>%
                                         adorn_totals(where = "col", name = "Sum_genus.decontam") %>%
                                         mutate(Percent = Sum_genus.decontam/sum(Sum_genus.decontam)*100) %>%
                                         filter(Percent > 0.001) %>%
                                         select(-one_of(c("Sum_genus.decontam","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts) %>%
                                         gather("Species","Counts",2:ncol(.)) %>%
                                         mutate(Counts = ifelse(Counts > 0, 1, 0)) %>%
                                         spread(Species, Counts) %>%
                                         column_to_rownames("SampleID")
  
# calculate observed genus for each sample with janitor 'adorn_totals'
malt_genus.decontam_filtered.os <- presabs_table_filtered %>%
  adorn_totals(., where = "col", name = "Observed_genera") %>%
  select(Observed_genera) %>%
  rownames_to_column("SampleID")

plot_box_whisker(malt_genus.decontam_filtered.os, "Observed_genera", "Env", "Env", "Number of genera")
plot_box_whisker(malt_genus.decontam_filtered.os, "Observed_genera", "Village", "Village", "Number of genera")

```

```{r shannon_genera}
# Shannon diversity with the package vegan
malt_genus.decontam.matrix <- as.matrix(malt_genus.decontam %>%
                                         select(-one_of(outliers)) %>%
                                         adorn_totals(where = "col", name = "Sum_genus.decontam") %>%
                                         mutate(Percent = Sum_genus.decontam/sum(Sum_genus.decontam)*100) %>%
                                         select(-one_of(c("Sum_genus.decontam","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts) %>% 
                                         column_to_rownames("SampleID"))

malt_genus.decontam.shannon <- as.data.frame(diversity(malt_genus.decontam.matrix, index = "shannon"))
names(malt_genus.decontam.shannon)[1] <- "Shannon_index"

# merge with the metadata file and plot
malt_genus.decontam.shannon <- malt_genus.decontam.shannon %>%
  rownames_to_column("SampleID") 

plot_box_whisker(malt_genus.decontam.shannon, "Shannon_index", "Env", "Env", "Shannon Index")
plot_box_whisker(malt_genus.decontam.shannon, "Shannon_index", "Village", "Village", "Shannon Index")


# now with the filtered data
malt_genus.decontam_filtered.matrix <- as.matrix(malt_genus.decontam %>%
                                         select(-one_of(outliers)) %>%
                                         adorn_totals(where = "col", name = "Sum_genus.decontam") %>%
                                         mutate(Percent = Sum_genus.decontam/sum(Sum_genus.decontam)*100) %>%
                                         filter(Percent > 0.001) %>%
                                         select(-one_of(c("Sum_genus.decontam","Percent"))) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts) %>% 
                                         column_to_rownames("SampleID"))

malt_genus.decontam_filtered.shannon <- as.data.frame(diversity(malt_genus.decontam_filtered.matrix, index = "shannon"))
names(malt_genus.decontam_filtered.shannon)[1] <- "Shannon_index"

# merge with the metadata file and plot
malt_genus.decontam_filtered.shannon <- malt_genus.decontam_filtered.shannon %>%
  rownames_to_column("SampleID") 

plot_box_whisker(malt_genus.decontam_filtered.shannon, "Shannon_index", "Env", "Env", "Shannon Index")
plot_box_whisker(malt_genus.decontam_filtered.shannon, "Shannon_index", "Village", "Village", "Shannon Index")

```

```{r evenness_pielou_genus}
evenness.genus <- diversity(malt_genus.decontam.matrix)
evenness.genus <- as.data.frame(evenness.genus/log(specnumber(malt_genus.decontam.matrix)))
names(evenness.genus)[1] <- "Pielou_evenness"

# merge with the metadata file and plot
evenness.genus <- evenness.genus %>%
  rownames_to_column("SampleID") 

plot_box_whisker(evenness.genus, "Pielou_evenness", "Env", "Env", "Pielou's evenness")
plot_box_whisker(evenness.genus, "Pielou_evenness", "Village", "Village", "Pielou's evenness")


# now with the filtered data
evenness_filtered.genus <- diversity(malt_genus.decontam_filtered.matrix)
evenness_filtered.genus <- as.data.frame(evenness_filtered.genus/log(specnumber(malt_genus.decontam_filtered.matrix)))
names(evenness_filtered.genus)[1] <- "Pielou_evenness"

# merge with the metadata file and plot
evenness_filtered.genus <- evenness_filtered.genus %>%
  rownames_to_column("SampleID") 

plot_box_whisker(evenness_filtered.genus, "Pielou_evenness", "Env", "Env", "Pielou's evenness")
plot_box_whisker(evenness_filtered.genus, "Pielou_evenness", "Village", "Village", "Pielou's evenness")

```
In this dataset EXB059.A1601 and LIB050.A0105 have no detected genera, and these are the 2 samples that were removed in the warning notice (they end up with NA for the evenness calculation).


Now make a heatmap of the top 50 most abundant species and genera (as an arbitrary cut-off)
```{r heatmaps, eval = F}
malt_species.decontam_top50 <- malt_species.decontam %>%
                                         select(-one_of(outliers)) %>%
                                         # select(-contains("SRR")) %>%
                                         select(-contains("10M")) %>%
                                         select(-contains("EXB")) %>%
                                         select(-contains("LIB")) %>%
                                         adorn_totals(where = "col", name = "Sum_species.decontam") %>%
                                         mutate(Percent = Sum_species.decontam/sum(Sum_species.decontam)*100) %>%
                                         top_n(50, Percent) %>%
                                         select(-one_of("Sum_species.decontam","Percent")) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts)

rownames(malt_species.decontam_top50) <- malt_species.decontam_top50$SampleID
malt_species.decontam_top50 <- malt_species.decontam_top50 %>% select(-SampleID)

# CLR-transform the matrix for plotting the heatmap
malt_species.decontam_top50_clr = malt_species.decontam_top50+1
#malt_species.decontam_top50_clr <- as.matrix(logratio.transfo(malt_species.decontam_top50_clr, logratio = "CLR"))
malt_species.decontam_top50_clr <- clr(malt_species.decontam_top50_clr)
malt_species.decontam_top50_clr <- as.matrix(malt_species.decontam_top50_clr)


malt_species.decontam_top50_clr.heatmap <- heatmap.2(malt_species.decontam_top50_clr, Colv = T, Rowv = T, dendrogram = "both",
         trace = "none", col = viridis_pal(),
         margins = c(8,6), srtCol = 30, cexCol = 0.6, cexRow = 0.5,
          key = T, keysize = 0.8, main = "50 most abundant species")


malt_genus.decontam_top50 <- malt_genus.decontam %>%
                                         select(-one_of(outliers)) %>%
                                         # select(-contains("SRR")) %>%
                                         select(-contains("10M")) %>%
                                         select(-contains("EXB")) %>%
                                         select(-contains("LIB")) %>%
                                         adorn_totals(where = "col", name = "Sum_genus.decontam") %>%
                                         mutate(Percent = Sum_genus.decontam/sum(Sum_genus.decontam)*100) %>%
                                         top_n(50, Percent) %>%
                                         select(-one_of("Sum_genus.decontam","Percent")) %>%
                                         gather("SampleID","Counts",2:ncol(.)) %>%
                                         spread(Species,Counts)

rownames(malt_genus.decontam_top50) <- malt_genus.decontam_top50$SampleID
malt_genus.decontam_top50 <- malt_genus.decontam_top50 %>% select(-SampleID)

# CLR-transform the matrix for plotting the heatmap
malt_genus.decontam_top50_clr = malt_genus.decontam_top50+1
#malt_genus.decontam_top50_clr <- as.matrix(logratio.transfo(malt_genus.decontam_top50_clr, logratio = "CLR"))
malt_genus.decontam_top50_clr <- clr(malt_genus.decontam_top50_clr)
malt_genus.decontam_top50_clr <- as.matrix(malt_genus.decontam_top50_clr)


malt_genus.decontam_top50_clr.heatmap <- heatmap.2(malt_genus.decontam_top50_clr, Colv = T, Rowv = T, dendrogram = "both",
         trace = "none", col = viridis_pal(),
         margins = c(8,6), srtCol = 30, cexCol = 0.6, cexRow = 0.5,
          key = T, keysize = 0.8, main = "50 most abundant genus")

```



