---
title: "Cameroon calculus/plaque MALT differentially abundant taxa by Songbird"
author: "Irina Velsko"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
    github_document:
        fig_width: 8
        fig_height: 6
        toc: true
---

```{r load_libraries, echo = F, message = F}
library(knitr)
library(data.table)
library(janitor)
library(tidyverse)
library(gplots)
library(viridis)
library(cowplot)
# source("./02-scripts.backup/ancom_v2.1.R") # source the ancom2 script to be able to use it here
opts_chunk$set(echo=F, warning=F, message=F, dpi=150)
```

Set the working directory of the notebook to the parent directory of the 
notebook itself.  

```{r setup}
knitr::opts_knit$set(root.dir = normalizePath(".."))

```

Load the taxonomy tables and songbird differentials tables
```{r load_data}

# load the taxonomy tables that have been decontaminated and the column headers are cleaned up
load("./05-results.backup/CMC_taxonomy_tables.RData")

songbird_noblanks_species_env <- fread("./05-results.backup/songbird_noblanks_species_env.tsv")
songbird_cmc_species_eg <- fread("./05-results.backup/songbird_cmc_species_eg.tsv")
songbird_cmc_species_me <- fread("./05-results.backup/songbird_cmc_species_me.tsv")
songbird_cmc_species_ts <- fread("./05-results.backup/songbird_cmc_species_ts.tsv")
songbird_cmc_species_sex <- fread("./05-results.backup/songbird_cmc_species_sex.tsv")

```

Load the ancom-BC differential abundance tables for comparison
```{r load_ancom_diffs}

ancomBC_noblanks_sp_HMP_HG <- fread("./05-results.backup/ancomBC_species_noblanks_HMP_HG.tsv")
ancomBC_noblanks_sp_JAE_HG <- fread("./05-results.backup/ancomBC_species_noblanks_JAE_HG.tsv")
ancomBC_cmc_sp_eg <- fread("./05-results.backup/ancomBC_species_cmc_ethnicgroup.tsv")
ancomBC_cmc_sp_Fa_FC <- fread("./05-results.backup/ancomBC_species_cmc_FA_FC.tsv")
ancomBC_cmc_sp_Fa_LR <- fread("./05-results.backup/ancomBC_species_cmc_FA_LR.tsv")
ancomBC_cmc_sp_ts <- fread("./05-results.backup/ancomBC_species_cmc_toothsite.tsv")
ancomBC_cmc_sp_sex <- fread("./05-results.backup/ancomBC_species_cmc_sex.tsv")

```


Read in the tables that have the species that are in MR1, for filtering the results later
```{r mr_species}
# read in the files that have the MR1, MR2, MR3 species/genera
mr_species <- fread("./05-results.backup/minres_oblimin_noyanomami.fa_species.tsv")
mr_cmc_species <- fread("./05-results.backup/minres_oblimin_cmc.fa_species.tsv")

```


List the outliers from the plot in section `raw_plot` in `Cameroon_h-g_taxonomy.Rmd` and remove them from the taxonomy table. 
```{r outliers}
outliers <- c("CMC002.B0201","CMC032.B0101","CMC037.B0201","CMC047.A0201",
              "EXB059.A1901","EXB059.A1301","EXB059.A1101","EXB059.A0901",
              "SRR1646041","SRR1646032","SRR1646042","SRR1646045","SRR1646030","SRR1646047")

outliersF <- str_c(outliers, collapse = "|")

```


```{r input_tables}
# new metadata table without factors, in case that's messing this up (Baka isn't appearing)
metadata <- fread("./00-documentation.backup/01-cameroon_hunter_gatherer_metadata.tsv")
metadata <- metadata %>%
  rename(SampleID = `#SampleID`) %>%
  filter(!SampleID %in% outliers,
         !str_detect(SampleID, "EXB|LIB|10M|SRR164")) %>%
  # rename(Sample.ID = SampleID) %>%
  as_tibble() %>%
  mutate(Market_economy = str_replace(Market_economy, "Forrest camp", "ForrestCamp"),
         Market_economy = str_replace(Market_economy, "Logging road", "LoggingRoad"),
         Market_economy_split = str_replace(Market_economy_split, "Forrest camp", "ForrestCamp"),
         Market_economy_split = str_replace(Market_economy_split, "Logging road", "LoggingRoad"),
         Market_economy_split = str_replace(Market_economy_split, "Industrial calculus", "IndustrialCalculus"),
         Market_economy_split = str_replace(Market_economy_split, "Industrial plaque", "IndustrialPlaque"),
         Industry = str_replace(Industry, "Non-industrial","nonIndustrial")) %>%
  select(SampleID, Env, Industry, Age_group, Env, Ethnic_Group, Market_economy, Market_economy_split, Sex, Tooth_site, Village)

# now re-name it for input to ancom-bc so I have to change fewer names
meta_data <- metadata


# species, no blanks, no Yanomami
malt_species.decontam_filtered_noblanks_noyanomami <- malt_species.decontam %>%
  select(-one_of(outliers)) %>%
  select(-matches("EXB|LIB|10M|SRR164")) %>%
  adorn_totals(where = "col") %>%
  filter(Total > 0) %>%
  select(-Total)

# species, CMC samples only
malt_species.decontam_filtered_cmc <- malt_species.decontam %>%
 select(-one_of(outliers)) %>%
 select(-matches("EXB|LIB|JAE|SRR")) %>%
  adorn_totals(where = "col") %>%
  filter(Total > 0) %>%
  select(-Total)

```

# Industrial and Cameroon Samples
```{r}
zeros <- songbird_noblanks_species_env %>%
  filter(`C(Env, Treatment('HunterGatherer'))[T.HMP]` >= 3 | `C(Env, Treatment('HunterGatherer'))[T.HMP]` <= -3) %>%
  rename(Species = featureid) %>%
  # mutate(Species = as_factor(Species)) %>%
  full_join(., ancomBC_noblanks_sp_HMP_HG %>%
               select(Species, diff.abn, `structural.zero (HMP)`, `structural.zero (HunterGatherer)`) %>%
               rename(ancomBC_da = diff.abn)) %>%
  mutate(ancomBC_da = str_replace(ancomBC_da, "TRUE","Yes"),
         ancomBC_da = replace_na(ancomBC_da, "No")) %>%
  drop_na(Intercept) %>%
  arrange(`C(Env, Treatment('HunterGatherer'))[T.HMP]`) %>%
  mutate(Species = as_factor(Species)) %>%
  ggplot(., aes(x = Species, y = `C(Env, Treatment('HunterGatherer'))[T.HMP]`, fill = ancomBC_da)) +
    geom_col() +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 0.5)) +
    ylab("Log fold change") +
    xlab("Species")

  
```



```{r plot_env}

songbird_HMP_CMC_plot <- songbird_noblanks_species_env %>%
  filter(`C(Env, Treatment('HunterGatherer'))[T.HMP]` >= 3 | `C(Env, Treatment('HunterGatherer'))[T.HMP]` <= -3) %>%
  rename(Species = featureid) %>%
  full_join(., ancomBC_noblanks_sp_HMP_HG %>%
               select(Species, diff.abn) %>%
               rename(ancomBC_da = diff.abn)) %>%
  mutate(ancomBC_da = str_replace(ancomBC_da, "TRUE","Yes"),
         ancomBC_da = str_replace(ancomBC_da, "FALSE","No"),
         ancomBC_da = replace_na(ancomBC_da, "No")) %>%
  drop_na(Intercept) %>%
  arrange(`C(Env, Treatment('HunterGatherer'))[T.HMP]`) %>%
  mutate(Species = as_factor(Species)) %>%
  ggplot(., aes(x = Species, y = `C(Env, Treatment('HunterGatherer'))[T.HMP]`, fill = ancomBC_da)) +
    geom_col() +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 0.5)) +
    ylab("Log fold change") +
    xlab("Species") +
    ggtitle("Cameroon plaque vs HMP plaque")

songbird_HMP_CMC_plot

songbird_JAE_CMC_plot <- songbird_noblanks_species_env %>%
  filter(`C(Env, Treatment('HunterGatherer'))[T.JAE]` >= 3 | `C(Env, Treatment('HunterGatherer'))[T.JAE]` <= -3) %>%
  rename(Species = featureid) %>%
  full_join(., ancomBC_noblanks_sp_JAE_HG %>%
               select(Species, diff.abn) %>%
               rename(ancomBC_da = diff.abn)) %>%
  mutate(ancomBC_da = str_replace(ancomBC_da, "TRUE","Yes"),
         ancomBC_da = str_replace(ancomBC_da, "FALSE","No"),
         ancomBC_da = replace_na(ancomBC_da, "No")) %>%
  drop_na(Intercept) %>%
  arrange(`C(Env, Treatment('HunterGatherer'))[T.JAE]`) %>%
  mutate(Species = as_factor(Species)) %>%
  ggplot(., aes(x = Species, y = `C(Env, Treatment('HunterGatherer'))[T.JAE]`, fill = ancomBC_da)) +
    geom_col() +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ylab("Log fold change") +
    xlab("Species") +
    ggtitle("Cameroon plaque vs JAE calculus")

songbird_JAE_CMC_plot

ggsave("./06-publication/supplemental_figures/SXX20/panel_parts/SXX20_species_songbird_hmp.pdf", plot = songbird_HMP_CMC_plot,
       device = "pdf", scale = 1, width = 8, height = 5, units = c("in"), dpi = 300)
ggsave("./06-publication/supplemental_figures/SXX20/panel_parts/SXX20_species_songbird_jae.pdf", plot = songbird_JAE_CMC_plot,
       device = "pdf", scale = 1, width = 11, height = 5, units = c("in"), dpi = 300)

 # ggsave("~/Desktop/songbird_HMP_CMC_plot.png", plot = songbird_HMP_CMC_plot, device = "png",
 #        scale = 1, width = 8, height = 4, units = c("in"), dpi = 300)
 # 
 # ggsave("~/Desktop/songbird_JAE_CMC_plot.png", plot = songbird_JAE_CMC_plot, device = "png",
 #        scale = 1, width = 8, height = 4, units = c("in"), dpi = 300)

```

# Cameroon samples only

## Ethnic Group
```{r plot_eg}

songbird_cmc_species_eg %>%
  filter(`C(Ethnic_Group, Treatment('Nzime'))[T.Baka]` >= 1 | `C(Ethnic_Group, Treatment('Nzime'))[T.Baka]` <= -1) %>%
  rename(Species = featureid) %>%
  full_join(., ancomBC_cmc_sp_eg %>%
               select(Species, diff.abn) %>%
               rename(ancomBC_da = diff.abn)) %>%
  mutate(ancomBC_da = str_replace(ancomBC_da, "TRUE","Yes"),
         ancomBC_da = replace_na(ancomBC_da, "No")) %>%
  drop_na(Intercept) %>%
  arrange(`C(Ethnic_Group, Treatment('Nzime'))[T.Baka]`) %>%
  mutate(Species = as_factor(Species)) %>%
  ggplot(., aes(x = Species, y = `C(Ethnic_Group, Treatment('Nzime'))[T.Baka]`, fill = ancomBC_da)) +
    geom_col() +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ylab("Log fold change") +
    xlab("Species")


```


## Market Economy
```{r plot_me}

me_FC <- songbird_cmc_species_me %>%
  filter(`C(Market_economy, Treatment('Farming'))[T.Forrest camp]` >= 1 | `C(Market_economy, Treatment('Farming'))[T.Forrest camp]` <= -1) %>%
  rename(Species = featureid) %>%
  full_join(., ancomBC_cmc_sp_Fa_FC %>%
               select(Species, diff.abn) %>%
               rename(ancomBC_da = diff.abn)) %>%
  mutate(ancomBC_da = str_replace(ancomBC_da, "TRUE","Yes"),
         ancomBC_da = replace_na(ancomBC_da, "No")) %>%
  drop_na(Intercept) %>%
  arrange(`C(Market_economy, Treatment('Farming'))[T.Forrest camp]`) %>%
  mutate(Species = as_factor(Species)) %>%
  ggplot(., aes(x = Species, y = `C(Market_economy, Treatment('Farming'))[T.Forrest camp]`, fill = ancomBC_da)) +
    geom_col() +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ylab("Log fold change") +
    xlab("Species") +
    ggtitle("Forrest Camp vs. Farming")

me_FC

me_LR <- songbird_cmc_species_me %>%
  filter(`C(Market_economy, Treatment('Farming'))[T.Logging road]` >= 1 | `C(Market_economy, Treatment('Farming'))[T.Logging road]` <= -1) %>%
  rename(Species = featureid) %>%
  full_join(., ancomBC_cmc_sp_Fa_LR %>%
               select(Species, diff.abn) %>%
               rename(ancomBC_da = diff.abn)) %>%
  mutate(ancomBC_da = str_replace(ancomBC_da, "TRUE","Yes"),
         ancomBC_da = replace_na(ancomBC_da, "No")) %>%
  drop_na(Intercept) %>%
  arrange(`C(Market_economy, Treatment('Farming'))[T.Logging road]`) %>%
  mutate(Species = as_factor(Species)) %>%
  ggplot(., aes(x = Species, y = `C(Market_economy, Treatment('Farming'))[T.Logging road]`, fill = ancomBC_da)) +
    geom_col() +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ylab("Log fold change") +
    xlab("Species") +
    ggtitle("Logging Road vs. Farming")
me_LR

ggsave("./06-publication/supplemental_figures/SXX20/panel_parts/SXX20_species_songbird_me_FC.pdf", plot = me_FC,
       device = "pdf", scale = 1, width = 9, height = 5, units = c("in"), dpi = 300)
ggsave("./06-publication/supplemental_figures/SXX20/panel_parts/SXX20_species_songbird_me_LR.pdf", plot = me_LR,
       device = "pdf", scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

```


## Tooth Site
```{r plot_ts}

songbird_ts <- songbird_cmc_species_ts %>%
  filter(`C(Tooth_site, Treatment('Posterior'))[T.Anterior]` >= 2 | `C(Tooth_site, Treatment('Posterior'))[T.Anterior]` <= -2) %>%
  rename(Species = featureid) %>%
  full_join(., ancomBC_cmc_sp_ts %>%
               select(Species, diff.abn) %>%
               rename(ancomBC_da = diff.abn)) %>%
  mutate(ancomBC_da = str_replace(ancomBC_da, "TRUE","Yes"),
         ancomBC_da = str_replace(ancomBC_da, "FALSE","No"),
         ancomBC_da = replace_na(ancomBC_da, "No")) %>%
  drop_na(Intercept) %>%
  arrange(`C(Tooth_site, Treatment('Posterior'))[T.Anterior]`) %>%
  mutate(Species = as_factor(Species)) %>%
  ggplot(., aes(x = Species, y = `C(Tooth_site, Treatment('Posterior'))[T.Anterior]`, fill = ancomBC_da)) +
    geom_col() +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ylab("Log fold change") +
    xlab("Species")

songbird_ts

ggsave("./06-publication/supplemental_figures/SXX20/panel_parts/SXX20_species_songbird_tooth_site.pdf", plot = songbird_ts,
       device = "pdf", scale = 1, width = 6, height = 5, units = c("in"), dpi = 300)

```



## Sex
```{r plot_sex}

songbird_cmc_species_sex %>%
  filter(`C(Sex, Treatment('Male'))[T.Female]` >= 1 | `C(Sex, Treatment('Male'))[T.Female]` <= -1) %>%
  arrange(`C(Sex, Treatment('Male'))[T.Female]`) %>%
  rename(Species = featureid) %>%
  full_join(., ancomBC_cmc_sp_sex %>%
               select(Species, diff.abn) %>%
               rename(ancomBC_da = diff.abn)) %>%
  mutate(ancomBC_da = str_replace(ancomBC_da, "TRUE","Yes"),
         ancomBC_da = replace_na(ancomBC_da, "No")) %>%
  drop_na(Intercept) %>%
  mutate(Species = as_factor(Species)) %>%
  ggplot(., aes(x = Species, y = `C(Sex, Treatment('Male'))[T.Female]`, fill = ancomBC_da)) +
    geom_col() +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ylab("Log fold change") +
    xlab("Species")


```










































