---
title: "Cameroon hunter-gatherer calculus/plaque SEED heat maps"
author: "Irina Velsko"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
    github_document:
        fig_width: 8
        fig_height: 6
        toc: true
---


```{r load_libraries, echo = F, message = F}
library(knitr)
library(data.table)
library(psych)
library(vegan)
# library(ape)
library(mixOmics)
library(compositions)
# library(MASS)
library(janitor)
# library(rstatix)
# library(pander)
library(tidyverse)
# library(gplots)
# library(ggrepel)
library(viridis)
library(cowplot)
opts_chunk$set(echo=F, warning=F, message=F, dpi=150)
```

Set the working directory of the notebook to the parent directory of the 
notebook itself.  

```{r setup}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r load_data}
## load the species and genus tables generated with HUMAnN2
load("./05-results.backup/CMC_seed_tables.RData")

seed_L4.decontam_noblanks.fa_seed <- fread("./05-results.backup/minres_oblimin_noblanks.fa_seed.tsv")
seed_L4.decontam_cmc.fa_seed <- fread("./05-results.backup/minres_oblimin_cmc.fa_seed.tsv")

```

```{r load_metadata}

load("./05-results.backup/CMC_metadata.RData")

```

## Heatmaps
Make a heat map of the species in MR1 and MR2 to see how they're different between the groups
```{r metadata_heatmaps}
# minres_oblimin_nobl_noyano_fa_species <- fread("./05-results.backup/minres_oblimin_noyanomami.fa_species.tsv", sep = "\t")
# minres_oblimin_cmc_fa_species <- fread("./05-results.backup/minres_oblimin_cmc.fa_species.tsv", sep = "\t")

# read in metadata again to get it without all the factors
metadata_hm <- fread("00-documentation.backup/01-cameroon_hunter_gatherer_metadata.tsv")

# order the samples so they can be ordered this way in the heat maps

# order the CMC samples by Village
cmc_order <- metadata_hm %>%
  rename(SampleID = `#SampleID`)  %>%
  as_tibble() %>%
  filter(str_detect(SampleID, "CMC")) %>%
  mutate(Village = fct_relevel(Village, "202","205","204","203","206")) %>%
  select(SampleID, Village) %>%
  arrange(Village) %>%
  pull(SampleID)

# order the HMP samples by subplaque, supraplaque
hmp_order <- metadata_hm %>%
  rename(SampleID = `#SampleID`)  %>%
  as_tibble() %>%
  filter(str_detect(SampleID, "SRR")) %>%
  filter(!str_detect(SampleID, "10M|SRR164")) %>%
  mutate(Description = fct_relevel(Description, "HMP_subPlaque","HMP_supPlaque")) %>%
  select(SampleID, Description) %>%
  arrange(Description) %>%
  pull(SampleID)

# keep JAE samples in alphabetical order
jae_order <- metadata_hm %>%
  rename(SampleID = `#SampleID`)  %>%
  as_tibble() %>%
  filter(str_detect(SampleID, "JAE")) %>%
  filter(!str_detect(SampleID, "10M")) %>%
  pull(SampleID)

# combine all 3 ordered vectors into a final ordered vector
each_order <- c(cmc_order, jae_order, hmp_order)

```

To change the heatmap label colors by different metadata categories swap out the
sections below
```{r heatmap_label_colors, eval = F}

Village_colors %>%
               as_data_frame(.) %>%
               rename(Color = value) %>%
               mutate(Village = c("202","205","204","203","206","Yanomami","HMP","HMP10M","JAE10M","JAE","ExtractionGauze","ExtractionBlank","LibraryBlank")) %>%
               filter(!str_detect(Village, "Yanomami|10M|Gauze|Blank")), by = "Village"

Sex_colors %>%
               as_data_frame(.) %>%
               rename(Color = value) %>%
               mutate(Sex = c("Female","Male","Unknown","ExtractionGauze","ExtractionBlank","LibraryBlank")) %>%
               filter(!str_detect(Sex, "Gauze|Blank")), by = "Sex"


Age_group_colors %>%
               as_data_frame(.) %>%
               rename(Color = value) %>%
               mutate(Age_group = c("18-29","30-39","40-49","50-59","60+","10-18","HMP","HMP10M","ExtractionGauze","ExtractionBlank","LibraryBlank")) %>%
               filter(!str_detect(Age_group, "10|10M|Gauze|Blank")), by = "Age_group"


```

Try changing the sample order to group them by the different metadata categories
(currently have Industrialization/Ethnic group/Market economy/Village grouping).
Need to look at: Tooth site, Age group, Sex, others? Also try clustering the
samples to see if they fall into any recognizable categories.
```{r heatmap_noblanks_noyanomami}
# select the SEED proteins in MR1 and MR2 for the FA
seed_L4.decontam_FA_all <- seed_L4.decontam %>%
                                  select(-matches("SRR164|10M|EXB|LIB|CMC032.B")) %>%
                                  separate(., Pathway, into = c("SEED","L2","L3","Pathway"), sep = ";") %>%
                                  select(-c("SEED","L2","L3")) %>%
                                  inner_join(., seed_L4.decontam_noblanks.fa_seed %>%
                                               filter(Residual == "MR2") %>%
                                               select(-Residual)) %>%
                                  gather("SampleID","Counts",2:ncol(.)) %>%
                                  mutate(Counts = Counts + 1) %>%
                                  spread(Pathway,Counts) %>%
                                  column_to_rownames("SampleID")

# CLR-transform the matrix for plotting the heatmap
seed_L4.decontam_FA_all_clr <- clr(seed_L4.decontam_FA_all)
seed_L4.decontam_FA_all_clr <- as.matrix(seed_L4.decontam_FA_all_clr)

# orient the matrix with correct rows/columns to run the distance calculation for the seed protein, not the samples (columns are SampleID, rows are proteins)
seed_L4.decontam_FA_all_clr_sp <- seed_L4.decontam_FA_all_clr %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  gather("Pathway","Counts",2:ncol(.)) %>%
  spread(SampleID,Counts) %>%
  column_to_rownames("Pathway")

# perform a distance calculation for hierarchical clustering for the protein ordering
seed_L4.decontam_FA_all_clr_dist <- vegdist(seed_L4.decontam_FA_all_clr_sp, method = "euclidian")
# cluster the data with hclust
seed_L4.decontam_FA_all_clr_dist_hr <- hclust(seed_L4.decontam_FA_all_clr_dist)
# seed_L4.decontam_FA_all_clr_dist_hr$order gives the dendrogram order from the bottom to the top, but using the row numbers not the row names
# pull this order out and store as a vector to make the order a factor in the next step for plotting
sp_all_order <- seed_L4.decontam_FA_all_clr_dist_hr$order %>%
  as.character.numeric_version(.) %>%
  str_trim(.)

# use the dendrogram order to pull out the SEED protein in the correct order for the heatmap
seed_L4.decontam_FA_all_clr_order <- seed_L4.decontam_FA_all_clr %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  gather("Pathway","CLR",2:ncol(.)) %>%
  spread(SampleID,CLR) %>%
  mutate(SpOrder = as.character(seq(1,869))) %>% # MR1 - 700; MR2 - 869
  select(Pathway, SpOrder, everything()) %>%
  mutate(SpOrder = fct_relevel(SpOrder, sp_all_order)) %>%
  arrange(SpOrder) %>%
  pull(Pathway)

# make the protein a factor so they're in the correct order for the heatmap using the vector from the step above
seed_L4.decontam_FA_all_clr_long <- seed_L4.decontam_FA_all_clr %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  gather("Pathway","CLR",2:ncol(.)) %>%
  spread(SampleID,CLR) %>%
  mutate(Pathway = fct_relevel(Pathway, seed_L4.decontam_FA_all_clr_order)) %>%
  arrange(Pathway) %>%
  gather("SampleID","CLR", 2:ncol(.))


# now cluster by sample
# convert the matrix to a data.frame
seed_L4.decontam_FA_all_clr_smp <- seed_L4.decontam_FA_all_clr %>%
  as.data.frame()

# perform a distance calculation for hierarchical clustering for the Pathway ordering
seed_L4.decontam_FA_all_clr_smp_dist <- vegdist(seed_L4.decontam_FA_all_clr_smp, method = "euclidian")

# cluster the data with hclust
seed_L4.decontam_FA_all_clr_smp_dist_hr <- hclust(seed_L4.decontam_FA_all_clr_smp_dist)
# seed_L4.decontam_FA_all_clr_dist_hr$order gives the dendrogram order from the bottom to the top, but using the row numbers not the row names
# pull this order out and store as a vector to make the order a factor in the next step for plotting
smp_all_order <- seed_L4.decontam_FA_all_clr_smp_dist_hr$order %>%
  as.character.numeric_version(.) %>%
  str_trim(.)

# use the dendrogram order to pull out the Pathway in the correct order for the heatmap
seed_L4.decontam_FA_all_clr_smp_order <- seed_L4.decontam_FA_all_clr %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  gather("Pathway","CLR",2:ncol(.)) %>%
  spread(Pathway,CLR) %>%
  mutate(SmpOrder = as.character(seq(1,113))) %>%
  select(SampleID, SmpOrder, everything()) %>%
  mutate(SmpOrder = fct_relevel(SmpOrder, smp_all_order)) %>%
  arrange(SmpOrder) %>%
  pull(SampleID)

# make the SampleIDs a factor so they're in the correct order for the heatmap using the vector from the step above
# Plot the heatmap with samples ordered by the clustering from above
# make the SampleIDs a factor so they're in the correct order for the heatmap using the vector from the step above
seed_L4.decontam_FA_all_clr_long <- seed_L4.decontam_FA_all_clr %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  gather("Species","CLR",2:ncol(.)) %>%
  spread(SampleID,CLR) %>%
  mutate(Species = fct_relevel(Species, seed_L4.decontam_FA_all_clr_order)) %>%
  arrange(Species) %>%
  gather("SampleID","CLR", 2:ncol(.)) %>%
  inner_join(., metadata %>%
               select(SampleID, Sex), by = "SampleID") %>%
  inner_join(., Sex_colors %>%
               as_data_frame(.) %>%
               rename(Color = value) %>%
               mutate(Sex = c("Female","Male","Unknown","ExtractionGauze","ExtractionBlank","LibraryBlank")) %>%
               filter(!str_detect(Sex, "Gauze|Blank")), by = "Sex") %>%
  mutate(SampleID = fct_relevel(SampleID, seed_L4.decontam_FA_all_clr_smp_order)) %>%
  arrange(SampleID)

vector_color = seed_L4.decontam_FA_all_clr_long %>%
  select(SampleID, Color) %>%
  unique %>%
  pull(Color) %>%
  as.character()

# make a heatmap with ggplot
heatmap_smp_cluster <- ggplot(seed_L4.decontam_FA_all_clr_long, aes(SampleID, Species, fill = CLR)) +
  geom_tile() +
  scale_fill_viridis_c(option = "A") +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.95, color = vector_color),
        plot.margin = margin(0, 0, 0, 0, "pt"),
        axis.title.y = element_blank(),
        legend.position = "left")

heatmap_smp_cluster


# Plot the heatmap with samples ordered by Village, then alphabetical
# make the SampleIDs a factor so they're in the correct order for the heatmap using the vector from the step above
seed_L4.decontam_FA_all_clr_long <- seed_L4.decontam_FA_all_clr %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  gather("Species","CLR",2:ncol(.)) %>%
  spread(SampleID,CLR) %>%
  mutate(Species = fct_relevel(Species, seed_L4.decontam_FA_all_clr_order)) %>%
  arrange(Species) %>%
  gather("SampleID","CLR", 2:ncol(.)) %>%
  inner_join(., metadata %>%
               select(SampleID, Village), by = "SampleID") %>%
  inner_join(., Village_colors %>%
               as_data_frame(.) %>%
               rename(Color = value) %>%
               mutate(Village = c("202","205","204","203","206","Yanomami","HMP","HMP10M","JAE10M","JAE","ExtractionGauze","ExtractionBlank","LibraryBlank")), by = "Village") %>%
  mutate(SampleID = fct_relevel(SampleID, each_order)) %>%
  arrange(SampleID)

vector_color = seed_L4.decontam_FA_all_clr_long %>%
  select(SampleID, Color) %>%
  unique %>%
  pull(Color) %>%
  as.character()

# make a heatmap with ggplot
heatmap_smp_village <- ggplot(seed_L4.decontam_FA_all_clr_long, aes(SampleID, Species, fill = CLR)) +
  geom_tile() +
  scale_fill_viridis_c(option = "A") +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.95, color = vector_color),
        plot.margin = margin(0, 0, 0, 0, "pt"),
        axis.title.y = element_blank(),
        legend.position = "left")

heatmap_smp_village

# ggsave("05-results.backuppresentation_pdfs/noblanks_fa_heatmap.pdf", plot = heatmap_base, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

```

## Cameroon samples only
```{r heatmap_cmc_noyanomami}
# select the SEED proteins in MR1 and MR2 for the FA
seed_L4.decontam_FA_cmc <- seed_L4.decontam %>%
                                  select(-matches("SRR|JAE|EXB|LIB|CMC032.B")) %>%
                                  separate(., Pathway, into = c("SEED","L2","L3","Pathway"), sep = ";") %>%
                                  select(-c("SEED","L2","L3")) %>%
                                  inner_join(., seed_L4.decontam_cmc.fa_seed %>%
                                               filter(Residual == "MR2") %>%
                                               select(-Residual)) %>%
                                  gather("SampleID","Counts",2:ncol(.)) %>%
                                  mutate(Counts = Counts + 1) %>%
                                  spread(Pathway,Counts) %>%
                                  column_to_rownames("SampleID")

# CLR-transform the matrix for plotting the heatmap
seed_L4.decontam_FA_cmc_clr <- clr(seed_L4.decontam_FA_cmc)
seed_L4.decontam_FA_cmc_clr <- as.matrix(seed_L4.decontam_FA_cmc_clr)

# orient the matrix with correct rows/columns to run the distance calculation for the seed protein, not the samples (columns are SampleID, rows are proteins)
seed_L4.decontam_FA_cmc_clr_sp <- seed_L4.decontam_FA_cmc_clr %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  gather("Pathway","Counts",2:ncol(.)) %>%
  spread(SampleID,Counts) %>%
  column_to_rownames("Pathway")

# perform a distance calculation for hierarchical clustering for the protein ordering
seed_L4.decontam_FA_cmc_clr_dist <- vegdist(seed_L4.decontam_FA_cmc_clr_sp, method = "euclidian")
# cluster the data with hclust
seed_L4.decontam_FA_cmc_clr_dist_hr <- hclust(seed_L4.decontam_FA_cmc_clr_dist)
# seed_L4.decontam_FA_cmc_clr_dist_hr$order gives the dendrogram order from the bottom to the top, but using the row numbers not the row names
# pull this order out and store as a vector to make the order a factor in the next step for plotting
sp_cmc_order <- seed_L4.decontam_FA_cmc_clr_dist_hr$order %>%
  as.character.numeric_version(.) %>%
  str_trim(.)

# use the dendrogram order to pull out the SEED protein in the correct order for the heatmap
seed_L4.decontam_FA_cmc_clr_order <- seed_L4.decontam_FA_cmc_clr %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  gather("Pathway","CLR",2:ncol(.)) %>%
  spread(SampleID,CLR) %>%
  mutate(SpOrder = as.character(seq(1,712))) %>%
  select(Pathway, SpOrder, everything()) %>%
  mutate(SpOrder = fct_relevel(SpOrder, sp_cmc_order)) %>%
  arrange(SpOrder) %>%
  pull(Pathway)

# make the protein a factor so they're in the correct order for the heatmap using the vector from the step above
seed_L4.decontam_FA_cmc_clr_long <- seed_L4.decontam_FA_cmc_clr %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  gather("Pathway","CLR",2:ncol(.)) %>%
  spread(SampleID,CLR) %>%
  mutate(Pathway = fct_relevel(Pathway, seed_L4.decontam_FA_cmc_clr_order)) %>%
  arrange(Pathway) %>%
  gather("SampleID","CLR", 2:ncol(.))


# now cluster by sample
# convert the matrix to a data.frame
seed_L4.decontam_FA_cmc_clr_smp <- seed_L4.decontam_FA_cmc_clr %>%
  as.data.frame()

# perform a distance calculation for hierarchical clustering for the Pathway ordering
seed_L4.decontam_FA_cmc_clr_smp_dist <- vegdist(seed_L4.decontam_FA_cmc_clr_smp, method = "euclidian")

# cluster the data with hclust
seed_L4.decontam_FA_cmc_clr_smp_dist_hr <- hclust(seed_L4.decontam_FA_cmc_clr_smp_dist)
# seed_L4.decontam_FA_cmc_clr_dist_hr$order gives the dendrogram order from the bottom to the top, but using the row numbers not the row names
# pull this order out and store as a vector to make the order a factor in the next step for plotting
smp_cmc_order <- seed_L4.decontam_FA_cmc_clr_smp_dist_hr$order %>%
  as.character.numeric_version(.) %>%
  str_trim(.)

# use the dendrogram order to pull out the Pathway in the correct order for the heatmap
seed_L4.decontam_FA_cmc_clr_smp_order <- seed_L4.decontam_FA_cmc_clr %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  gather("Pathway","CLR",2:ncol(.)) %>%
  spread(Pathway,CLR) %>%
  mutate(SmpOrder = as.character(seq(1,83))) %>%
  select(SampleID, SmpOrder, everything()) %>%
  mutate(SmpOrder = fct_relevel(SmpOrder, smp_cmc_order)) %>%
  arrange(SmpOrder) %>%
  pull(SampleID)

# make the SampleIDs a factor so they're in the correct order for the heatmap using the vector from the step above
# Plot the heatmap with samples ordered by the clustering from above
# make the SampleIDs a factor so they're in the correct order for the heatmap using the vector from the step above
seed_L4.decontam_FA_cmc_clr_long <- seed_L4.decontam_FA_cmc_clr %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  gather("Species","CLR",2:ncol(.)) %>%
  spread(SampleID,CLR) %>%
  mutate(Species = fct_relevel(Species, seed_L4.decontam_FA_cmc_clr_order)) %>%
  arrange(Species) %>%
  gather("SampleID","CLR", 2:ncol(.)) %>%
  inner_join(., metadata %>%
               select(SampleID, Sex), by = "SampleID") %>%
  inner_join(., Sex_colors %>%
               as_data_frame(.) %>%
               rename(Color = value) %>%
               mutate(Sex = c("Female","Male","Unknown","ExtractionGauze","ExtractionBlank","LibraryBlank")) %>%
               filter(!str_detect(Sex, "Gauze|Blank")), by = "Sex") %>%
  mutate(SampleID = fct_relevel(SampleID, seed_L4.decontam_FA_cmc_clr_smp_order)) %>%
  arrange(SampleID)

vector_color = seed_L4.decontam_FA_cmc_clr_long %>%
  select(SampleID, Color) %>%
  unique %>%
  pull(Color) %>%
  as.character()

# make a heatmap with ggplot
heatmap_smp_cluster <- ggplot(seed_L4.decontam_FA_cmc_clr_long, aes(SampleID, Species, fill = CLR)) +
  geom_tile() +
  scale_fill_viridis_c(option = "A") +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.95, color = vector_color),
        plot.margin = margin(0, 0, 0, 0, "pt"),
        axis.title.y = element_blank(),
        legend.position = "left")

heatmap_smp_cluster


# Plot the heatmap with samples ordered by Village, then alphabetical
# make the SampleIDs a factor so they're in the correct order for the heatmap using the vector from the step above
seed_L4.decontam_FA_cmc_clr_long <- seed_L4.decontam_FA_cmc_clr %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  gather("Species","CLR",2:ncol(.)) %>%
  spread(SampleID,CLR) %>%
  mutate(Species = fct_relevel(Species, seed_L4.decontam_FA_cmc_clr_order)) %>%
  arrange(Species) %>%
  gather("SampleID","CLR", 2:ncol(.)) %>%
  inner_join(., metadata %>%
               select(SampleID, Village), by = "SampleID") %>%
  inner_join(., Village_colors %>%
               as_data_frame(.) %>%
               rename(Color = value) %>%
               mutate(Village = c("202","205","204","203","206","Yanomami","HMP","HMP10M","JAE10M","JAE","ExtractionGauze","ExtractionBlank","LibraryBlank")), by = "Village") %>%
  mutate(SampleID = fct_relevel(SampleID, each_order)) %>%
  arrange(SampleID)

vector_color = seed_L4.decontam_FA_cmc_clr_long %>%
  select(SampleID, Color) %>%
  unique %>%
  pull(Color) %>%
  as.character()

# make a heatmap with ggplot
heatmap_smp_village <- ggplot(seed_L4.decontam_FA_cmc_clr_long, aes(SampleID, Species, fill = CLR)) +
  geom_tile() +
  scale_fill_viridis_c(option = "A") +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.95, color = vector_color),
        plot.margin = margin(0, 0, 0, 0, "pt"),
        axis.title.y = element_blank(),
        legend.position = "left")

heatmap_smp_village

# ggsave("05-results.backuppresentation_pdfs/cmc_fa_heatmap.pdf", plot = heatmap_base, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

```


